---
title: "AMR_RGI_analysis"
output: html_document
date: "2024-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(dplyr)
library(tidyr)
library(readxl)
library(stringr)
library(KEGGREST)
library(speedyseq)

# For visualization and statistical tests
library(phyloseq)
library(randomcoloR)
library(ggplot2)
theme_set(theme_bw())
library(ComplexHeatmap)
library(circlize)  # For color scaling
library(vegan)
library(cowplot)
library(microViz)
library(patchwork) # for arranging groups of plots
library(breakaway)
```

# Read in data

## Load raw data 

First read in the RGI data table. Note, the table that came from the RGI program was kinda tough in the first column. So I changed the first column to have just the contig/gene ID, then got rid of the empty columns (Contig, start, stop, orientation). I also got rid of teh SNPs column and predicted_DNA columns. I also just tried saving it as an Excel document to see if it works better. 
```{r}
amr <- read_excel("../data/FLK_prok_coassembly_RGI_AMR.xlsx")

# Make a short ID that corresponds to the AMR gene family name 
amr.fam <- data.frame("AMR_Gene_Family" = unique(amr$AMR_Gene_Family))
amr.fam$AMR_Gene_Family_ID <- paste0("amrgenefam_", seq_len(nrow(amr.fam)))

amr.new <- amr %>%
  left_join(amr.fam, by = "AMR_Gene_Family")

table(amr.new$AMR_Gene_Family_ID == "NA")
```

Need to read in the count, taxonomy, and metadata data for each gene. 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata
metadata <- as.data.frame(read_excel("../data/RRC_sample_CoreSiteData_v10_RRCCoreData_SP1.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)

# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)

# Read in the taxonomy
prok_taxa <- read.table("../data/FLK_prok_coassembly_tax_mmseqs2.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

## Subset the metadata
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced samples
```

# Data analysis 

I am interested in how many different types of hits there were. Start by capturing some summary statistics of the different groupings. 

```{r}
ggplot(amr, aes(x = Cut_Off, fill = Resistance_Mechanism)) +
  geom_bar(position = "fill")

ggplot(amr, aes(x = Cut_Off, fill = Resistance_Mechanism)) +
  geom_bar()
```
Generate some summary statistics regarding the different AMR categories of all 200,000 hits
```{r}
amr %>% 
  dplyr::count(Model_type)

amr %>% 
  dplyr::count(Cut_Off)

x <- amr %>% 
  dplyr::count(Resistance_Mechanism) 

amr %>% 
  dplyr::count(Drug_Class)

amr %>% 
  dplyr::count(AMR_Gene_Family)
```
Let's look at the 49 "strict" hits. What are some different stats for them?
```{r}
strict_amr <- amr %>% 
  filter(Cut_Off == "Strict")

strict_amr %>% 
  dplyr::count(Drug_Class)

strict_amr %>% 
  dplyr::count(AMR_Gene_Family)
```

# Heatmap

Let's make a heatmap of the different genes that are "Strict" AMR gene hits. 

## prep data
```{r}
# add a column of contig ids to the prok reads table
prokreads_nozero$Genes <- rownames(prokreads_nozero)

# First, I need to get a matrix of the abundances of genes that have "strict" hits
strict_amr_counts <- strict_amr %>%
  left_join(prokreads_nozero, by = c("ORF_ID" = "Genes")) 

strict_amr_counts_matrix <- as.matrix(strict_amr_counts[, 17:57])

# Add the rownames (Gene IDs)
rownames(strict_amr_counts_matrix) <- strict_amr_counts$ORF_ID

# Scale using z-score standardization for each gene (row) across samples (columsn)
z_scores_amr <- t(apply(strict_amr_counts_matrix, 1, scale))  # Scale rows (genes)
colnames(z_scores_amr) <- colnames(strict_amr_counts_matrix)
```

## Make the heatmap
```{r}
# Define the color scale
col_fun <- colorRamp2(c(-1, 0, 6), c("blue", "white", "red"))  # Z-score scale: low -> blue, high -> red

# Generate the heatmap
Heatmap(
  z_scores_amr,
  name = "Z-Score",
  col = col_fun,
  cluster_rows = TRUE,  # Cluster genes
  cluster_columns = TRUE,  # Cluster samples
  show_row_dend = TRUE,  # Show row dendrogram
  show_column_dend = TRUE,  # Show column dendrogram
  row_title = "AMR Genes",
  column_title = "Coral Samples",
  row_names_gp = gpar(fontsize = 8),  # Adjust font size
  column_names_gp = gpar(fontsize = 8),
  heatmap_legend_param = list(title = "Gene Abundance")
)
```

This is a great start! I want to add colors to the samples so I know which ones are diseased or healthy based on the "Disease at sampling category". 
```{r}
coral_sample_annotation <- HeatmapAnnotation(
  Group = metadata_prok$Disease_on_colony_when_coring_binary,
  col = list(Group = c("Unaffected" = "#0C7BDC", "Diseased" = "#FFC20A"))
)

gene_labels <- strict_amr_counts$AMR_Gene_Family

# Include the annotation in the heatmap
Heatmap(
  z_scores_amr,
  name = "Z-Score",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  row_title = "AMR Genes",
  column_title = "Coral Samples",
  top_annotation = coral_sample_annotation,  # Add sample group annotation
  row_labels = gene_labels,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  heatmap_legend_param = list(title = "Gene Abundance")
)

# Move the legend
draw(
Heatmap(
  z_scores_amr,
  name = "Z-Score",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  row_title = "AMR Genes",
  column_title = "Coral Samples",
  top_annotation = coral_sample_annotation,  # Add sample group annotation
  row_labels = gene_labels,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  heatmap_legend_param = list(title = "Gene Abundance")
),
  heatmap_legend_side = "right",          # Heatmap legend on the left
  annotation_legend_side = "right"       # Annotation legend on the left
)
```

Random metadata graph that I need
```{r}
ggplot(metadata_prok, aes(x = "", y = "", fill = Zoox_composition_New)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  scale_fill_brewer(palette = "Dark2") +
  labs(fill = "Symbiont Composition")
```

# AMR analysis

## PCA of AMR genes.

Phyloseq object of AMR genes. 

```{r}
amr.ps <- amr %>% 
  dplyr::select(ORF_ID, Cut_Off, Best_Hit_ARO, Drug_Class, Resistance_Mechanism, AMR_Gene_Family) %>%
  as.matrix

amr.ps %>% head

amr.tax <- amr %>% 
  left_join(prok_taxa, by = c("ORF_ID" = "contig")) %>%
  dplyr::select(ORF_ID, Cut_Off, Best_Hit_ARO, Drug_Class, Resistance_Mechanism, AMR_Gene_Family, annotation, complete_classification) %>%
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = TRUE) %>%
  as.matrix

rownames(amr.ps) <- amr.ps[,1]
rownames(amr.tax) <- amr.tax[,1]

prokreads_nozero$Genes <- NULL

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(amr.tax)

ps.amr <- phyloseq(NUMREADS, META, AMR)
ps.amr
```

Looks like there are 178,736 anti-microbial resistance genes that are non-zero abundance in the dataset. There were a total of 200,100 different genes. So lots to work with. I will make a PCA of only AMR genes. I will make it like I do for other PCAs. 

```{r}
# transform the data using centered log-ratio transformations
ps.amr_clr <- microbiome::transform(ps.amr, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.amr_clr_euc <- ordinate(ps.amr_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
ps.amr.clr_meta <- as(sample_data(ps.amr_clr), "data.frame")
ps.amr.clr_otu <- otu_table(ps.amr_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
ps.amr.euc_diss <- vegdist(ps.amr.clr_otu, method = "euclidean")
```

### Disease
```{r}
sample_data(ps.amr_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.amr_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))
sample_data(ps.amr_clr)$Zoox_background <- factor(sample_data(ps.amr_clr)$Zoox_background, levels = c("none", "C", "CD", "D"))

# plot PCA
disease_pca <- plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary", 
                shape ="Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))
disease_pca

# Make the shape the background symbiont 
plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary", 
                shape ="Zoox_composition_New") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling", shape = "Symbiont Composition") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  scale_shape_manual(values = c(16, 15, 17, 18, 3, 8))

# colony disease at sampling PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.euc_diss ~ Disease_on_colony_when_coring_binary, data = ps.amr.clr_meta, permutations = 999) 
```
### Symbiont composition
```{r}

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Zoox_composition_New", shape = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  ggrepel::geom_text_repel(label = sample_data(ps.amr_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Zoox_composition_New", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

symbiont_pca$layers

symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + geom_point(size=4, alpha=0.8)
symbiont_pca_final

# colony symbiont composition PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.euc_diss ~ Zoox_composition_New, data = ps.amr.clr_meta, permutations = 999) 
```
### Colony Fate
```{r}
sample_data(ps.amr_clr)$Fate_after_SP1 <- 
              factor(sample_data(ps.amr_clr)$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# plot PCA - facet by fate
pca_fate <- plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  facet_wrap(. ~ Fate_after_SP1) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate")
pca_fate

# plot PCA - facet by fate
plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  facet_wrap(. ~ Fate_after_SP1) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate") +
  ggrepel::geom_text_repel(label = sample_data(ps.amr_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6)

# colony disease fate
set.seed(10)
adonis2(formula = ps.amr.euc_diss ~ Fate_after_SP1, data = ps.amr.clr_meta, permutations = 999) 
```

## Stacked bar plot

The majority of AMR genes have an unclassified annotation

```{r}
amr.taxonomy <- tax_table(ps.amr) %>%
  as.data.frame()

# How many fully unclassified taxa?
amr.taxonomy %>%
  dplyr::count(annotation) %>%
  arrange(desc(n)) %>%
  head()

# For visualizing, just graph the AMR genes that show up more than 100 counts in the total dataset
ps.amr.filt <- ps.amr %>%
  filter_taxa(function(x) sum(x) > 100, TRUE)

ps_ra<-transform_sample_counts(ps.amr.filt, function(OTU) OTU/sum(OTU))
ps_ra_nofilt<-transform_sample_counts(ps.amr, function(OTU) OTU/sum(OTU))

# How many different phyla comprise the most abundant amr genes?
tax_table(ps.amr.filt)[,"Phylum"] %>% unique
tax_table(ps.amr.filt)[,"Class"] %>% unique

# How many different phyla comprise the amr genes?
tax_table(ps.amr)[,"Phylum"] %>% unique
tax_table(ps.amr)[,"Class"] %>% unique

plot_bar(ps.amr.filt, x = "CoralID", fill="Class")+
  geom_bar(aes(fill=Class), stat="identity", position = "stack") +
  theme(legend.position = "none") +
  theme(strip.text=element_text(face="bold")) +
  facet_grid(. ~ Disease_on_colony_when_coring_binary, scales = "free", space = "free") 

plot_bar(ps.amr.filt, x = "CoralID", fill="Phylum")+
  geom_bar(aes(fill=Phylum), stat="identity", position = "stack") +
  theme(legend.position = "none") +
  theme(strip.text=element_text(face="bold")) +
  facet_grid(. ~ Disease_on_colony_when_coring_binary, scales = "free", space = "free") 

plot_bar(ps_ra, x = "CoralID", fill="Class")+
  geom_bar(aes(fill=Class), stat="identity", position = "stack") +
  theme(legend.position = "none") +
  theme(strip.text=element_text(face="bold")) +
  facet_grid(. ~ Disease_on_colony_when_coring_binary, scales = "free", space = "free") 

plot_bar(ps_ra, x = "CoralID", fill="Phylum")+
  geom_bar(aes(fill=Phylum), stat="identity", position = "stack") +
  theme(legend.position = "none") +
  theme(strip.text=element_text(face="bold")) +
  facet_grid(. ~ Disease_on_colony_when_coring_binary, scales = "free", space = "free") 

p <- plot_bar(ps.amr.filt, x = "CoralID", fill="Class")+
  geom_bar(aes(fill=Class), stat="identity", position = "stack") +
  theme(strip.text=element_text(face="bold")) +
  facet_grid(. ~ Disease_on_colony_when_coring_binary, scales = "free", space = "free") 

legend <- cowplot::get_legend(p) #show the legend because it isn't printed in the gridded plot
grid.newpage()
grid.draw(legend)
```
Looks like a whopping 129,772 AMR genes don't have a taxonomic ID. I will remove those to get a better look at what taxa are making up the classified AMR genes. 

### Resistance mechanism
It'd be nice to make a stacked bar plot of the general mechanisms. This will plot the top 8, which are making up the majority of the different groupings
```{r}
ps.amr.taxfix <- ps.amr %>%
  transmute_tax_table(Resistance_Mechanism, AMR_Gene_Family, Best_Hit_ARO, ORF_ID) %>%
  tax_fix()

sample_data(ps.amr.taxfix)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.amr.taxfix)$Disease_on_colony_when_coring_binary,
                                                                          levels = c("Unaffected", "Diseased"))

rest.mech <- ps.amr.taxfix %>%
  comp_barplot(tax_level = "Resistance_Mechanism", 
               other_name = "Other resistance mechanism", # set custom name for the "other" category
               facet_by = "Disease_on_colony_when_coring_binary") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Coral Colony ID", fill = "Resistance Mechanism")

tax_table(ps.amr.taxfix)[,1] %>% unique
```

### AMR Gene Family
```{r}
# Make a new phyloseq object with the unique AMR gene family IDs
amr.ps.new <- amr.new %>% 
  dplyr::select(AMR_Gene_Family, AMR_Gene_Family_ID, ORF_ID) %>%
  as.matrix
#Resistance_Mechanism, AMR_Gene_Family, 
amr.ps.new %>% head

rownames(amr.ps.new) <- amr.ps.new[,3]

prokreads_nozero$Genes <- NULL

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(amr.ps.new)

ps.amr.new <- phyloseq(NUMREADS, META, AMR)
ps.amr.new

# Agglomerate by the gene family
ps.genefam <- ps.amr.new %>%
  tax_glom(taxrank = "AMR_Gene_Family_ID") %>% 
  tax_fix()

ps.genefam

sample_data(ps.genefam)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.genefam)$Disease_on_colony_when_coring_binary,
                                                                          levels = c("Unaffected", "Diseased"))


# Visualize the amr gene families, using the ID, since it is shorter. I can match up the actual name later
genefam <- ps.genefam %>%
  comp_barplot(tax_level = "AMR_Gene_Family_ID", 
               other_name = "Other Gene Family", # set custom name for the "other" category
               facet_by = "Disease_on_colony_when_coring_binary", 
               n_taxa = 15,
               merge_other = TRUE) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Coral Colony ID", fill = "AMR Gene Family")

amrfamilymatches <- filter(as.data.frame(amr.ps.new), AMR_Gene_Family_ID %in% c("amrgenefam_5", 
                                                            "amrgenefam_1",
                                                            "amrgenefam_32",
                                                            "amrgenefam_95",
                                                            "amrgenefam_17",
                                                            "amrgenefam_58",
                                                            "amrgenefam_6",
                                                            "amrgenefam_197",
                                                            "amrgenefam_88",
                                                            "amrgenefam_163",
                                                            "amrgenefam_101",
                                                            "amrgenefam_77",
                                                            "amrgenefam_62",
                                                            "amrgenefam_87",
                                                            "amrgenefam_74"
                                                            ))
# Get unique combinations of col1 and col2
unique_combinations <- unique(amrfamilymatches[c("AMR_Gene_Family", "AMR_Gene_Family_ID")])
print(unique_combinations)

# How many gene families are there? 437
tax_table(ps.genefam)[,1] %>% unique
```




## Beta-lactamases

Search through the genes that are beta-lactamases. 

```{r}
amr.new$AMR_Gene_Family %>% head

betalactamases <- amr.new[grep("beta-lactamase", amr.new$AMR_Gene_Family), ]

betalactamases$AMR_Gene_Family %>% unique %>% length
dim(betalactamases)
```

It looks like there are 53,467 genes that are classified as beta-lactamases in the dataset. This includes 253 unique types. 

Combine the beta-lactamases with the taxonomy
```{r}
bl.tax <- betalactamases %>% 
  left_join(prok_taxa, by = c("ORF_ID" = "contig")) %>%
  dplyr::select(ORF_ID, Cut_Off, Best_Hit_ARO, Drug_Class, Resistance_Mechanism, AMR_Gene_Family, annotation, complete_classification) %>%
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = TRUE) %>%
  as.data.frame()

write.table(bl.tax, "../data/beta-lactamase_AMR.txt", sep="\t", row.names = FALSE, col.names = TRUE)

bl.tax %>% head
rownames(bl.tax) <- bl.tax[,1]

# How many fully unclassified taxa?
bl.tax %>%
  dplyr::count(annotation) %>%
  arrange(desc(n))

classonly <- bl.tax %>%
  dplyr::count(Class) %>%
  arrange(desc(n))

phylaonly <- bl.tax %>%
  dplyr::count(Phylum) %>%
  arrange(desc(n))

# Only show the top 10 most abundant Phyla Aggregate all others into "others
plotting_phyla <- phylaonly %>% 
  mutate(rank = rank(-n), 
         Phylum = ifelse(rank <= 10, Phylum, 'Other Phyla'))

# Make a pie chart of the taxonomy - PHYLA

ggplot(plotting_phyla, aes(x = "", y = n, fill = Phylum)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(fill = "Phylum")

# Class level pie chart
plotting_class <- classonly %>%
  mutate(rank = rank(-n),
         Class = ifelse(rank <= 10, Class, 'Other Class')) %>%
  group_by(Class) %>%
  summarise(new_n = sum(n))

safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

betalactam.tax <- ggplot(plotting_class, aes(x = "", y = new_n, fill = Class)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(fill = "Class") +
  scale_fill_manual(values = safe_colorblind_palette)

betalactam.tax
```

### Stacked bar plot
In addition, I want to make a stacked bar plot of only the beta-lactamase genes. What are the different types that are most abundant in the dataset?
```{r}
# Make a new phyloseq object with the unique AMR gene family IDs
bl.new <- betalactamases %>% 
  dplyr::select(AMR_Gene_Family, AMR_Gene_Family_ID, ORF_ID) %>%
  as.matrix
#Resistance_Mechanism, AMR_Gene_Family, 
bl.new %>% head

# How many total gene families and total beta lactamases
bl.new[,1] %>% unique %>% length
bl.new %>% dim()

rownames(bl.new) <- bl.new[,3]

prokreads_nozero$Genes <- NULL

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(bl.new)

bl.psobj <- phyloseq(NUMREADS, META, AMR)
bl.psobj

# 47,892 beta-lactamase genes with positive abundance

# Agglomerate by the gene family #253 taxa
bl.psobj.agg <- bl.psobj %>%
  tax_glom(taxrank = "AMR_Gene_Family") %>% # Not sure how it gets to 441 when there are 437 unique taxa
  tax_fix()

bl.psobj.agg

sample_data(bl.psobj.agg)$Disease_on_colony_when_coring_binary <- factor(sample_data(bl.psobj.agg)$Disease_on_colony_when_coring_binary,
                                                                          levels = c("Unaffected", "Diseased"))

# Visualize the amr gene families, using the ID, since it is shorter. I can match up the actual name later
blactams <- bl.psobj.agg %>%
  comp_barplot(tax_level = "AMR_Gene_Family", 
               other_name = "Other beta-Lactamase gene family", # set custom name for the "other" category
               facet_by = "Disease_on_colony_when_coring_binary", 
               n_taxa = 15,
               merge_other = TRUE) +
  labs(x = "Coral Colony ID", y = "Relative Abundance", fill = "AMR Gene Family") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

blactams

```


### PCA

```{r}
NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(as.matrix(bl.tax))

ps.amr.bl <- phyloseq(NUMREADS, META, AMR)
ps.amr.bl
```
Prep data for the plot
```{r}
# transform the data using centered log-ratio transformations
ps.amr.bl_clr <- microbiome::transform(ps.amr.bl, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.amr.bl_clr_euc <- ordinate(ps.amr.bl_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
ps.amr.bl.clr_meta <- as(sample_data(ps.amr.bl_clr), "data.frame")
ps.amr.bl.clr_otu <- otu_table(ps.amr.bl_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
ps.amr.bl.euc_diss <- vegdist(ps.amr.bl.clr_otu, method = "euclidean")
```

Make the plot
```{r}
sample_data(ps.amr.bl_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.amr.bl_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA
disease_pca <- plot_ordination(ps.amr.bl_clr, ps.amr.bl_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary", 
                shape ="Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))
disease_pca

# Make the shape the background symbiont 
amr_pca <- plot_ordination(ps.amr.bl_clr, ps.amr.bl_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary", 
                shape ="Zoox_composition_New") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling", shape = "Symbiodiniaceae Composition") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
 scale_shape_manual(values = c(16, 15, 17, 18, 3, 8))

amr_pca

# colony disease at sampling PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Disease_on_colony_when_coring_binary, data = ps.amr.bl.clr_meta, permutations = 999) 

# colony zoox PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Zoox_composition_New, data = ps.amr.bl.clr_meta, permutations = 999) 
```

# Amoxicillin treatments
```{r}
metadata_prok$Disease_on_colony_when_coring_binary <- factor(metadata_prok$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

treatments <- ggplot(metadata_prok, aes(x = Disease_on_colony_when_coring_binary,
                          y = Total_treatments_before_coring)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary, 
                  shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Total amoxicillin treatments") +
  theme(legend.text.align = 0) +
  labs(color = "Disease at sampling", x = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))

treatments

plot_ordination(ps.amr.bl_clr, ps.amr.bl_clr_euc, type="samples", 
                color="Total_treatments_before_coring", 
                shape ="Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Total amoxicillin treatments", shape = "Disease at sampling")

# PERMANOVA for total treatments before coring
set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Total_treatments_before_coring, data = ps.amr.bl.clr_meta, permutations = 999) 

# Test between the two groupings for amox treatments
shapiro.test(metadata_prok$Total_treatments_before_coring) # not normal

wilcox.test(Total_treatments_before_coring ~ Disease_on_colony_when_coring_binary, data = metadata_prok, exact = FALSE) # significant difference
```
Total treatments is not significant with the actual permanova. 

# Diversity of beta-lactamases

I would expect that as the number of treatments increases, that the diversity of beta-lactamase genes increases in the coral microbiome in response to the antibiotic pressure. I need to adjust the table so that I can calculate the number of different beta-lactamases in each coral sample. 
```{r}
# Use the agglomerated phyloseq object of just beta lactamases from above
bl.psobj.agg %>% otu_table

# Change levels
sample_data(bl.psobj.agg)$Disease_on_colony_when_coring_binary <- factor(sample_data(bl.psobj.agg)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# Look at the diversity of beta-lactamases (richness)
plot_richness(bl.psobj.agg)

plot_richness(bl.psobj.agg, measures = "Observed", x = "Total_treatments_before_coring", color = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Richness of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

plot_richness(bl.psobj.agg, measures = "Shannon", x = "Total_treatments_before_coring", color = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Shannon diversity of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

plot_richness(bl.psobj.agg, measures = "Observed", x = "Disease_on_colony_when_coring_binary", color = "Disease_on_colony_when_coring_binary") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Richness of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

plot_richness(bl.psobj.agg, measures = "Shannon", x = "Disease_on_colony_when_coring_binary", color = "Disease_on_colony_when_coring_binary") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Shannon diversity of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

colnames(sample_data(bl.psobj.agg))
```

Graph richness but NOT using the agglomerated table. This way you treat each individual beta-lactamase gene as a different gene. This could be different individual genes, even if they are annotated as the same beta-lactamase genes. 
```{r}
# Use the agglomerated phyloseq object of just beta lactamases from above
bl.psobj %>% otu_table

# Change levels
sample_data(bl.psobj)$Disease_on_colony_when_coring_binary <- factor(sample_data(bl.psobj)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# Look at the diversity of beta-lactamases (richness)
plot_richness(bl.psobj)

plot_richness(bl.psobj, measures = "Observed", x = "Total_treatments_before_coring", color = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Richness of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

plot_richness(bl.psobj, measures = "Shannon", x = "Total_treatments_before_coring", color = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Shannon diversity of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

plot_richness(bl.psobj, measures = "Observed", x = "Disease_on_colony_when_coring_binary", color = "Disease_on_colony_when_coring_binary") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Richness of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

plot_richness(bl.psobj, measures = "Shannon", x = "Disease_on_colony_when_coring_binary", color = "Disease_on_colony_when_coring_binary") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Shannon diversity of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")

## MS FIGURE
plot_richness(bl.psobj, measures = "Observed", x = "Total_treatments_before_coring", color = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ylab("Richness of Beta-lactamase genes") +
  xlab("Total amoxicillin treatments before coring") +
  labs(color = "Disease at sampling")
```


Use breakaway to test the richness estimation and covariates
```{r}
richness_bl <- bl.psobj.agg %>% breakaway
plot(richness_bl, physeq=bl.psobj.agg, color="Disease_on_colony_when_coring_binary")

richness_bl.og <- bl.psobj %>% breakaway
plot(richness_bl.og, physeq=bl.psobj, color="Disease_on_colony_when_coring_binary")

# statistical inference
meta.bl <- bl.psobj %>%
  sample_data %>%
  as_tibble %>%
  mutate("sample_names" = bl.psobj %>% sample_names )

combined_richness.bl <- meta.bl %>%
  left_join(summary(richness_bl.og),
            by = "sample_names") %>%
  mutate(Disease_on_colony_when_coring_binary = 
           factor(Disease_on_colony_when_coring_binary,
                  levels=c("Unaffected", "Diseased")))

# Test for an impact of disease on richness
bt_disease_fixed <- betta(formula = estimate ~ Disease_on_colony_when_coring_binary, 
                      ses = error, data = combined_richness.bl)
bt_disease_fixed$table

# Test for an impact of treatments on richness
bt_treatments_fixed <- betta(formula = estimate ~ Total_treatments_before_coring, 
                      ses = error, data = combined_richness.bl)
bt_treatments_fixed$table

# Plot estimated richness (adding conf. intervals is horrible for graphing...so I will keep them in the table)
ggplot(combined_richness.bl, aes(x = Total_treatments_before_coring, y = estimate)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = "gray") +
  geom_point(size = 4, aes(color = Disease_on_colony_when_coring_binary)) +
  geom_smooth(method = "lm", se = FALSE)

richness <- ggplot(combined_richness.bl, aes(x = Total_treatments_before_coring, y = estimate)) +
  geom_point(size = 4, aes(color = Disease_on_colony_when_coring_binary,
                           shape = Disease_on_colony_when_coring_binary)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(x = "Total amoxicillin treatments before coring", y = "Estimated richness of beta-lactamase genes", color = "Disease at Sampling", shape = "Disease at Sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) 
richness
```

# MAIN MS FIG
I would like to put all the beta-lactamase genes together
```{r}
ggarrange(ggarrange(ggarrange(treatments, richness, ncol = 2, common.legend = TRUE, labels = c("a.", "b."), legend = "bottom"), 
          betalactam.tax, widths = c(1.5, 1), labels = c("", "c.")), 
          blactams, nrow = 2, labels = c("", "d."))

wrap_plots(wrap_plots(treatments, richness),
           wrap_plots(betalactam.tax, amr_pca), nrow = 2, heights = c(1,1.3), axis_titles = "collect")

#ggsave("../figures/AMR_beta-lactamase_wPCA.pdf", width = 13, height = 9)
```

### SUPP FIGURE - Graphs together
```{r}
wrap_plots(rest.mech, genefam,blactams, nrow = 3, axis_titles = "collect_x")

#ggsave("../figures/AMR_all_betalactams_stacked_bar.pdf", width = 12, height = 15)
```