---
title: "AMR_RGI_analysis"
output: html_document
date: "2024-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(dplyr)
library(tidyr)
library(readxl)
library(stringr)
library(KEGGREST)
library(speedyseq)

# For visualization and statistical tests
library(phyloseq)
library(randomcoloR)
library(ggplot2)
theme_set(theme_bw())
library(ComplexHeatmap)
library(circlize)  # For color scaling
library(vegan)
library(cowplot)
library(microViz)
library(patchwork) # for arranging groups of plots
library(breakaway)
```

# Read in data

## Load raw data 

First read in the RGI data table. Note, the table that came from the RGI program was kinda tough in the first column. So I changed the first column to have just the contig/gene ID, then got rid of the empty columns (Contig, start, stop, orientation). I also got rid of teh SNPs column and predicted_DNA columns. I also just tried saving it as an Excel document to see if it works better. 
```{r}
amr <- read_excel("../data/FLK_prok_coassembly_RGI_AMR.xlsx")

# Make a short ID that corresponds to the AMR gene family name 
amr.fam <- data.frame("AMR_Gene_Family" = unique(amr$AMR_Gene_Family))
amr.fam$AMR_Gene_Family_ID <- paste0("amrgenefam_", seq_len(nrow(amr.fam)))

amr.new <- amr %>%
  left_join(amr.fam, by = "AMR_Gene_Family")

table(amr.new$AMR_Gene_Family_ID == "NA")
```

Need to read in the count, taxonomy, and metadata data for each gene. 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata
metadata <- as.data.frame(read_excel("../data/RRC_sample_CoreSiteData_v10_RRCCoreData_SP1_Keys_OFAV.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)

# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)

# Read in the taxonomy
prok_taxa <- read.table("../data/FLK_prok_coassembly_tax_mmseqs2.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

## Subset the metadata
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced samples
```

Generate some summary statistics regarding the different AMR categories of all 200,000 hits
```{r}
amr %>% 
  dplyr::count(Model_type)

amr %>% 
  dplyr::count(Cut_Off)

x <- amr %>% 
  dplyr::count(Resistance_Mechanism) 

amr %>% 
  dplyr::count(Drug_Class)

amr %>% 
  dplyr::count(AMR_Gene_Family)

# How many unique genes had AMR "hits"

amr$ORF_ID %>% unique %>% length
```



# AMR analysis

## PCA of AMR genes.

Phyloseq object of AMR genes. 

```{r}
amr.ps <- amr %>% 
  dplyr::select(ORF_ID, Cut_Off, Best_Hit_ARO, Drug_Class, Resistance_Mechanism, AMR_Gene_Family) %>%
  as.matrix

amr.ps %>% head

amr.tax <- amr %>% 
  left_join(prok_taxa, by = c("ORF_ID" = "contig")) %>%
  dplyr::select(ORF_ID, Cut_Off, Best_Hit_ARO, Drug_Class, Resistance_Mechanism, AMR_Gene_Family, annotation, complete_classification) %>%
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = TRUE) %>%
  as.matrix

rownames(amr.ps) <- amr.ps[,1]
rownames(amr.tax) <- amr.tax[,1]

prokreads_nozero$Genes <- NULL

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(amr.tax)

ps.amr <- phyloseq(NUMREADS, META, AMR)
ps.amr
```

Looks like there are 178,736 anti-microbial resistance genes that are non-zero abundance in the dataset. There were a total of 200,100 different genes. So lots to work with. I will make a PCA of only AMR genes. I will make it like I do for other PCAs. 

```{r}
# transform the data using centered log-ratio transformations
ps.amr_clr <- microbiome::transform(ps.amr, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.amr_clr_euc <- ordinate(ps.amr_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
ps.amr.clr_meta <- as(sample_data(ps.amr_clr), "data.frame")
ps.amr.clr_otu <- otu_table(ps.amr_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
ps.amr.euc_diss <- vegdist(ps.amr.clr_otu, method = "euclidean")
```

### Disease
```{r}
sample_data(ps.amr_clr)$Condition_At_Sampling <- factor(sample_data(ps.amr_clr)$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

# plot PCA
disease_pca <- plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Condition_At_Sampling", 
                shape ="Condition_At_Sampling") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Condition at sampling", shape = "Condition at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))
disease_pca

# Make the shape the background symbiont 
plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Condition_At_Sampling", 
                shape ="Zoox_composition_New") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Condition at sampling", shape = "Symbiont Composition") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  scale_shape_manual(values = c(16, 15, 17, 18, 3, 8))

# colony disease at sampling PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.euc_diss ~ Condition_At_Sampling, data = ps.amr.clr_meta, permutations = 999) 
```
### Symbiont composition
```{r}

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Zoox_composition_New", shape = "Condition_At_Sampling") +
  geom_point(size = 4) +
  coord_fixed() +
  ggrepel::geom_text_repel(label = sample_data(ps.amr_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Symbiodiniaceae composition", shape = "Condition at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Zoox_composition_New", shape = "Condition_At_Sampling") +
  coord_fixed() +
  labs(color = "Symbiodiniaceae composition", shape = "Condition at sampling")

symbiont_pca$layers

symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + geom_point(size=4, alpha=0.8)
symbiont_pca_final

# colony symbiont composition PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.euc_diss ~ Zoox_composition_New, data = ps.amr.clr_meta, permutations = 999) 
```
### Colony Fate
```{r}
sample_data(ps.amr_clr)$Fate_after_SP1 <- 
              factor(sample_data(ps.amr_clr)$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# plot PCA - facet by fate
pca_fate <- plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  scale_color_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  labs(color = "Colony Fate")
pca_fate

# plot PCA - facet by fate
plot_ordination(ps.amr_clr, ps.amr_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  scale_color_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  labs(color = "Colony Fate") +
  ggrepel::geom_text_repel(label = sample_data(ps.amr_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6)

# colony disease fate
set.seed(10)
adonis2(formula = ps.amr.euc_diss ~ Fate_after_SP1, data = ps.amr.clr_meta, permutations = 999) 
```

## Stacked bar plot


A whopping 129,772 AMR genes don't have a taxonomic ID. I will remove those to get a better look at what taxa are making up the classified AMR genes. 

### Resistance mechanism
It'd be nice to make a stacked bar plot of the general mechanisms. This will plot the top 8, which are making up the majority of the different groupings
```{r}
ps.amr.taxfix <- ps.amr %>%
  transmute_tax_table(Resistance_Mechanism, AMR_Gene_Family, Best_Hit_ARO, ORF_ID) %>%
  tax_fix()

sample_data(ps.amr.taxfix)$Condition_At_Sampling <- factor(sample_data(ps.amr.taxfix)$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

rest.mech <- ps.amr.taxfix %>%
  comp_barplot(tax_level = "Resistance_Mechanism", 
               other_name = "Other resistance mechanism", # set custom name for the "other" category
               facet_by = "Condition_At_Sampling") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Coral Colony ID", fill = "Resistance Mechanism")

rest.mech

tax_table(ps.amr.taxfix)[,1] %>% unique
```

### AMR Gene Family
```{r}
# Make a new phyloseq object with the unique AMR gene family IDs
amr.ps.new <- amr.new %>% 
  dplyr::select(AMR_Gene_Family, AMR_Gene_Family_ID, ORF_ID) %>%
  as.matrix
#Resistance_Mechanism, AMR_Gene_Family, 
amr.ps.new %>% head

rownames(amr.ps.new) <- amr.ps.new[,3]

prokreads_nozero$Genes <- NULL

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(amr.ps.new)

ps.amr.new <- phyloseq(NUMREADS, META, AMR)
ps.amr.new

# Agglomerate by the gene family
ps.genefam <- ps.amr.new %>%
  tax_glom(taxrank = "AMR_Gene_Family_ID") %>% 
  tax_fix()

ps.genefam

sample_data(ps.genefam)$Condition_At_Sampling <- factor(sample_data(ps.genefam)$Condition_At_Sampling,
                                                        levels = c("Apparently Healthy", "Actively Diseased"))


# Visualize the amr gene families, using the ID, since it is shorter. I can match up the actual name later
genefam <- ps.genefam %>%
  comp_barplot(tax_level = "AMR_Gene_Family_ID", 
               other_name = "Other Gene Family", # set custom name for the "other" category
               facet_by = "Condition_At_Sampling", 
               n_taxa = 15,
               merge_other = TRUE) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Coral Colony ID", fill = "AMR Gene Family")
genefam

amrfamilymatches <- filter(as.data.frame(amr.ps.new), AMR_Gene_Family_ID %in% c("amrgenefam_5", 
                                                            "amrgenefam_1",
                                                            "amrgenefam_32",
                                                            "amrgenefam_95",
                                                            "amrgenefam_17",
                                                            "amrgenefam_58",
                                                            "amrgenefam_6",
                                                            "amrgenefam_197",
                                                            "amrgenefam_88",
                                                            "amrgenefam_163",
                                                            "amrgenefam_101",
                                                            "amrgenefam_77",
                                                            "amrgenefam_62",
                                                            "amrgenefam_87",
                                                            "amrgenefam_74"
                                                            ))
# Get unique combinations of col1 and col2
unique_combinations <- unique(amrfamilymatches[c("AMR_Gene_Family", "AMR_Gene_Family_ID")])
print(unique_combinations)

# How many gene families are there? 437
tax_table(ps.genefam)[,1] %>% unique
```




## Beta-lactamases

Search through the genes that are beta-lactamases. 

```{r}
amr.new$AMR_Gene_Family %>% head

betalactamases <- amr.new[grep("beta-lactamase", amr.new$AMR_Gene_Family), ]

betalactamases$AMR_Gene_Family %>% unique %>% length
dim(betalactamases)
```

It looks like there are 53,467 genes that are classified as beta-lactamases in the dataset. This includes 253 unique types. 

Combine the beta-lactamases with the taxonomy
```{r}
bl.tax <- betalactamases %>% 
  left_join(prok_taxa, by = c("ORF_ID" = "contig")) %>%
  dplyr::select(ORF_ID, Cut_Off, Best_Hit_ARO, Drug_Class, Resistance_Mechanism, AMR_Gene_Family, annotation, complete_classification) %>%
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = TRUE) %>%
  as.data.frame()

#write.table(bl.tax, "../data/beta-lactamase_AMR.txt", sep="\t", row.names = FALSE, col.names = TRUE)

bl.tax %>% head
rownames(bl.tax) <- bl.tax[,1]

# How many fully unclassified taxa?
bl.tax %>%
  dplyr::count(annotation) %>%
  arrange(desc(n))

classonly <- bl.tax %>%
  dplyr::count(Class) %>%
  arrange(desc(n))

phylaonly <- bl.tax %>%
  dplyr::count(Phylum) %>%
  arrange(desc(n))

# Only show the top 10 most abundant Phyla Aggregate all others into "others
plotting_phyla <- phylaonly %>% 
  mutate(rank = rank(-n), 
         Phylum = ifelse(rank <= 10, Phylum, 'Other Phyla'))

# Class level pie chart
plotting_class <- classonly %>%
  mutate(rank = rank(-n),
         Class = ifelse(rank <= 10, Class, 'Other Class')) %>%
  group_by(Class) %>%
  summarise(new_n = sum(n))

safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

betalactam.tax <- ggplot(plotting_class, aes(x = "", y = new_n, fill = Class)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(fill = "Class") +
  scale_fill_manual(values = safe_colorblind_palette)

betalactam.tax
```

### Stacked bar plot
In addition, I want to make a stacked bar plot of only the beta-lactamase genes. What are the different types that are most abundant in the dataset?
```{r}
# Make a new phyloseq object with the unique AMR gene family IDs
bl.new <- betalactamases %>% 
  dplyr::select(AMR_Gene_Family, AMR_Gene_Family_ID, ORF_ID) %>%
  as.matrix
#Resistance_Mechanism, AMR_Gene_Family, 
bl.new %>% head

# How many total gene families and total beta lactamases
bl.new[,1] %>% unique %>% length
bl.new %>% dim()

rownames(bl.new) <- bl.new[,3]

prokreads_nozero$Genes <- NULL

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(bl.new)

bl.psobj <- phyloseq(NUMREADS, META, AMR)
bl.psobj

# 47,892 beta-lactamase genes with positive abundance

# Agglomerate by the gene family #253 taxa
bl.psobj.agg <- bl.psobj %>%
  tax_glom(taxrank = "AMR_Gene_Family") %>% # Not sure how it gets to 441 when there are 437 unique taxa
  tax_fix()

bl.psobj.agg

sample_data(bl.psobj.agg)$Condition_At_Sampling <- factor(sample_data(bl.psobj.agg)$Condition_At_Sampling,
                                                        levels = c("Apparently Healthy", "Actively Diseased"))

# Visualize the amr gene families, using the ID, since it is shorter. I can match up the actual name later
blactams <- bl.psobj.agg %>%
  comp_barplot(tax_level = "AMR_Gene_Family", 
               other_name = "Other beta-Lactamase gene family", # set custom name for the "other" category
               facet_by = "Condition_At_Sampling", 
               n_taxa = 15,
               merge_other = TRUE) +
  labs(x = "Coral Colony ID", y = "Relative Abundance", fill = "AMR Gene Family") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

blactams

```


### PCA

```{r}
NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
AMR = tax_table(as.matrix(bl.tax))

ps.amr.bl <- phyloseq(NUMREADS, META, AMR)
ps.amr.bl
```
Prep data for the plot
```{r}
# transform the data using centered log-ratio transformations
ps.amr.bl_clr <- microbiome::transform(ps.amr.bl, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.amr.bl_clr_euc <- ordinate(ps.amr.bl_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
ps.amr.bl.clr_meta <- as(sample_data(ps.amr.bl_clr), "data.frame")
ps.amr.bl.clr_otu <- otu_table(ps.amr.bl_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
ps.amr.bl.euc_diss <- vegdist(ps.amr.bl.clr_otu, method = "euclidean")
```

Make a plot
```{r}
sample_data(ps.amr.bl_clr)$Condition_At_Sampling <- factor(sample_data(ps.amr.bl_clr)$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

sample_data(ps.amr.bl_clr)$Fate_after_SP1 <- 
              factor(sample_data(ps.amr.bl_clr)$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))


# Make the shape the background symbiont 
plot_ordination(ps.amr.bl_clr, ps.amr.bl_clr_euc, type="samples", 
                color="Fate_after_SP1", 
                shape ="Zoox_composition_New") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Colony Fate", shape = "Symbiodiniaceae Composition") +
  scale_color_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  scale_shape_manual(values = c(16, 15, 17, 18, 3, 8))


# colony condition at sampling PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Condition_At_Sampling, data = ps.amr.bl.clr_meta, permutations = 999) 

# colony zoox PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Zoox_composition_New, data = ps.amr.bl.clr_meta, permutations = 999) 

# colony fate PERMANOVA
set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Fate_after_SP1, data = ps.amr.bl.clr_meta, permutations = 999) 
```

# Amoxicillin treatments

I am interested in how the number of treatments in different colonies varies and is potentially influencing the number of beta-lactamase genes, which would confer resistance on colonies. 
Some important variables:
- Recent_Treatment_Less_than_two_months (was the treatment recent? In some cases, corals never had treatments.)
- Estimated_Num_treatments_Per_sq_m (scaled to size, which makes the treatments a bit more comparable across colonies)
- Number_treatments_PreSampling (treatments leading up to the sampling in May/June 2021)
- Colony fate = what did the colony end up doing (potentially related to treatments)
```{r}
metadata_prok$Fate_after_SP1 <- factor(metadata_prok$Fate_after_SP1, levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

sample_data(ps.amr.bl_clr)$Recent_Treatment_Less_than_two_months <- factor(sample_data(ps.amr.bl_clr)$Recent_Treatment_Less_than_two_months, levels = c("No", "Yes", "NoTreatments"))


plot_ordination(ps.amr.bl_clr, ps.amr.bl_clr_euc, type="samples", 
                color="Estimated_Num_treatments_Per_sq_m", 
                shape ="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Amoxicilling treatments per sq m.", shape = "Colony Fate")

plot_ordination(ps.amr.bl_clr, ps.amr.bl_clr_euc, type="samples", 
                color="Number_treatments_PreSampling", 
                shape ="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Amoxicillin treatments pre-sampling", shape = "Colony Fate")

amr_pca <- plot_ordination(ps.amr.bl_clr, ps.amr.bl_clr_euc, type="samples", 
                color="Recent_Treatment_Less_than_two_months", 
                shape ="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Recent treatment (<2 mo.)", shape = "Colony Fate")+
  scale_shape_manual(values = c(16, 15, 17, 18)) +
  scale_color_manual(values = c("darkgreen", "orange", "gray"))

amr_pca 

# PERMANOVA for amoxicillin treatments
set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Estimated_Num_treatments_Per_sq_m, data = ps.amr.bl.clr_meta, permutations = 999) 

set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Number_treatments_PreSampling, data = ps.amr.bl.clr_meta, permutations = 999) 

set.seed(10)
adonis2(formula = ps.amr.bl.euc_diss ~ Recent_Treatment_Less_than_two_months, data = ps.amr.bl.clr_meta, permutations = 999) 
```
What if I only subset corals that had any history of treatment? Remove all corals that were not treated at ALL over the course of the study?

```{r}
# Subset to only corals with any treatment history
ps.amr.bl.treated <- subset_samples(ps.amr.bl, Recent_Treatment_Less_than_two_months != "NoTreatments")

# transform the data using centered log-ratio transformations
ps.amr.bl.treated_clr <- microbiome::transform(ps.amr.bl.treated, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.amr.bl.treated_clr_euc <- ordinate(ps.amr.bl.treated_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
ps.amr.bl.treated.clr_meta <- as(sample_data(ps.amr.bl.treated_clr), "data.frame")
ps.amr.bl.treated.clr_otu <- otu_table(ps.amr.bl.treated_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
ps.amr.bl.treated.euc_diss <- vegdist(ps.amr.bl.treated.clr_otu, method = "euclidean")

sample_data(ps.amr.bl_clr)$Recent_Treatment_Less_than_two_months <- factor(sample_data(ps.amr.bl_clr)$Recent_Treatment_Less_than_two_months, levels = c("No", "Yes"))

# PCAs on only corals that had some history of antibiotic treatment
plot_ordination(ps.amr.bl.treated_clr, ps.amr.bl.treated_clr_euc, type="samples", 
                color="Estimated_Num_treatments_Per_sq_m", 
                shape ="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Amoxicilling treatments per sq m.", shape = "Colony Fate")

plot_ordination(ps.amr.bl.treated_clr, ps.amr.bl.treated_clr_euc, type="samples", 
                color="Number_treatments_PreSampling", 
                shape ="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Amoxicillin treatments pre-sampling", shape = "Colony Fate")

plot_ordination(ps.amr.bl.treated_clr, ps.amr.bl.treated_clr_euc, type="samples", 
                color="Recent_Treatment_Less_than_two_months", 
                shape ="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Recent treatment (<2 mo.)", shape = "Colony Fate")


# PERMANOVA for amoxicillin treatments
set.seed(10)
adonis2(formula = ps.amr.bl.treated.euc_diss ~ Estimated_Num_treatments_Per_sq_m, data = ps.amr.bl.treated.clr_meta, permutations = 999) 

set.seed(10)
adonis2(formula = ps.amr.bl.treated.euc_diss ~ Number_treatments_PreSampling, data = ps.amr.bl.treated.clr_meta, permutations = 999) 

set.seed(10)
adonis2(formula = ps.amr.bl.treated.euc_diss ~ Recent_Treatment_Less_than_two_months, data = ps.amr.bl.treated.clr_meta, permutations = 999) 
```
Honestly, the results are worse when I relate the whole composition to only corals with some history of antibiotic treatment. 

# Diversity of beta-lactamases

I would expect that as the number of treatments increases, that the diversity of beta-lactamase genes increases in the coral microbiome in response to the antibiotic pressure. I need to adjust the table so that I can calculate the number of different beta-lactamases in each coral sample. 

Graph richness but NOT using the agglomerated table. This way you treat each individual beta-lactamase gene as a different gene. This could be different individual genes, even if they are annotated as the same beta-lactamase genes. 

Use breakaway to test the richness estimation and covariates
```{r}
# Change levels
sample_data(bl.psobj)$Condition_At_Sampling <- factor(sample_data(bl.psobj.agg)$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

sample_data(bl.psobj)$Fate_after_SP1 <- factor(sample_data(bl.psobj.agg)$Fate_after_SP1, levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

richness_bl.og <- bl.psobj %>% breakaway
plot(richness_bl.og, physeq=bl.psobj, color="Condition_At_Sampling")

# statistical inference
meta.bl <- bl.psobj %>%
  sample_data %>%
  as_tibble %>%
  mutate("sample_names" = bl.psobj %>% sample_names )

combined_richness.bl <- meta.bl %>%
  left_join(summary(richness_bl.og),
            by = "sample_names") %>%
  mutate(Condition_At_Sampling = 
           factor(Condition_At_Sampling,
                  levels=c("Apparently Healthy", "Actively Diseased"))) %>%
  mutate(Recent_Treatment_Less_than_two_months = 
           factor(Recent_Treatment_Less_than_two_months,
                  levels=c("No", "Yes", "NoTreatments"))) %>%
  mutate(Fate_after_SP1 = 
           factor(Fate_after_SP1,
                  levels=c("Recovered", "Unaffected", "Will be diseased", "Diseased")))

# Test for an impact of disease on richness
bt_disease_fixed <- betta(formula = estimate ~ Fate_after_SP1, 
                      ses = error, data = combined_richness.bl)
bt_disease_fixed$table

# Test for an impact of treatments on richness - scaled by size
bt_treatments_fixed <- betta(formula = estimate ~ Estimated_Num_treatments_Per_sq_m, 
                      ses = error, data = combined_richness.bl)
bt_treatments_fixed$table

# Test for an impact of treatments on richness
bt_treatments_fixed <- betta(formula = estimate ~ Number_treatments_PreSampling, 
                      ses = error, data = combined_richness.bl)
bt_treatments_fixed$table

# Test for an impact of treatment history (recent = <2 months, versus a while ago)
bt_treatments_fixed <- betta(formula = estimate ~ Recent_Treatment_Less_than_two_months, 
                      ses = error, data = combined_richness.bl)
bt_treatments_fixed$table

# Plot estimated richness 
richness <- ggplot(combined_richness.bl, aes(x = Total_treatments_before_coring, y = estimate)) +
  geom_point(size = 4, aes(color = Recent_Treatment_Less_than_two_months,
                           shape = Fate_after_SP1)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(x = "Total amoxicillin treatments before sampling", y = "Estimated richness of beta-lactamase genes", color = "Recent treatment (<2 mo.)", shape = "Colony Fate") +
  scale_shape_manual(values = c(16, 15, 17, 18)) +
  scale_color_manual(values = c("darkgreen", "orange", "gray"))
richness
```

# MAIN MS FIG
I would like to put all the beta-lactamase genes together
```{r}
#ggarrange(ggarrange(ggarrange(treatments, richness, ncol = 2, common.legend = TRUE, labels = c("a.", "b."), legend = "bottom"), 
#          betalactam.tax, widths = c(1.5, 1), labels = c("", "c.")), 
#          blactams, nrow = 2, labels = c("", "d."))

wrap_plots(wrap_plots(richness, amr_pca, guides = "collect", widths = c(1,1.5)),
           wrap_plots(betalactam.tax), nrow = 2, heights = c(1,1.3), axis_titles = "collect")

ggsave("../figures/AMR_beta-lactamase_wPCA_08.07.25.pdf", width = 12, height = 8)
```

### SUPP FIGURE - Graphs together
```{r}
wrap_plots(rest.mech, genefam,blactams, nrow = 3, axis_titles = "collect_x")

#ggsave("../figures/AMR_all_betalactams_stacked_bar_08.07.25.pdf", width = 12, height = 15)
```