---
title: "BGC_analysis"
output: html_document
date: "2024-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(dplyr)
library(tidyr)
library(readxl)
library(stringr)
library(KEGGREST)

# For visualization
library(phyloseq)
library(randomcoloR)
library(ggplot2)
theme_set(theme_bw())
library(ComplexHeatmap)
library(circlize)  # For color scaling
library(RColorBrewer)
```

# Read in data

## Load raw data 

First read in the RGI data table. Note, the table that came from the RGI program was kinda tough in the first column. So I changed the first column to have just the contig/gene ID, then got rid of the empty columns (Contig, start, stop, orientation). I also got rid of teh SNPs column and predicted_DNA columns. I also just tried saving it as an Excel document to see if it works better. 
```{r}
bgc <- read_excel("../data/FLK_prok_coassembly_antismash_BGC.xlsx")
```

Need to read in the count, taxonomy, and metadata data for each gene. 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)

# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)

# Read in the taxonomy
prok_taxa <- read.table("../data/FLK_prok_coassembly_tax_mmseqs2.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

## Subset the metadata
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced samples
```
# Explore
Explore the results

```{r}
n <- 30
colrs <- brewer.pal.info[brewer.pal.info$colorblind == TRUE, ]
col_vec = unlist(mapply(brewer.pal, colrs$maxcolors, rownames(colrs)))
col <- sample(col_vec, n)

ggplot(bgc, aes(x = "", fill = Type)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = col)

ggplot(bgc, aes(x = On_contig_edge, fill = Type)) +
  geom_bar()+
  scale_fill_manual(values = col)
```

Let's look at the individual BGCs. They are on full contigs, but all my annotations are on the predicted genes. I will investigate the annotations for the genes from each contig. 
First, start with some data manipulation of the taxa. 
```{r}
prok_contig_taxa <- prok_taxa %>%
  mutate(Gene = contig) %>% 
  separate(contig, into = c("base", "contig", "gene_ID" ), sep = "_", remove = FALSE) %>% 
  mutate(FullContig = paste0(base, "_", contig))
```

Now that I have the data in a format I like. I will investigate the 3 full BGC taxonomic annotations. 

### Terpene BGC
The terpene is on contig k127_167047
```{r}
terpene_tax <- prok_contig_taxa %>% filter(FullContig == "k127_167047") %>%
  separate(complete_classification, into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";")

ggplot(terpene_tax, aes(x = "", y = n, fill = Family)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()
```
### Opine-like-metallophore BGC
The metallophore is on contig k127_491492
This one is interesting. It is a BGC for a zincophore, for harvesting zinc!!!
```{r}
metallophore_tax <- prok_contig_taxa %>% filter(FullContig == "k127_491492") %>%
  separate(complete_classification, into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";")

ggplot(metallophore_tax, aes(x = "", y = n, fill = Family)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()
```
### RRE-containing BGC
The RRE BGC is on contig k127_3016148
This one is interesting. It is a BGC for a zincophore, for harvesting zinc!!!
```{r}
RRE_tax <- prok_contig_taxa %>% filter(FullContig == "k127_3016148") %>%
  separate(complete_classification, into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";")

ggplot(RRE_tax, aes(x = "", y = n, fill = Family)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()
```



