---
title: "Metadata_code_to_reproduce_Figure_1"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(dplyr)
library(readxl)
library(stringr)
library(tidyr)
library(tidyverse)
theme_set(theme_bw())
library(ggpubr)
library(ggalluvial)
library(scales)
```

## Read in data
```{r}
# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID
```


# MS FIGURE Alluvial plot - three timepoints
I want to make an alluvial plot, but with pre-disease, disease at sampling, and disease after sampling, and color it by colony fate. The point of this plot is to demonstrate that we tracked these colonies over a long period of time, but that the corals were sampled only during one time period. 

```{r}
# Subset only the FLK data first
FLK_data <- metadata %>%
  filter(Region == "Lower Keys") %>% # only look at FLK samples
  filter(Spp == "Ofav") %>% # remove ofra colonies since I didn't analyze their metagenomes
  filter(CoralID != "N49") %>% # remove the coral colony that wasn't sequenced
  dplyr::select(CoralID, Disease_before_SP1, Disease_on_colony_when_coring_binary, Disease_after_SP1, Fate_after_SP1) %>%
  group_by(Disease_before_SP1, Disease_on_colony_when_coring_binary, Disease_after_SP1, Fate_after_SP1) %>%
  dplyr::summarise(Count = n()) %>%
  ungroup()

FLK_data$Fate_after_SP1 <- factor(FLK_data$Fate_after_SP1, levels = c("Unaffected", "Recovered", "Will be diseased", "Diseased"))

addline_format <- function(x,...){
    gsub("\\(", "\n(", x)
}

ggplot(FLK_data,
       aes(axis1 = Disease_before_SP1, axis2 = Disease_on_colony_when_coring_binary, axis3 = Disease_after_SP1,
           y = Count)) +
  geom_alluvium(aes(fill = Fate_after_SP1), alpha = 0.9) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Pre-collection condition", "Disease at collection", "Post-collection condition"),
                   expand = c(0.15, 0.05),
                   labels = addline_format(c("Pre-collection condition (2019 - Apr 2021)", 
                        "Disease at collection (May-June 2021)", "Post-collection condition (July 2021 - July 2023)"))) +
  labs(y = "Number of coral colonies", fill = "Colony Fate") +
  theme_minimal() +
  scale_fill_manual(values = c("black", "gray20", "gray70", "gray90"))

#ggsave("../figures/alluvial_plot_fate.pdf", width = 9, height = 5)
```

Calculate numbers to go in the manuscript
```{r}
FLK_data

20/41 # 20 colonies diseased at sampling
(3+3+4+11)/41 # 21 colonies unaffected at sampling

(20+3+3)/41 # colonies that were diseased at sampling and diseased sometime after SP1 out of total colonies

(3 + 3)/41 # colonies unaffected at sampling, but got disease after SP1 (will be diseased) out of total colonies

4/41 # colonies that recoverd from sctld out of all colonies

11/41 # colonies never impacted during monitoring out of all colonies
```

# SUPPLEMENT - SYMBIONT COMPOSITION
I would like to show a figure displaying the abundance of different symbiont composition groupings within the program. 
I could do this using an overall pie chart. I could also make a stacked bar plot. 
```{r}
FLK_metadata <- metadata %>%
  filter(Region == "Lower Keys") %>% # only look at FLK samples
  filter(Spp == "Ofav") %>% # remove ofra colonies since I didn't analyze their metagenomes
  filter(CoralID != "N49") # remove the coral colony that wasn't sequenced

FLK_metadata$Zoox_composition

# There are a mix of zooxanthellae
ggplot(FLK_metadata, aes(x = "", y = "", fill = Zoox_background)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()

# one coral has Cladocopium zoox
ggplot(FLK_metadata, aes(x = "", y = "", fill = Zoox_dominant)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()

# Zoox_composition
ggplot(FLK_metadata, aes(x = "", y = "", fill = Zoox_composition)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()

# can I make a plot of both?
FLK_zoox <- FLK_metadata %>% 
  dplyr::select(CoralID, Zoox_composition, Breviolum_density, Cladocopium_density, Durusdinium_density) %>%
  gather(key = "Symbiont", value = "Symbiont_Density", Breviolum_density:Durusdinium_density)

ggplot(FLK_zoox, aes(x = CoralID, y = Symbiont_Density, fill = Symbiont)) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(. ~ Zoox_composition, scales = "free")

ggplot(FLK_zoox, aes(x = CoralID, y = Symbiont_Density, fill = Symbiont)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(. ~ Zoox_composition, scales = "free")
```
Shit...what I am learning here is that there are HUGE errors in the metadata columns associated with symbiont composition. What I am going to do next is RECALCULATE THE columns that I would need. 
I will start by playing around only with the symbiont relevant columns
Look at this to confirm that I made the correct designation (dominant symbiont and minor symbiont, if applicable)
```{r}
zoox_comp <- FLK_metadata %>% 
  dplyr::select(CoralID, Zoox_dominant, Zoox_background, Total_Symb_Density, Breviolum_density, Cladocopium_density, Durusdinium_density, Zoox_composition) %>%
  mutate(Symb_density_check = Breviolum_density + Cladocopium_density + Durusdinium_density) %>%
  mutate(symb_density_match = ifelse(Total_Symb_Density == Symb_density_check, "good", "ERROR")) %>% # Looks like everything was correctly calculated
  mutate(B_prop = Breviolum_density / Total_Symb_Density) %>%
  mutate(C_prop = Cladocopium_density / Total_Symb_Density) %>%
  mutate(D_prop = Durusdinium_density / Total_Symb_Density) %>% 
  dplyr::select(Zoox_dominant, Zoox_background, B_prop, C_prop, D_prop, Zoox_composition) %>% 
  mutate(B_main = ifelse(B_prop == 1, "B", "other"))
zoox_comp
```
Everything looks correct now! 

## SUPPLEMENTARY FIGURE
I want to show the composition of the symbiotic community in these corals. I will do this by using faceting and also showing the overall results. 
```{r}
# Pie chart of the different corals and their composition
pie <- ggplot(FLK_metadata, aes(x = "", y = "", fill = Zoox_composition)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")

# can I make a stacked bar plot of the actual symbiont composition?
FLK_zoox <- FLK_metadata %>% 
  dplyr::select(CoralID, Zoox_composition, Disease_on_colony_when_coring_binary, Breviolum_density, Cladocopium_density, Durusdinium_density) %>%
  gather(key = "Symbiont", value = "Symbiont_Density", Breviolum_density:Durusdinium_density)

barplot <- ggplot(FLK_zoox, aes(x = CoralID, y = Symbiont_Density, fill = Symbiont)) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(. ~ Disease_on_colony_when_coring_binary + Zoox_composition, scales = "free_x", ncol = 5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("gray80", "gray60", "black")) +
  theme(legend.position = "bottom")
```

Put the plots together
```{r}
ggarrange(pie, barplot, widths = c(1,1.5), labels = c("a.", "b."))
#ggsave("../figures/zoox_composition.pdf", width = 12, height = 5)

table(FLK_metadata$Zoox_composition)

(5+6+8)/41

table(FLK_metadata$treat_type)

summary(FLK_metadata$Total_treatments_before_coring)
```

