---
title: "Betadiversity_code_to_reproduce_Figure_2"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(readxl)
library(stringr)
library(dplyr)
library(tidyr)

# For stats and visualization
library(phyloseq)
library(vegan)
library(microbiome)
library(ggplot2)
theme_set(theme_bw())
library(ggpubr)
library(devtools)
library(pairwiseAdonis)
```

# Load Data

## Read in tables, remove zeros
```{r}
# Load table with count abundances from salmon ("numreads")
numreads <- read.table("../data/FLK_coassembly_salmon_quant_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
numreads <- numreads[, -1] # get rid of the first column cause it is just numbered lines

# Load the table with taoxnomic information from mmseqs2
taxa <- read.table("../data/FLK_coassembly_tax_mmseqs2_lca.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

# Load the table with functional information from eggnog mapper
fxn <- read_excel("../data/FLK_coassembly_fxn_eggnog.xlsx", skip = 2, na = "-")

# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many contigs are zero across all samples?
table(rowSums(numreads) == 0)

# remove them
numreads_nozero <- numreads[rowSums(numreads) != 0, ]
dim(numreads_nozero)

# any samples with no reads?
table(colSums(numreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(numreads_nozero) <- str_replace_all(colnames(numreads_nozero), pattern = "[X]", "")
colnames(numreads_nozero) <- str_split_i(colnames(numreads_nozero), "[_]", 1)
```

## Make phyloseq object of count-related data 
```{r}
idx <- match(colnames(numreads_nozero), rownames(metadata))
metadata_3 <- metadata[idx,] # subset the metadata to only the sequenced samples

NUMREADS = otu_table(numreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_3)

ps.count <- phyloseq(NUMREADS, META)
ps.count

# save phyloseq object for easy use in the future
saveRDS(ps.count, "../data/FLK_coassembly_phyloseq_count.rds")
```

## Filter ps by abundance
```{r}
# Just to get fewer contigs/genes/taxa, sum > 10 number of reads total to remove the more sparse contigs
ps.count.filt <- filter_taxa(ps.count, function(x) sum(x) > 10, TRUE) # 1,743,873 contigs left
ps.count.filt
```

## Save PS object for quicker code 
```{r}
saveRDS(ps.count.filt, "../data/FLK_coassembly_phyloseq_count.filt.rds")
```

# Load PS object 
NOTE: Only do this to read in the RDS phyloseq object that you saved if it already exists. 
```{r}
ps.count.filt <- readRDS("../data/FLK_coassembly_phyloseq_count.filt.rds")

# transform the data using centered log-ratio transformations
ps.count.filt_clr <- microbiome::transform(ps.count.filt, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.count.filt_clr_euc <- ordinate(ps.count.filt_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
count.clr_meta <- as(sample_data(ps.count.filt_clr), "data.frame")
count.clr_otu <- otu_table(ps.count.filt_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
count.euc_diss <- vegdist(count.clr_otu, method = "euclidean")
```

# Beta Diversity Analysis

## Fate
### PCA fate
```{r}
sample_data(ps.count.filt_clr)$Fate_after_SP1 <- 
              factor(sample_data(ps.count.filt_clr)$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# plot PCA - facet by fate
pca_fate <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  facet_wrap(. ~ Fate_after_SP1) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate")
pca_fate

# Conduct statistical tests on Fate 
# Use the dissimilarity matrix from where you read in the ps object
set.seed(10)
adonis2(formula = count.euc_diss ~ Fate_after_SP1, data = count.clr_meta, permutations = 999) 
```

### betadisper fate
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.fate <- betadisper(count.euc_diss, count.clr_meta$Fate_after_SP1)

# save the distances and combine with metadata
distances <- disp.fate$distances
disp.fate.meta <- cbind(distances, count.clr_meta)

# Set the levels so the graphs match
disp.fate.meta$Fate_after_SP1 <- factor(disp.fate.meta$Fate_after_SP1, levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# Are the data normally distributed in each group?
ggqqplot(disp.fate.meta$distances) # not normal looking
shapiro.test(disp.fate.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Fate_after_SP1, data = disp.fate.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.fate.meta$distances, disp.fate.meta$Fate_after_SP1, p.adjust.method = "BH") 

# Plot the results
betadisp_fate <- ggplot(disp.fate.meta, aes(x = Fate_after_SP1, y = distances)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Fate_after_SP1), size = 4) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate", shape = "Reef", x = "Colony Fate")

betadisp_fate
```

### Supp. Figure - Fate
```{r}
ggarrange(pca_fate, betadisp_fate, labels = c("a.", "b."), common.legend = TRUE)
ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

### PCA fate subset
```{r}
# subset the unaffected and diseased fates (no recovered or will be diseased)
ps.count.filt_clr_subset <- subset_samples(ps.count.filt_clr, Fate_after_SP1 %in% c("Unaffected", "Diseased"))

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.count.filt_clr_euc_subset <- ordinate(ps.count.filt_clr_subset, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
count.clr_meta_subset <- as(sample_data(ps.count.filt_clr_subset), "data.frame")
count.clr_otu_subset <- otu_table(ps.count.filt_clr_subset) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
count.euc_diss_subset <- vegdist(count.clr_otu_subset, method = "euclidean")

sample_data(ps.count.filt_clr_subset)$Fate_after_SP1 <- 
              factor(sample_data(ps.count.filt_clr_subset)$Fate_after_SP1, 
                                 levels = c("Unaffected", "Diseased"))

# plot PCA - facet by fate
pca_fate <- plot_ordination(ps.count.filt_clr_subset, ps.count.filt_clr_euc_subset, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  scale_color_manual(values = c("#648FFF", "#DC267F")) +
  labs(color = "Colony Fate")
pca_fate

# Conduct statistical tests on Fate 
# Use the dissimilarity matrix from where you read in the ps object
set.seed(10)
adonis2(formula = count.euc_diss_subset ~ Fate_after_SP1, data = count.clr_meta_subset, permutations = 999) 
```

## Symbiodiniaceae background
### PCA symbiont
```{r}
sample_data(ps.count.filt_clr)$Zoox_background <- factor(sample_data(ps.count.filt_clr)$Zoox_background, levels = c("none", "C", "CD", "D"))
sample_data(ps.count.filt_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.count.filt_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  ggrepel::geom_text_repel(label = sample_data(ps.count.filt_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

symbiont_pca$layers

symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + geom_point(size=4, alpha=0.8)
symbiont_pca_final

# Conduct statistical tests on the background symbiont
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Zoox_background, data = count.clr_meta, permutations = 999) 

#ggsave("../figures/PCA_zoox.png", width = 6, height = 4)
```
Note that for the symbiodiniaceae results, sample 3421 is the one colony with a dominant genus different (cladocopium) than all the others (breviolum). That coral is all the way out on its own in terms of a functional microbiome and was also consistently diseased. 

### betadisper symbiont
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.symbiont <- betadisper(count.euc_diss, count.clr_meta$Zoox_background)

# save the distances and combine with metadata
distances <- disp.symbiont$distances
disp.symbiont.meta <- cbind(distances, count.clr_meta)

# Set the levels so the graphs match
disp.symbiont.meta$Zoox_background <- factor(disp.symbiont.meta$Zoox_background, levels = c("none", "C", "CD", "D"))
disp.symbiont.meta$Disease_on_colony_when_coring_binary <-
  factor(disp.symbiont.meta$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# Are the data normally distributed in each group?
ggqqplot(disp.symbiont.meta$distances) # not normal looking
shapiro.test(disp.symbiont.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Zoox_background, data = disp.symbiont.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.symbiont.meta$distances, disp.symbiont.meta$Zoox_background, p.adjust.method = "BH") 

# Plot the results
betadisp_symbiont <- ggplot(disp.symbiont.meta, aes(x = Zoox_background, y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Zoox_background, shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Background Symbiodiniaceae", x = "Background Symbiodiniaceae", shape = "Disease at sampling")

betadisp_symbiont
```
### Figure - Symbiodiniaceae
```{r}
ggarrange(symbiont_pca_final, betadisp_symbiont, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

## Clone
### PCA clone
```{r}
# plot PCA
clone_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Clone", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  facet_wrap(. ~ Clone) +
  ggrepel::geom_text_repel(label = sample_data(ps.count.filt_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(shape = "Disease at sampling")


clone_pca$layers

clone_pca$layers <- clone_pca$layers[-1]

clone_pca_final <- clone_pca + geom_point(size=4, alpha=0.8)
clone_pca_final

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Clone, data = count.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_clone.png", width = 6, height = 4)
```
## Location
### PCA location
```{r}
# plot PCA
location_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Location", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Reef", shape = "Disease at sampling")

location_pca$layers

location_pca$layers <- location_pca$layers[-1]

location_pca_final <- location_pca + geom_point(size=4, alpha=0.8)
location_pca_final

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Location, data = count.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_reef.png", width = 6, height = 4)
```


## Disease at sampling
### PCA disease at sampling
```{r}
sample_data(ps.count.filt_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.count.filt_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA
disease_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary", 
                shape ="Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))
disease_pca

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Disease_on_colony_when_coring_binary, data = count.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_disease_binary.png", width = 6, height = 4)
```

### betadisper disease at sampling
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.disease <- betadisper(count.euc_diss, count.clr_meta$Disease_on_colony_when_coring_binary)

# save the distances and combine with metadata
distances <- disp.disease$distances
disp.disease.meta <- cbind(distances, count.clr_meta)

# Are the data normally distributed in each group?
ggqqplot(disp.disease.meta$distances) # not normal looking
shapiro.test(disp.disease.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Disease_on_colony_when_coring_binary, data = disp.disease.meta)

# Using the Wilcoxon rank sum test to examine if there is a significant difference between unaffected and diseased coral microbiomes:
wilcox.test(distances ~ Disease_on_colony_when_coring_binary, 
            data = disp.disease.meta) 

# Set levels
disp.disease.meta$Disease_on_colony_when_coring_binary <- factor(disp.disease.meta$Disease_on_colony_when_coring_binary, 
                                                                 levels = c("Unaffected", "Diseased"))

# Plot the results
betadisp_disease <- ggplot(disp.disease.meta, aes(x = Disease_on_colony_when_coring_binary, 
                                                  y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary, 
                  shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Background Symbiodiniaceae", x = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))


betadisp_disease
```

### Figure - Disease at sampling
```{r}
ggarrange(disease_pca, betadisp_disease, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

## Figure - disease at sampling and symbiodiniaceae
```{r}
ggarrange(ggarrange(disease_pca, betadisp_disease, 
                    labels = c("a.", "b."), common.legend = TRUE), 
          ggarrange(symbiont_pca_final, betadisp_symbiont, 
                    labels = c("c.", "d."), common.legend = TRUE, legend = "bottom"), 
          nrow = 2)
ggsave("../figures/PCA_disease_symbiont.pdf", width = 8, height = 8)
```

# Beta Diversity - Bacteria only 
Since we are ultimately doing so many filtering steps to remove eukaryotic genetic signatures, it doesn't make sense to include them in the permanova and PCA-based analyses. I will first filter out to the taxa annotated, then to the taxa annotated as bacteria or archaea. 

I am going to do this over two steps. 
1. I will filter the taxa (`prune_taxa`) using only the taxa that were annotated functionally or taxonomically
2. Further filter the taxa by those that were annotated to bacteria or archaea (remove any "root" annotations or unclassified annotations)

## manipulate data
```{r}
ps.count <- readRDS("../data/FLK_coassembly_phyloseq_count.rds")
```

Manipulate that function file to only show prok. taxonomic IDs. 
```{r}
prok_only_fxn <- fxn %>%
  separate(max_annot_lvl, 
           into = c("id", "max_level"), 
           sep = "[|]", remove = FALSE) %>%
  filter(!(max_level %in% c("Eukaryota", "Metazoa", "Fungi", "-"))) %>%
  separate(eggNOG_OGs, 
           into = c("eggnog_ID", "root", "Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[|]", remove = FALSE) %>% # Soe taxa have two roots, so need to get id of the first Domain column, then columns in Phylum that have Eukaryotes
  separate(Domain, into = c("actual_domain"), sep = ",") %>%
  filter(actual_domain != "Eukaryota") %>%
  separate(Phylum, into = c("actual_phylum"), sep = ",") %>%
  filter(actual_phylum != "Eukaryota") %>% 
  separate(Class, into = c("actual_class"), sep = ",") %>%
  filter(actual_class != "Eukaryota") %>%
  separate(Order, into = c("actual_order"), sep = ",") %>%
  filter(actual_order != "Eukaryota") %>%
  separate(Family, into = c("actual_family"), sep = ",")

# Are there any eukaryotes left? NO!
prok_only_fxn$actual_family %>% unique %>% sort

# what about in the "root" classification? NO!
prok_only_fxn %>% 
  filter(actual_family == "root") %>%
  dplyr::select(eggNOG_OGs) %>%
  View

#Make a list of only the contig/gene ids since I will use that to filter the phyloseq object
fxn_prok_genes <- prok_only_fxn$query
```

Manipulate the taxa file to only show prok. taxonomic IDs. Keep a list of the contig/gene ids. 
```{r}
prok_taxa <- taxa %>%
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE) %>% 
  dplyr::filter(annotation != "unclassified") %>%
  dplyr::filter(Domain != "-_root")

# Check what types of annotations I have (yay, no root or eukaryotes!)
prok_taxa$Domain %>% unique %>% sort

# save a list of the gene/contig ids
taxa_prok_genes <- prok_taxa$contig
```


Now I need to put the lists together and prune the LARGE 2.3 million gene count table to only the genes that had some sort of bacterial annotation (either functional or taxonomic)

```{r}
taxa_fxn_prok_genes <- c(taxa_prok_genes, fxn_prok_genes) #172,214 genes. There may be repeats

ps.count.prok <- prune_taxa(taxa_fxn_prok_genes, ps.count)
ps.count.prok
```

Looks like this filtered the dataset down to 130,806 taxa that was annotated as bacteria/archaea or was functionally annotated as from bacteria/archaea. 

# Transform prok-only data

Should I clr-transform the data before or after the filtering. Start with trying it after the taxa filtering. 
```{r}
# transform the data using centered log-ratio transformations
ps.count.prok_clr <- microbiome::transform(ps.count.prok, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.count.prok_clr_euc <- ordinate(ps.count.prok_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
prok.clr_meta <- as(sample_data(ps.count.prok_clr), "data.frame")
prok.clr_otu <- otu_table(ps.count.prok_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
prok.euc_diss <- vegdist(prok.clr_otu, method = "euclidean")
```

## Fate - prok
### PCA fate
```{r}
sample_data(ps.count.prok_clr)$Fate_after_SP1 <- 
              factor(sample_data(ps.count.prok_clr)$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# plot PCA - facet by fate
pca_fate <- plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  facet_wrap(. ~ Fate_after_SP1) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate")
pca_fate

# plot PCA - facet by fate
plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  facet_wrap(. ~ Fate_after_SP1) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate") +
  ggrepel::geom_text_repel(label = sample_data(ps.count.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6)

# Conduct statistical tests on Fate 
# Use the dissimilarity matrix from where you read in the ps object
set.seed(10)
adonis2(formula = prok.euc_diss ~ Fate_after_SP1, data = prok.clr_meta, permutations = 999) 
```
### pariwise permanova
I would like to conduct a pairwise permanova, which refers to doing pairwise tests between the various groupings, rather than getting just an overall look at the data. 
NOTE: Julie had to get a personal access token from GitHub to get the installation to work. I'll try without, then I'll get the token, which is like what I get for using github on the hipergator. 
Pairwise permanova worked with installation using:
```
install.packages('devtools')
library(devtools)
install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
```
I think since I had a personal access token already, maybe it worked. 
```{r}
# Conduct statistical tests on Fate with the pairwise Permanova 
# Use the dissimilarity matrix from where you read in the ps object
set.seed(10)
pairwise.Fate <- pairwise.adonis2(prok.euc_diss ~ Fate_after_SP1, data = prok.clr_meta, permutations = 999) 
print(pairwise.Fate)

set.seed(10)
pairwise.Fate.zoox <- pairwise.adonis2(prok.euc_diss ~ Fate_after_SP1/Zoox_background, data = prok.clr_meta, strata = 'Zoox_background') 
print(pairwise.Fate.zoox)

pairwise.Fate.zoox <- pairwise.adonis2(prok.euc_diss ~ Fate_after_SP1/Zoox_background, data = prok.clr_meta) 
print(pairwise.Fate.zoox)

pairwise.Fate.zoox <- pairwise.adonis2(prok.euc_diss ~ Fate_after_SP1, data = prok.clr_meta, strata = 'Zoox_background') 
print(pairwise.Fate.zoox)
```


### betadisper fate
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.fate <- betadisper(prok.euc_diss, prok.clr_meta$Fate_after_SP1)

# save the distances and combine with metadata
distances <- disp.fate$distances
disp.fate.meta <- cbind(distances, prok.clr_meta)

# Set the levels so the graphs match
disp.fate.meta$Fate_after_SP1 <- factor(disp.fate.meta$Fate_after_SP1, levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# Are the data normally distributed in each group?
ggqqplot(disp.fate.meta$distances) # not normal looking
shapiro.test(disp.fate.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Fate_after_SP1, data = disp.fate.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.fate.meta$distances, disp.fate.meta$Fate_after_SP1, p.adjust.method = "BH") 

# Plot the results
betadisp_fate <- ggplot(disp.fate.meta, aes(x = Fate_after_SP1, y = distances)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Fate_after_SP1), size = 4) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate", shape = "Reef", x = "Colony Fate")

betadisp_fate
```

### Supp. Figure - Fate
```{r}
ggarrange(pca_fate, betadisp_fate, labels = c("a.", "b."), common.legend = TRUE)
```

### PCA fate subset
```{r}
# subset the unaffected and diseased fates (no recovered or will be diseased)
ps.count.prok_clr_subset <- subset_samples(ps.count.prok_clr, Fate_after_SP1 %in% c("Unaffected", "Diseased"))

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.count.prok_clr_euc_subset <- ordinate(ps.count.prok_clr_subset, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
prok.clr_meta_subset <- as(sample_data(ps.count.prok_clr_subset), "data.frame")
prok.clr_otu_subset <- otu_table(ps.count.prok_clr_subset) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
prok.euc_diss_subset <- vegdist(prok.clr_otu_subset, method = "euclidean")

sample_data(ps.count.prok_clr_subset)$Fate_after_SP1 <- 
              factor(sample_data(ps.count.prok_clr_subset)$Fate_after_SP1, 
                                 levels = c("Unaffected", "Diseased"))

# plot PCA - facet by fate
pca_fate <- plot_ordination(ps.count.prok_clr_subset, ps.count.prok_clr_euc_subset, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  scale_color_manual(values = c("#648FFF", "#DC267F")) +
  labs(color = "Colony Fate")
pca_fate

# Conduct statistical tests on Fate 
# Use the dissimilarity matrix from where you read in the ps object
set.seed(10)
adonis2(formula = prok.euc_diss_subset ~ Fate_after_SP1, data = prok.clr_meta_subset, permutations = 999) 
```

## Symbiodiniaceae - zoox
### PCA symbiont
```{r}
sample_data(ps.count.prok_clr)$Zoox_background <- factor(sample_data(ps.count.prok_clr)$Zoox_background, levels = c("none", "C", "CD", "D"))
sample_data(ps.count.prok_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.count.prok_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  ggrepel::geom_text_repel(label = sample_data(ps.count.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

symbiont_pca$layers

symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + geom_point(size=4, alpha=0.8)
symbiont_pca_final

# Conduct statistical tests on the background symbiont
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = prok.euc_diss ~ Zoox_background, data = prok.clr_meta, permutations = 999) 

#ggsave("../figures/PCA_zoox.png", width = 6, height = 4)
```
Note that for the symbiodiniaceae results, sample 3421 is the one colony with a dominant genus different (cladocopium) than all the others (breviolum). That coral is all the way out on its own in terms of a functional microbiome and was also consistently diseased. 

### betadisper symbiont
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.symbiont <- betadisper(prok.euc_diss, prok.clr_meta$Zoox_background)

# save the distances and combine with metadata
distances <- disp.symbiont$distances
disp.symbiont.meta <- cbind(distances, prok.clr_meta)

# Set the levels so the graphs match
disp.symbiont.meta$Zoox_background <- factor(disp.symbiont.meta$Zoox_background, levels = c("none", "C", "CD", "D"))
disp.symbiont.meta$Disease_on_colony_when_coring_binary <-
  factor(disp.symbiont.meta$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# Are the data normally distributed in each group?
ggqqplot(disp.symbiont.meta$distances) # not normal looking
shapiro.test(disp.symbiont.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Zoox_background, data = disp.symbiont.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.symbiont.meta$distances, disp.symbiont.meta$Zoox_background, p.adjust.method = "BH") 

# Plot the results
betadisp_symbiont <- ggplot(disp.symbiont.meta, aes(x = Zoox_background, y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Zoox_background, shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Background Symbiodiniaceae", x = "Background Symbiodiniaceae", shape = "Disease at sampling")

betadisp_symbiont
```
### Figure - Symbiodiniaceae
```{r}
ggarrange(symbiont_pca_final, betadisp_symbiont, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

## Symbiodin. Remove 3421
Adjust the phyloseq object to not have the 3421 coral sample. Re-do the clr-transformations and such. 
```{r}
ps.count.prok.3421rm <- subset_samples(ps.count.prok, CoralID != "3421")

# transform the data using centered log-ratio transformations
ps.count.prok.3421rm_clr <- microbiome::transform(ps.count.prok.3421rm, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.count.prok.3421rm_clr_euc <- ordinate(ps.count.prok.3421rm_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
prok.clr.3421rm_meta <- as(sample_data(ps.count.prok.3421rm_clr), "data.frame")
prok.clr.3421rm_otu <- otu_table(ps.count.prok.3421rm_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
prok.3421rm_euc_diss <- vegdist(prok.clr.3421rm_otu, method = "euclidean")
```
### PCA
```{r}
sample_data(ps.count.prok.3421rm_clr)$Zoox_background <- factor(sample_data(ps.count.prok.3421rm_clr)$Zoox_background, levels = c("none", "C", "CD", "D"))
sample_data(ps.count.prok.3421rm_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.count.prok.3421rm_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.count.prok.3421rm_clr, ps.count.prok.3421rm_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  ggrepel::geom_text_repel(label = sample_data(ps.count.prok.3421rm_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.count.prok.3421rm_clr, ps.count.prok.3421rm_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

symbiont_pca$layers

symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + geom_point(size=4, alpha=0.8)
symbiont_pca_final

# Conduct statistical tests on the background symbiont
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = prok.3421rm_euc_diss ~ Zoox_background, data = prok.clr.3421rm_meta, permutations = 999) 
```
### Betadisper
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.symbiont.3421rm <- betadisper(prok.3421rm_euc_diss, prok.clr.3421rm_meta$Zoox_background)

# save the distances and combine with metadata
distances <- disp.symbiont.3421rm$distances
disp.symbiont.3421rm.meta <- cbind(distances, prok.clr.3421rm_meta)

# Set the levels so the graphs match
disp.symbiont.3421rm.meta$Zoox_background <- factor(disp.symbiont.3421rm.meta$Zoox_background, levels = c("none", "C", "CD", "D"))
disp.symbiont.3421rm.meta$Disease_on_colony_when_coring_binary <-
  factor(disp.symbiont.3421rm.meta$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# Are the data normally distributed in each group?
ggqqplot(disp.symbiont.3421rm.meta$distances) # not normal looking
shapiro.test(disp.symbiont.3421rm.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Zoox_background, data = disp.symbiont.3421rm.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.symbiont.3421rm.meta$distances, disp.symbiont.3421rm.meta$Zoox_background, p.adjust.method = "BH") 

# Plot the results
betadisp_symbiont <- ggplot(disp.symbiont.3421rm.meta, aes(x = Zoox_background, y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Zoox_background, shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Background Symbiodiniaceae", x = "Background Symbiodiniaceae", shape = "Disease at sampling")

betadisp_symbiont
```
### Figure - Symbiodiniaceae without 3421
```{r}
ggarrange(symbiont_pca_final, betadisp_symbiont, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

```{r}
sample_data(ps.count.prok_clr)$Zoox_background <- factor(sample_data(ps.count.prok_clr)$Zoox_background, levels = c("none", "C", "CD", "D"))
sample_data(ps.count.prok_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.count.prok_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  ggrepel::geom_text_repel(label = sample_data(ps.count.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

symbiont_pca$layers

symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + geom_point(size=4, alpha=0.8)
symbiont_pca_final

# Conduct statistical tests on the background symbiont
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = prok.euc_diss ~ Zoox_background, data = prok.clr_meta, permutations = 999) 

#ggsave("../figures/PCA_zoox.png", width = 6, height = 4)
```



## Disease - prok
### PCA disease at sampling
```{r}
sample_data(ps.count.prok_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.count.prok_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA
disease_pca <- plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary", 
                shape ="Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))
disease_pca

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = prok.euc_diss ~ Disease_on_colony_when_coring_binary, data = prok.clr_meta, permutations = 999) 

# Do the adonis2 using strata to control for background symbiodiniaceae 
# Code for this found at https://github.com/vegandevs/vegan/discussions/600 
set.seed(10)
stratatest <- with(prok.clr_meta, how(nperm = 999, blocks = Zoox_background))
adonis2(prok.euc_diss ~ Disease_on_colony_when_coring_binary, data = prok.clr_meta, permutations = stratatest) 

# ggsave("../figures/PCA_disease_binary.png", width = 6, height = 4)
```

### betadisper disease at sampling
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.disease <- betadisper(prok.euc_diss, prok.clr_meta$Disease_on_colony_when_coring_binary)

# save the distances and combine with metadata
distances <- disp.disease$distances
disp.disease.meta <- cbind(distances, prok.clr_meta)

# Are the data normally distributed in each group?
ggqqplot(disp.disease.meta$distances) # not normal looking
shapiro.test(disp.disease.meta$distances) # data not normal

# Using the Wilcoxon rank sum test to examine if there is a significant difference between unaffected and diseased coral microbiomes. Need wilcox test becouse I have two groups. 
wilcox.test(distances ~ Disease_on_colony_when_coring_binary, 
            data = disp.disease.meta)

disp.disease.meta.no1220 <- filter(disp.disease.meta, CoralID != "1220")
wilcox.test.no1220 <- wilcox.test(distances ~ Disease_on_colony_when_coring_binary, 
            data = disp.disease.meta.no1220)

# Set levels
disp.disease.meta$Disease_on_colony_when_coring_binary <- factor(disp.disease.meta$Disease_on_colony_when_coring_binary, 
                                                                 levels = c("Unaffected", "Diseased"))

# Plot the results
betadisp_disease <- ggplot(disp.disease.meta, aes(x = Disease_on_colony_when_coring_binary, 
                                                  y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary, 
                  shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Disease at sampling", x = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  ggrepel::geom_text_repel(label = sample_data(ps.count.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6)


betadisp_disease
```

### Figure - Disease at sampling
```{r}
ggarrange(disease_pca, betadisp_disease, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

## Disease - no 1220
Adjust the phyloseq object to not have the 1220 coral sample. Re-do the clr-transformations and such. 
```{r}
ps.count.prok.1220rm <- subset_samples(ps.count.prok, CoralID != "1220")

# transform the data using centered log-ratio transformations
ps.count.prok.1220rm_clr <- microbiome::transform(ps.count.prok.1220rm, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.count.prok.1220rm_clr_euc <- ordinate(ps.count.prok.1220rm_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
prok.clr.1220rm_meta <- as(sample_data(ps.count.prok.1220rm_clr), "data.frame")
prok.clr.1220rm_otu <- otu_table(ps.count.prok.1220rm_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
prok.1220rm_euc_diss <- vegdist(prok.clr.1220rm_otu, method = "euclidean")
```

### PCA disease at sampling
```{r}
sample_data(ps.count.prok.1220rm_clr)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.count.prok.1220rm_clr)$Disease_on_colony_when_coring_binary, levels = c("Unaffected", "Diseased"))

# plot PCA
disease_pca <- plot_ordination(ps.count.prok.1220rm_clr, ps.count.prok.1220rm_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary", 
                shape ="Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A"))
disease_pca

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = prok.1220rm_euc_diss ~ Disease_on_colony_when_coring_binary, data = prok.clr.1220rm_meta, permutations = 999) 

# ggsave("../figures/PCA_disease_binary.png", width = 6, height = 4)
```

### betadisper disease at sampling
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.disease.1220rm <- betadisper(prok.1220rm_euc_diss, prok.clr.1220rm_meta$Disease_on_colony_when_coring_binary)
plot(disp.disease.1220rm)

# save the distances and combine with metadata
distances1220rm <- disp.disease.1220rm$distances
disp.disease.1220rm.meta <- cbind(distances1220rm, prok.clr.1220rm_meta)

# Are the data normally distributed in each group?
ggqqplot(disp.disease.1220rm.meta$distances) # not normal looking
shapiro.test(disp.disease.1220rm.meta$distances) # data not normal

# Using the Wilcoxon rank sum test to examine if there is a significant difference between unaffected and diseased coral microbiomes. Need wilcox test becouse I have two groups. 
wilcox.test(distances1220rm ~ Disease_on_colony_when_coring_binary, 
            data = disp.disease.1220rm.meta) 

# Set levels
disp.disease.1220rm.meta$Disease_on_colony_when_coring_binary <- factor(disp.disease.1220rm.meta$Disease_on_colony_when_coring_binary, 
                                                                 levels = c("Unaffected", "Diseased"))

# Plot the results
betadisp_disease <- ggplot(disp.disease.1220rm.meta, aes(x = Disease_on_colony_when_coring_binary, 
                                                  y = distances1220rm)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary, 
                  shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Disease at sampling", x = "Disease at sampling", shape = "Disease at sampling") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) 
 # ggrepel::geom_text_repel(label = sample_data(ps.count.prok.1220rm_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6)


betadisp_disease
```

### Figure - Disease at sampling
```{r}
ggarrange(disease_pca, betadisp_disease, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

## Clone
### PCA clone
```{r}
# plot PCA
clone_pca <- plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Clone", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  facet_wrap(. ~ Clone) +
  ggrepel::geom_text_repel(label = sample_data(ps.count.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(shape = "Disease at sampling")


clone_pca$layers

clone_pca$layers <- clone_pca$layers[-1]

clone_pca_final <- clone_pca + geom_point(size=4, alpha=0.8)
clone_pca_final

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = prok.euc_diss ~ Clone, data = prok.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_clone.png", width = 6, height = 4)
```
## Location
### PCA location
```{r}
# plot PCA
location_pca <- plot_ordination(ps.count.prok_clr, ps.count.prok_clr_euc, type="samples", 
                color="Location", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Reef", shape = "Disease at sampling")

location_pca$layers

location_pca$layers <- location_pca$layers[-1]

location_pca_final <- location_pca + geom_point(size=4, alpha=0.8)
location_pca_final

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = prok.euc_diss ~ Location, data = prok.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_reef.png", width = 6, height = 4)
```