---
title: "Betadiversity_code_to_reproduce_Figure_2"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(readxl)
library(stringr)
library(dplyr)
library(tidyr)

# For stats and visualization
library(phyloseq)
library(vegan)
library(microbiome)
library(ggplot2)
theme_set(theme_bw())
library(ggpubr)
library(devtools)
library(pairwiseAdonis)
```


# MS ANALYSIS - Prok ONLY

I recently re-ran Salmon to get gene abundances on all the prokaryote genes. The prokaryote genes were identified using EukRep. EukRep split the coassembly into prok and euk fractions. I then got predicted genes from prodigal (output both AA and nucleotide seqs). I took the prok fraction of predicted genes (nucleotide) and re-mapped the eukaryote-removed reads on the prok genes to get gene abundances. That is what I am now re-running all the PCA analyses on. 

## Data 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata - Use updated metadata (v10 with new symbiont and treatment categories)
metadata <- as.data.frame(read_excel("../data/RRC_sample_CoreSiteData_v10_RRCCoreData_SP1_Keys_OFAV.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)
nrow(prokreads)
# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names because R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)
```

## Phyloseq object
Start with a phyloseq object jsut with the count data (no annotation information). The annotations aren't needed for the PCAs I want to make. 
```{r}
## Make phyloseq object of count-related data 
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced Florida Keys samples, and no Orbicella franksi coral data. 

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)

ps.prok <- phyloseq(NUMREADS, META)
ps.prok
```

## Transform data and dissimilarity
I have a matrix of read counts in different coral samples for each gene. To accound for the compositional nature of this dataset (limited sequencing output), I am going to use centered log-ratio transformation on the read counts for the analysis of these data. I will visualize everything in Aitchison distance since it is more compositionally aware and robust to subsetting, which is what I effectively did sinc e I am not including the ECA data. 
```{r}
# transform the data using centered log-ratio transformations
ps.prok_clr <- microbiome::transform(ps.prok, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.prok_clr_euc <- ordinate(ps.prok_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
ps.prok.clr_meta <- as(sample_data(ps.prok_clr), "data.frame")
ps.prok.clr_otu <- otu_table(ps.prok_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix in vegan for PERMANOVA testing
ps.prok.euc_diss <- vegdist(ps.prok.clr_otu, method = "euclidean")
```

## PERMANOVA Stats
I want to investigate which variables are explaining differences in the microbiome. These are the variables of interest:
- colony fate
- disease at sampling
- symbiont composition (NEW AS OF 1/4/2024, UPDATED as of Mid-Jan. Carly Dennison gave me updated details. New column of interest with correct categories is "Zoox_composition_New")
- clone
- reef
```{r}
# Conduct PERMANOVA stats tests
# Use the dissimilarity matrix from where you read in the ps object

# colony disease fate
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Fate_after_SP1, data = ps.prok.clr_meta, permutations = 999) 

# zoox (symbiont) composition
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Zoox_composition_New, data = ps.prok.clr_meta, permutations = 999) 

# disease at sampling
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Condition_At_Sampling, data = ps.prok.clr_meta, permutations = 999) 

# clone
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Clone, data = ps.prok.clr_meta, permutations = 999) 

# reef location
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Location, data = ps.prok.clr_meta, permutations = 999) 

# Do the adonis2 using strata to control for symbiodiniaceae composition since I am interested in controlling for this effect. I am most interested in disease, but with symbiont composition controlling a huge portion of the community, it is difficult to disentangle
# Code for this found at https://github.com/vegandevs/vegan/discussions/600 
set.seed(10)
stratatest <- with(ps.prok.clr_meta, how(nperm = 999, blocks = Zoox_composition_New))
adonis2(ps.prok.euc_diss ~ Condition_At_Sampling, data = ps.prok.clr_meta, permutations = stratatest) 

# fate while controlling for symbiont composition
set.seed(10)
stratatest <- with(ps.prok.clr_meta, how(nperm = 999, blocks = Zoox_composition_New))
adonis2(ps.prok.euc_diss ~ Fate_after_SP1, data = ps.prok.clr_meta, permutations = stratatest) 

# Also do a pairwise permanova while controlling for symbiont composition since I am also interested in between-group permanova comparisons. 
set.seed(10)
pairwise.prok.Fate.zoox <- pairwise.adonis2(ps.prok.euc_diss ~ Fate_after_SP1/Zoox_composition_New, data = ps.prok.clr_meta, strata = 'Zoox_composition_New') 
print(pairwise.prok.Fate.zoox)
```

## Disease - prok
### PCA disease at sampling
```{r}
sample_data(ps.prok_clr)$Condition_At_Sampling <- 
              factor(sample_data(ps.prok_clr)$Condition_At_Sampling , 
                                 levels = c("Apparently Healthy", "Actively Diseased"))

# plot PCA
disease_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                color="Condition_At_Sampling", 
                shape ="Zoox_composition_New") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Condition at sampling", shape = "Symbiodiniaceae composition") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  scale_shape_manual(values = c(16, 15, 17, 18, 3, 8))

disease_pca

disease_pca_2 <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                shape = "Zoox_composition_New") +
  coord_fixed()

disease_pca_2$layers

disease_pca_2$layers <- disease_pca_2$layers[-1]

disease_pca_final <- disease_pca_2 + 
  geom_point(size=4, alpha = 0.9, aes(color = Condition_At_Sampling)) +
  scale_shape_manual(values = c(16, 15, 17, 18, 3, 8)) +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  labs(color = "Condition at sampling", shape = "Symbiodiniaceae composition")
disease_pca_final
```

### betadisper disease at sampling
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.disease <- betadisper(ps.prok.euc_diss, ps.prok.clr_meta$Condition_At_Sampling)

# save the distances and combine with metadata
distances <- disp.disease$distances
disp.disease.meta <- cbind(distances, ps.prok.clr_meta)

# Are the data normally distributed in each group?
ggqqplot(disp.disease.meta$distances) # not normal looking
shapiro.test(disp.disease.meta$distances) # data not normal

# Using the Wilcoxon rank sum test to examine if there is a significant difference between unaffected and diseased coral microbiomes. Need wilcox test becouse I have two groups. 
wilcox.test(distances ~ Condition_At_Sampling, 
            data = disp.disease.meta)

disp.disease.meta.no1220 <- filter(disp.disease.meta, CoralID != "1220")
wilcox.test.no1220 <- wilcox.test(distances ~ Condition_At_Sampling, 
            data = disp.disease.meta.no1220)
wilcox.test.no1220

# Set levels
disp.disease.meta$Condition_At_Sampling <- 
  factor(disp.disease.meta$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

# Plot the results
betadisp_disease <- ggplot(disp.disease.meta, aes(x = Condition_At_Sampling, 
                                                  y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Condition_At_Sampling, 
                  shape = Zoox_composition_New), 
              size = 5, alpha = 0.9) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  labs(color = "Condition at sampling", x = "Condition at sampling", shape = "Symbiodiniaceae composition") +
  scale_color_manual(values = c("#0C7BDC", "#FFC20A")) +
  scale_shape_manual(values = c(16, 15, 17, 18, 3, 8))

betadisp_disease
```

### Figure - Disease at sampling
```{r}
ggarrange(disease_pca_final, betadisp_disease, labels = c("a.", "b."), common.legend = TRUE, legend = "right", widths = c(2,1))
#ggsave("../figures/PCA_disease_at_sampling_ZooxNew.pdf", width = 10, height = 5)

### POTENTIAL NEW FIGURE FOR MAIN MS
disp.disease.meta$Fate_after_SP1 <- 
              factor(disp.disease.meta$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

betadisp_disease_new <- ggplot(disp.disease.meta, aes(x = Condition_At_Sampling, 
                                                  y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(fill = Fate_after_SP1, 
                  shape = Condition_At_Sampling), 
              size = 5, alpha = 0.9) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  labs(color = "Condition at sampling", x = "Condition at sampling", shape = "Condition at sampling") +
  scale_fill_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  scale_shape_manual(values = c(21, 24))

betadisp_disease_new
```

## Fate - prok
### PCA fate
```{r}
sample_data(ps.prok_clr)$Fate_after_SP1 <- 
              factor(sample_data(ps.prok_clr)$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))
sample_data(ps.prok_clr)$Condition_At_Sampling <- 
              factor(sample_data(ps.prok_clr)$Condition_At_Sampling, 
                                 levels = c("Apparently Healthy", "Actively Diseased"))

fate_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                shape = "Condition_At_Sampling") +
  coord_fixed() 
 # scale_color_manual(values = c("black", "black", "black", "black"))
 # scale_color_manual(values = c("gray40", "black",  "gray70", "gray90"))

fate_pca$layers

fate_pca$layers <- fate_pca$layers[-1]

fate_pca_final <- fate_pca + 
  geom_point(size=5, aes(fill = Fate_after_SP1)) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c("gray40", "black",  "gray70", "gray90")) +
  labs(fill = "Colony Fate", shape = "Condition at sampling")
fate_pca_final
```
### betadisper fate
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.fate <- betadisper(ps.prok.euc_diss, ps.prok.clr_meta$Fate_after_SP1)

# save the distances and combine with metadata
distances <- disp.fate$distances
disp.fate.meta <- cbind(distances, ps.prok.clr_meta)

# Set the levels so the graphs match
disp.fate.meta$Fate_after_SP1 <- factor(disp.fate.meta$Fate_after_SP1, levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))
disp.fate.meta$Condition_At_Sampling <- factor(disp.fate.meta$Condition_At_Sampling, 
                                                              levels = c("Apparently Healthy", "Actively Diseased"))

# Are the data normally distributed in each group?
ggqqplot(disp.fate.meta$distances) # not normal looking
shapiro.test(disp.fate.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Fate_after_SP1, data = disp.fate.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.fate.meta$distances, disp.fate.meta$Fate_after_SP1, p.adjust.method = "BH") 

# Plot the results
betadisp_fate <- ggplot(disp.fate.meta, aes(x = Fate_after_SP1, y = distances)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(fill = Fate_after_SP1, shape = Condition_At_Sampling), size = 5) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c("gray40", "black",  "gray70", "gray90")) +
  labs(fill = "Colony Fate", shape = "Condition at sampling", x = "Colony Fate")

betadisp_fate
```

### Supp. Figure - Fate
```{r}
ggarrange(fate_pca_final, betadisp_fate, labels = c("a.", "b."), common.legend = TRUE, legend = "right")
#ggsave("../figures/PCA_fate.pdf", width = 11, height = 5)

### ALTERNATE FOR MAIN FIGURE
betadisp_fate_new <- ggplot(disp.fate.meta, aes(x = Fate_after_SP1, y = distances)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(fill = Fate_after_SP1, shape = Condition_At_Sampling), size = 5) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  labs(fill = "Colony Fate", shape = "Condition at sampling", x = "Colony Fate")

betadisp_fate_new

fate_pca_new <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                shape = "Condition_At_Sampling") +
  coord_fixed() 
 # scale_color_manual(values = c("black", "black", "black", "black"))
 # scale_color_manual(values = c("gray40", "black",  "gray70", "gray90"))

fate_pca_new$layers

fate_pca_new$layers <- fate_pca_new$layers[-1]

fate_pca_final_new <- fate_pca_new + 
  geom_point(size=5, aes(fill = Fate_after_SP1)) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  labs(fill = "Colony Fate", shape = "Condition at sampling")
fate_pca_final_new

ggarrange(fate_pca_final_new, betadisp_fate_new, betadisp_disease_new, labels = c("a.", "b.", "c."), common.legend = TRUE, legend = "right", widths = c(2,1.5,1), nrow = 1)
#ggsave("../figures/PCA_fate_condition_combine.pdf", width = 13, height = 5)
```

## Symbiodiniaceae - zoox

Note: Make sure to use the brand new zooxanthellae groupings. These are from Carly Dennison as of mid-January. 
It is the `Zoox_composition_New` grouping.

### PCA symbiont
```{r}
sample_data(ps.prok_clr)$Condition_At_Sampling <- factor(sample_data(ps.prok_clr)$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                color="Zoox_composition_New", shape = "Condition_At_Sampling") +
  geom_point(size = 4) +
  coord_fixed() +
  scale_color_brewer(palette = "Dark2") +
  ggrepel::geom_text_repel(label = sample_data(ps.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Symbiodiniaceae composition", shape = "Condition at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                shape = "Condition_At_Sampling") +
  coord_fixed() 

symbiont_pca$layers
symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + 
  geom_point(size=5, alpha = 0.9, aes(fill = Zoox_composition_New)) +
  scale_fill_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(21, 24)) +
  labs(fill = "Symbiodiniaceae composition", shape = "Condition at sampling")
symbiont_pca_final
```
Note that for the symbiodiniaceae results, sample 3421 is the one colony with a dominant genus different (cladocopium) than all the others (breviolum). That coral is all the way out on its own in terms of a functional microbiome and was also consistently diseased. 

### betadisper symbiont
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.symbiont <- betadisper(ps.prok.euc_diss, ps.prok.clr_meta$Zoox_composition_New)

# save the distances and combine with metadata
distances <- disp.symbiont$distances
disp.symbiont.meta <- cbind(distances, ps.prok.clr_meta)

# Set the levels so the graphs match
disp.symbiont.meta$Condition_At_Sampling <-
  factor(disp.symbiont.meta$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

# Remove sample 3421 (BD symbionts) and 689 (CD symbionts) because there is only one sample in each grouping
disp.symbiont.meta.no3421or689 <- filter(disp.symbiont.meta, CoralID != "3421")
disp.symbiont.meta.no3421or689 <- filter(disp.symbiont.meta.no3421or689, CoralID != "689")

# Are the data normally distributed in each group?
ggqqplot(disp.symbiont.meta.no3421or689$distances) # normal looking
shapiro.test(disp.symbiont.meta.no3421or689$distances) # data normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# First I need to remove sample 3421 since there is only one coral in that group. It would likely skew the results. 

# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Zoox_composition_New, data = disp.symbiont.meta.no3421or689)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.symbiont.meta.no3421or689$distances, disp.symbiont.meta.no3421or689$Zoox_composition_New, p.adjust.method = "BH") 

# NEED TO DO AN ANOVA BECAUSE DATA ARE NORMAL
zoox_aov <- aov(distances ~ Zoox_composition_New, data = disp.symbiont.meta.no3421or689)

summary(zoox_aov)

TukeyHSD(zoox_aov)

# Plot the results
betadisp_symbiont <- ggplot(disp.symbiont.meta.no3421or689, aes(x = Zoox_composition_New, y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(fill = Zoox_composition_New, shape = Condition_At_Sampling), 
              size = 5, alpha = 0.9) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_fill_brewer(palette = "Dark2") +
    scale_shape_manual(values = c(21, 24)) +
  labs(fill = "Symbiodiniaceae composition", x = "Symbiodiniaceae composition", shape = "Condition at sampling")

betadisp_symbiont
```
### Figure - Symbiodiniaceae
```{r}
ggarrange(symbiont_pca_final, betadisp_symbiont, labels = c("d.", "e."), common.legend = TRUE, legend = "right")
#ggsave("../figures/PCA_zoox_composition_ZooxNew_Condition.pdf", width = 12, height = 5)
```
### MS FIGURE - Fate + Disease at sampling combined


## Clone & Location

### Clone
```{r}
# plot PCA
clone_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                color="Clone", shape = "Condition_At_Sampling") +
  coord_fixed() +
  geom_point(size = 4) +
  facet_wrap(. ~ Clone) +
  ggrepel::geom_text_repel(label = sample_data(ps.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(shape = "Condition at sampling") +
  theme(legend.position = "right")
```

### location
```{r}
# plot PCA
location_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                color="Location", shape = "Condition_At_Sampling") +
  coord_fixed() +
  labs(color = "Reef", shape = "Condition at sampling") +
  geom_point(size = 4) +
  theme(legend.position = "bottom")
```

### Figures together - SUPPLEMENT
```{r}
ggarrange(clone_pca, location_pca, labels = c("a.", "b."), widths = c(2,1))
ggsave("../figures/PCA_reef_clone_Jul25.pdf", width = 13, height = 5)
```
