---
title: "Betadiversity_code_to_reproduce_Figure_2"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(readxl)
library(stringr)
library(dplyr)

# For stats and visualization
library(phyloseq)
library(vegan)
library(microbiome)
library(ggplot2)
theme_set(theme_bw())
library(ggpubr)
```

# Load Data

## Read in tables, remove zeros
```{r}
# Load table with count abundances from salmon ("numreads")
numreads <- read.table("../data/FLK_coassembly_salmon_quant_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
numreads <- numreads[, -1] # get rid of the first column cause it is just numbered lines

# Load the table with taoxnomic information from mmseqs2
taxa <- read.table("../data/FLK_coassembly_tax_mmseqs2_lca.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

# Load the table with functional information from eggnog mapper
fxn <- read_excel("../data/FLK_coassembly_fxn_eggnog.xlsx", skip = 2, na = "-")

# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many contigs are zero across all samples?
table(rowSums(numreads) == 0)

# remove them
numreads_nozero <- numreads[rowSums(numreads) != 0, ]
dim(numreads_nozero)

# any samples with no reads?
table(colSums(numreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(numreads_nozero) <- str_replace_all(colnames(numreads_nozero), pattern = "[X]", "")
colnames(numreads_nozero) <- str_split_i(colnames(numreads_nozero), "[_]", 1)
```

## Make phyloseq object of count-related data 
```{r}
idx <- match(colnames(numreads_nozero), rownames(metadata))
metadata_3 <- metadata[idx,] # subset the metadata to only the sequenced samples

NUMREADS = otu_table(numreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_3)

ps.count <- phyloseq(NUMREADS, META)
ps.count
```

## Filter ps by abundance
```{r}
# Just to get fewer contigs/genes/taxa, sum > 10 number of reads total to remove the more sparse contigs
ps.count.filt <- filter_taxa(ps.count, function(x) sum(x) > 10, TRUE) # 1,743,873 contigs left
ps.count.filt
```

## Save PS object for quicker code 
```{r}
saveRDS(ps.count.filt, "../data/FLK_coassembly_phyloseq_count.filt.rds")
```

# Load PS object 
NOTE: Only do this to read in the RDS phyloseq object that you saved if it already exists. 
```{r}
ps.count.filt <- readRDS("../data/FLK_coassembly_phyloseq_count.filt.rds")

# transform the data using centered log-ratio transformations
ps.count.filt_clr <- microbiome::transform(ps.count.filt, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.count.filt_clr_euc <- ordinate(ps.count.filt_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
count.clr_meta <- as(sample_data(ps.count.filt_clr), "data.frame")
count.clr_otu <- otu_table(ps.count.filt_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix 
count.euc_diss <- vegdist(count.clr_otu, method = "euclidean")
```

# Beta Diversity Analysis

## Fate
### PCA fate
```{r}
sample_data(ps.count.filt_clr)$Fate_after_SP1 <- 
              factor(sample_data(ps.count.filt_clr)$Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# plot PCA - facet by fate
pca_fate <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Fate_after_SP1") +
  geom_point(size = 4) +
  coord_fixed() +
  facet_wrap(. ~ Fate_after_SP1) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate")
pca_fate

# Conduct statistical tests on Fate 
# Use the dissimilarity matrix from where you read in the ps object
set.seed(10)
adonis2(formula = count.euc_diss ~ Fate_after_SP1, data = count.clr_meta, permutations = 999) 
```

### betadisper fate
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.fate <- betadisper(count.euc_diss, count.clr_meta$Fate_after_SP1)

# save the distances and combine with metadata
distances <- disp.fate$distances
disp.fate.meta <- cbind(distances, count.clr_meta)

# Set the levels so the graphs match
disp.fate.meta$Fate_after_SP1 <- factor(disp.fate.meta$Fate_after_SP1, levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))

# Are the data normally distributed in each group?
ggqqplot(disp.fate.meta$distances) # not normal looking
shapiro.test(disp.fate.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Fate_after_SP1, data = disp.fate.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.fate.meta$distances, disp.fate.meta$Fate_after_SP1, p.adjust.method = "BH") 

# Plot the results
betadisp_fate <- ggplot(disp.fate.meta, aes(x = Fate_after_SP1, y = distances)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Fate_after_SP1), size = 4) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Colony Fate", shape = "Reef", x = "Colony Fate")

betadisp_fate
```

### Supp. Figure - Fate
```{r}
ggarrange(pca_fate, betadisp_fate, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

## Symbiodiniaceae background
### PCA symbiont
```{r}
sample_data(ps.count.filt_clr)$Zoox_background <- factor(sample_data(ps.count.filt_clr)$Zoox_background, levels = c("none", "C", "CD", "D"))

# plot PCA with labels to see which corals are farther apart
plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  ggrepel::geom_text_repel(label = sample_data(ps.count.filt_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Zoox_background", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Background Symbiodiniaceae", shape = "Disease at sampling")

symbiont_pca$layers

symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + geom_point(size=4, alpha=0.8)


# Conduct statistical tests on the background symbiont
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Zoox_background, data = count.clr_meta, permutations = 999) 

#ggsave("../figures/PCA_zoox.png", width = 6, height = 4)
```
Note that for the symbiodiniaceae results, sample 3421 is the one colony with a dominant genus different (cladocopium) than all the others (breviolum). That coral is all the way out on its own in terms of a functional microbiome and was also consistently diseased. 

### betadisper symbiont
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.symbiont <- betadisper(count.euc_diss, count.clr_meta$Zoox_background)

# save the distances and combine with metadata
distances <- disp.symbiont$distances
disp.symbiont.meta <- cbind(distances, count.clr_meta)

# Set the levels so the graphs match
disp.symbiont.meta$Zoox_background <- factor(disp.symbiont.meta$Zoox_background, levels = c("none", "C", "CD", "D"))

# Are the data normally distributed in each group?
ggqqplot(disp.symbiont.meta$distances) # not normal looking
shapiro.test(disp.symbiont.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Zoox_background, data = disp.symbiont.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.symbiont.meta$distances, disp.symbiont.meta$Zoox_background, p.adjust.method = "BH") 

# Plot the results
betadisp_symbiont <- ggplot(disp.symbiont.meta, aes(x = Zoox_background, y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Zoox_background, shape = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Background Symbiodiniaceae", x = "Background Symbiodiniaceae")

betadisp_symbiont
```
### Figure - Symbiodiniaceae
```{r}
ggarrange(symbiont_pca_final, betadisp_symbiont, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```

## Clone
### PCA clone
```{r}
# plot PCA
clone_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Clone", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  facet_wrap(. ~ Clone) +
  ggrepel::geom_text_repel(label = sample_data(ps.count.filt_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(shape = "Disease at sampling")


clone_pca$layers

clone_pca$layers <- clone_pca$layers[-1]

clone_pca_final <- clone_pca + geom_point(size=4, alpha=0.8)
clone_pca_final

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Clone, data = count.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_clone.png", width = 6, height = 4)
```
## Location
### PCA location
```{r}
# plot PCA
location_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Location", shape = "Disease_on_colony_when_coring_binary") +
  coord_fixed() +
  labs(color = "Reef", shape = "Disease at sampling")

location_pca$layers

location_pca$layers <- location_pca$layers[-1]

location_pca_final <- location_pca + geom_point(size=4, alpha=0.8)
location_pca_final

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Location, data = count.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_reef.png", width = 6, height = 4)
```


## Disease at sampling
### PCA disease at sampling
```{r}
# plot PCA
disease_pca <- plot_ordination(ps.count.filt_clr, ps.count.filt_clr_euc, type="samples", 
                color="Disease_on_colony_when_coring_binary") +
  geom_point(size = 4) +
  coord_fixed() +
  labs(color = "Disease at sampling")

# Conduct statistical tests on the clones 
# Use the dissimilarity matrix from an earlier chunk
set.seed(10)
adonis2(formula = count.euc_diss ~ Disease_on_colony_when_coring_binary, data = count.clr_meta, permutations = 999) 

# ggsave("../figures/PCA_disease_binary.png", width = 6, height = 4)
```

### betadisper disease at sampling
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.disease <- betadisper(count.euc_diss, count.clr_meta$Disease_on_colony_when_coring_binary)

# save the distances and combine with metadata
distances <- disp.disease$distances
disp.disease.meta <- cbind(distances, count.clr_meta)

# Are the data normally distributed in each group?
ggqqplot(disp.disease.meta$distances) # not normal looking
shapiro.test(disp.disease.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Disease_on_colony_when_coring_binary, data = disp.disease.meta)

# Using the Wilcoxon rank sum test to examine if there is a significant difference between unaffected and diseased coral microbiomes:
wilcox.test(distances ~ Disease_on_colony_when_coring_binary, 
            data = disp.disease.meta) 

# Plot the results
betadisp_disease <- ggplot(disp.disease.meta, aes(x = Disease_on_colony_when_coring_binary, 
                                                  y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(color = Disease_on_colony_when_coring_binary), 
              size = 4, alpha = 0.8) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  #scale_color_manual(values = c("#785EF0", "#648FFF", "#FFB000", "#DC267F")) +
  labs(color = "Background Symbiodiniaceae", x = "Background Symbiodiniaceae")

betadisp_disease
```

### Figure - Disease at sampling
```{r}
ggarrange(disease_pca, betadisp_disease, labels = c("a.", "b."), common.legend = TRUE)
# ggsave("../figures/PCA_fate.pdf", width = 9, height = 5)
```
