---
title: "Betadiversity_code_to_reproduce_figures"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(readxl)
library(stringr)
library(dplyr)
library(tidyr)

# For stats and visualization
library(phyloseq)
library(vegan)
library(microbiome)
library(ggplot2)
theme_set(theme_bw())
library(ggpubr)
library(devtools)
library(pairwiseAdonis)
```


# MS Analysis - prokaryotic microbiome beta diversity

The data included here are abundances on all prokaryote genes in a Orbicella faveolata coral metagenome co-assembly. The prokaryote genes were identified using EukRep. EukRep split the coassembly into prok and euk fractions. I then got predicted genes from prodigal (output both AA and nucleotide seqs). I took the prok fraction of predicted genes (nucleotide) and re-mapped the eukaryote-removed reads on the prok genes to get gene abundances. All analyses here represent genes and their abundances in the prokaryotic microbiome of O. faveolata. 

## Data 

Data tables can be downloaded from Zenodo: https://doi.org/10.5281/zenodo.11493775 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/FLK_OFAV_MG_prok_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata - Use updated metadata (v10 with new symbiont and treatment categories)
metadata <- as.data.frame(read_excel("../data/FLK_OFAV_MG_prok_coassembly_metadata_RRC_v10.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# remove genes are zero across all samples? Like 32k 
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]

# Fix the sample names because R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)
```

## Phyloseq object
Start with a phyloseq object jsut with the count data (no annotation information). The annotations aren't needed for the PCAs I make here.
```{r}
## Make phyloseq object of count-related data 
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced Florida Keys samples, and no Orbicella franksi coral data. 

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)

ps.prok <- phyloseq(NUMREADS, META)
ps.prok
```

## Transform data and dissimilarity
I have a matrix of read counts in different coral samples for each gene. To account for the compositional nature of this dataset (limited sequencing output), I am going to use centered log-ratio transformation on the read counts for the analysis of these data. I will visualize everything in Aitchison distance since it is more compositionally aware and robust to subsetting. 

```{r}
# transform the data using centered log-ratio transformations
ps.prok_clr <- microbiome::transform(ps.prok, 'clr')

# Generate aitchison distance ordination (euclidean distance on clr-transformed data)
ps.prok_clr_euc <- ordinate(ps.prok_clr, "RDA", "euclidean")

# Generate aitchison distance dissimilarity matrix with vegan for adonis2 test
ps.prok.clr_meta <- as(sample_data(ps.prok_clr), "data.frame")
ps.prok.clr_otu <- otu_table(ps.prok_clr) %>% as.matrix %>% t() #samples must be rows, so transpose

#get dissimilarity matrix in vegan for PERMANOVA testing
ps.prok.euc_diss <- vegdist(ps.prok.clr_otu, method = "euclidean")
```


# Figure 3a-c
a) Principal components analysis (PCA) of coral tissue prokaryotic functional microbiome beta diversity (Aitchison distance) colored by colony fate with point shape reflecting the condition of the coral at sampling. b) Prokaryotic functional beta diversity dispersion measured as the distance to centroid was greater in colonies with diseased fates relative to recovered or unaffected fates (Wilcoxon rank sum test, p < 0.05).  c) Prokaryotic functional beta diversity dispersion measured as the distance to centroid was greater in actively diseased colonies relative to apparently healthy colonies (Wilcoxon rank sum test, p < 0.05). 

## Figure 3b betadisper fate
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.fate <- betadisper(ps.prok.euc_diss, ps.prok.clr_meta$Fate_after_SP1)

# save the distances and combine with metadata
distances <- disp.fate$distances
disp.fate.meta <- cbind(distances, ps.prok.clr_meta)

# Set the levels so the graphs match
disp.fate.meta$Fate_after_SP1 <- factor(disp.fate.meta$Fate_after_SP1, levels = c("Unaffected", "Recovered", "Will be diseased", "Diseased"))
disp.fate.meta$Condition_At_Sampling <- factor(disp.fate.meta$Condition_At_Sampling, 
                                                              levels = c("Apparently Healthy", "Actively Diseased"))

# Are the data normally distributed in each group?
shapiro.test(disp.fate.meta$distances) # data not normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Fate_after_SP1, data = disp.fate.meta)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.fate.meta$distances, disp.fate.meta$Fate_after_SP1, p.adjust.method = "BH") 
```
## Figure 3c betadisper condition at sampling
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.disease <- betadisper(ps.prok.euc_diss, ps.prok.clr_meta$Condition_At_Sampling)

# save the distances and combine with metadata
distances <- disp.disease$distances
disp.disease.meta <- cbind(distances, ps.prok.clr_meta)

# Are the data normally distributed in each group?
shapiro.test(disp.disease.meta$distances) # data not normal

# Using the Wilcoxon rank sum test to examine if there is a significant difference between unaffected and diseased coral microbiomes. Need wilcox test becouse I have two groups. 
wilcox.test(distances ~ Condition_At_Sampling, 
            data = disp.disease.meta)

# Set levels
disp.disease.meta$Condition_At_Sampling <- 
  factor(disp.disease.meta$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

# Figure
disp.disease.meta$Fate_after_SP1 <- 
              factor(disp.disease.meta$Fate_after_SP1, 
                                 levels = c("Unaffected", "Recovered", "Will be diseased", "Diseased"))
```

## Figure 3a-c visualization
```{r}
fate_pca_new <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                shape = "Condition_At_Sampling") +
  coord_fixed() 
 # scale_color_manual(values = c("black", "black", "black", "black"))
 # scale_color_manual(values = c("gray40", "black",  "gray70", "gray90"))

fate_pca_new$layers

fate_pca_new$layers <- fate_pca_new$layers[-1]

fate_pca_final_new <- fate_pca_new + 
  geom_point(size=5, aes(fill = Fate_after_SP1)) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  labs(fill = "Colony Fate", shape = "Condition at sampling")

betadisp_fate_new <- ggplot(disp.fate.meta, aes(x = Fate_after_SP1, y = distances)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(fill = Fate_after_SP1, shape = Condition_At_Sampling), size = 5) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  labs(fill = "Colony Fate", shape = "Condition at sampling", x = "Colony Fate")


betadisp_disease_new <- ggplot(disp.disease.meta, aes(x = Condition_At_Sampling, 
                                                  y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(fill = Fate_after_SP1, 
                  shape = Condition_At_Sampling), 
              size = 5, alpha = 0.9) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  labs(color = "Condition at sampling", x = "Condition at sampling", shape = "Condition at sampling") +
  scale_fill_manual(values = c("#482173", "#2e6f8e",  "#29af7f", "#FFC20A")) +
  scale_shape_manual(values = c(21, 24))

ggarrange(fate_pca_final_new, betadisp_fate_new, betadisp_disease_new, labels = c("a.", "b.", "c."), common.legend = TRUE, legend = "right", widths = c(2,1.5,1), nrow = 1)
```

# Figure 3d-e 
d) PCA as in (a) colored by Symbiodiniaceae composition. e) Functional prokaryotic beta dispersion differed across Symbiodiniaceae composition groups, with corals containing Breviolum dominant + Cladocopium + Durusdinium exhibiting greater dispersion than Breviolum-only tissues (Tukey Honest Significant Differences post-hoc test, **p < 0.01). Note that the Cladocopium and Durusdinium and the Breviolum and Durusdinium groupings are omitted from panel (d) because there was only one coral in each group. All points are colored by Symbiodiniaceae composition groupings, and shape reflects the colony condition at the time of sampling. The groupings Breviolum and Durusdinium, and Cladocopium and Durusdinium contain both genera making up at least 10% of symbiont community. See Figure S3 in the manuscript for a visualization of the photoendosymbiont composition of all corals.

## Figure 3d - PCA zoox
```{r}
sample_data(ps.prok_clr)$Condition_At_Sampling <- factor(sample_data(ps.prok_clr)$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

# without the labels - want transparency
symbiont_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                shape = "Condition_At_Sampling") +
  coord_fixed() 

symbiont_pca$layers
symbiont_pca$layers <- symbiont_pca$layers[-1]

symbiont_pca_final <- symbiont_pca + 
  geom_point(size=5, alpha = 0.9, aes(fill = Zoox_composition_New)) +
  scale_fill_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(21, 24)) +
  labs(fill = "Symbiodiniaceae composition", shape = "Condition at sampling")
```
Note that for the symbiodiniaceae results, sample 3421 is the one colony with a dominant genus different (cladocopium) than all the others (breviolum). That coral is all the way out on its own in terms of a functional microbiome and was also consistently diseased. 

## Figure 3e - betadisper symbiont
```{r}
# Run betadisper function from vegan
# Hypothesis - Colony fate explains the variability in the functional microbiome
disp.symbiont <- betadisper(ps.prok.euc_diss, ps.prok.clr_meta$Zoox_composition_New)

# save the distances and combine with metadata
distances <- disp.symbiont$distances
disp.symbiont.meta <- cbind(distances, ps.prok.clr_meta)

# Set the levels so the graphs match
disp.symbiont.meta$Condition_At_Sampling <-
  factor(disp.symbiont.meta$Condition_At_Sampling, levels = c("Apparently Healthy", "Actively Diseased"))

# Remove sample 3421 (BD symbionts) and 689 (CD symbionts) because there is only one sample in each grouping
disp.symbiont.meta.no3421or689 <- filter(disp.symbiont.meta, CoralID != "3421")
disp.symbiont.meta.no3421or689 <- filter(disp.symbiont.meta.no3421or689, CoralID != "689")

# Are the data normally distributed in each group?
shapiro.test(disp.symbiont.meta.no3421or689$distances) # data normal

# Do a Kruskal-Wallis test (nonparametric) to see if there are significant differences in the dispersion by fate
# First I need to remove sample 3421 since there is only one coral in that group. It would likely skew the results. 

# My question: Is there a significant difference in dispersion between fate groups?
kruskal.test(distances ~ Zoox_composition_New, data = disp.symbiont.meta.no3421or689)

#Pairwise tests using the Wilcoxon rank sum test with Benjamini-Hochberg correction for multiple testing, which creates an adjusted P value:
pairwise.wilcox.test(disp.symbiont.meta.no3421or689$distances, disp.symbiont.meta.no3421or689$Zoox_composition_New, p.adjust.method = "BH") 

# NEED TO DO AN ANOVA BECAUSE DATA ARE NORMAL
zoox_aov <- aov(distances ~ Zoox_composition_New, data = disp.symbiont.meta.no3421or689)

summary(zoox_aov)

TukeyHSD(zoox_aov)

# Plot the results
betadisp_symbiont <- ggplot(disp.symbiont.meta.no3421or689, aes(x = Zoox_composition_New, y = distances)) +
  geom_boxplot(lwd = 0.5, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position=position_jitter(width=.1, height=0), 
              aes(fill = Zoox_composition_New, shape = Condition_At_Sampling), 
              size = 5, alpha = 0.9) +
  ylab("Distance to Centroid") +
  theme(legend.text.align = 0) +
  scale_fill_brewer(palette = "Dark2") +
    scale_shape_manual(values = c(21, 24)) +
  labs(fill = "Symbiodiniaceae composition", x = "Symbiodiniaceae composition", shape = "Condition at sampling")
```
## Figure 3d and 3e - Symbiodiniaceae 
```{r}
ggarrange(symbiont_pca_final, betadisp_symbiont, labels = c("d.", "e."), common.legend = TRUE, legend = "right")
```

# Figure S2 - Clone and Location

```{r}
# Clone PCA
clone_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                color="Clone", shape = "Condition_At_Sampling") +
  coord_fixed() +
  geom_point(size = 4) +
  facet_wrap(. ~ Clone) +
  ggrepel::geom_text_repel(label = sample_data(ps.prok_clr)$CoralID, nudge_x = 2, nudge_y = 5, color = "black", alpha = 0.6) +
  labs(shape = "Condition at sampling") +
  theme(legend.position = "right")

# Location PCA
location_pca <- plot_ordination(ps.prok_clr, ps.prok_clr_euc, type="samples", 
                color="Location", shape = "Condition_At_Sampling") +
  coord_fixed() +
  labs(color = "Reef", shape = "Condition at sampling") +
  geom_point(size = 4) +
  theme(legend.position = "bottom")

ggarrange(clone_pca, location_pca, labels = c("a.", "b."), widths = c(2,1))
```

# Figure S4 - Will be diseased analysis

I noticed that the Will be diseased category is not significantly different in microbial dispersion compared to the diseased category. We are interested in understanding if this has anything to do with the months until disease onset for these "Will be diseased" corals. I will check out my figures because I think I looked into this at some point. The valuable column here is "Months_after_sampling_when_observed_diseased".
I will do a regression. Essentially, I'd predict that when disease is coming soon, the dispersion will be higher (more dysbiotic).
I also think history will matter. Some of the "Will be diseased" corals had previous history of disease, but were unaffected at sampling and got diseased again in the future, while other "Will be diseased" corals had no history of disease, but still got diseaed. 

```{r}
disp.fate.meta.wbd <- filter(disp.fate.meta, Fate_after_SP1 == "Will be diseased")

# Does dispersion increase when disease is sooner to come? NO
lmtest <- lm(distances ~ Months_after_sampling_when_observed_diseased, data = disp.fate.meta.wbd)
summary(lmtest)

months <- ggplot(disp.fate.meta.wbd, aes(x = Months_after_sampling_when_observed_diseased, y = distances)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgray") +
  ggrepel::geom_text_repel(label = disp.fate.meta.wbd$CoralID, nudge_x = 2, nudge_y = 5, color = "black") +
  ylab("Distance to Centroid") +
  xlab("Months after sampling of disease onset")

# Does history of disease influence microbial dispersion for will be diseased colonies?
wilcox.test(distances ~ Disease_before_SP1, data = disp.fate.meta.wbd)

history <- ggplot(disp.fate.meta.wbd, aes(x = Disease_before_SP1, y = distances)) +
  geom_boxplot(lwd = 1, outlier.shape = NA, color = "darkgray") +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1, color = "darkgray") +
  geom_jitter(position=position_jitter(width=.1, height=0), size = 5) +
  ylab("Distance to Centroid") +
  ggrepel::geom_text_repel(label = disp.fate.meta.wbd$CoralID, color = "black") +
  xlab("Disease prior to sampling")

ggarrange(months, history)
```

# Table S4 - PERMANOVA Stats
I investigate which variables are explaining variability in the microbiome across O. faveolata corals. These are the variables of interest:    
* colony fate    
* disease at sampling     
* symbiont composition     
* clone    
* reef     
```{r}
# Conduct PERMANOVA stats tests
# Use the dissimilarity matrix from the "Transform data and dissimilarity" section

# colony disease fate
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Fate_after_SP1, data = ps.prok.clr_meta, permutations = 999) 

# zoox (symbiont) composition
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Zoox_composition_New, data = ps.prok.clr_meta, permutations = 999) 

# disease at sampling
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Condition_At_Sampling, data = ps.prok.clr_meta, permutations = 999) 

# clone
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Clone, data = ps.prok.clr_meta, permutations = 999) 

# reef location
set.seed(10)
adonis2(formula = ps.prok.euc_diss ~ Location, data = ps.prok.clr_meta, permutations = 999) 

# Do the adonis2 using strata to control for symbiodiniaceae composition since I am interested in controlling for this effect. I am most interested in disease, but with symbiont composition controlling a huge portion of the community, it is difficult to disentangle
# Code for this found at https://github.com/vegandevs/vegan/discussions/600 
set.seed(10)
stratatest <- with(ps.prok.clr_meta, how(nperm = 999, blocks = Zoox_composition_New))
adonis2(ps.prok.euc_diss ~ Condition_At_Sampling, data = ps.prok.clr_meta, permutations = stratatest) 

# fate while controlling for symbiont composition
set.seed(10)
stratatest <- with(ps.prok.clr_meta, how(nperm = 999, blocks = Zoox_composition_New))
adonis2(ps.prok.euc_diss ~ Fate_after_SP1, data = ps.prok.clr_meta, permutations = stratatest) 
```