---
title: "DA_code_to_reproduce_Figure_3"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(dplyr)
library(tidyr)
library(readxl)
library(stringr)
library(KEGGREST)

# For visualization
library(corncob)
library(DESeq2)
library(phyloseq)
library(randomcoloR)
```

# Read in data

## Load raw data 

```{r}
# Load table with count abundances from salmon ("numreads")
numreads <- read.table("../data/FLK_coassembly_salmon_quant_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
numreads <- numreads[, -1] # get rid of the first column cause it is just numbered lines

# Load the table with taoxnomic information from mmseqs2
taxa <- read.table("../data/FLK_coassembly_tax_mmseqs2_lca.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

# Load the table with functional information from eggnog mapper
fxn <- read_excel("../data/FLK_coassembly_fxn_eggnog.xlsx", skip = 2, na = "-")

# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many contigs are zero across all samples?
table(rowSums(numreads) == 0)

# remove them
numreads_nozero <- numreads[rowSums(numreads) != 0, ]
dim(numreads_nozero)

# any samples with no reads?
table(colSums(numreads_nozero) == 0)

# Fix the sample names
colnames(numreads_nozero) <- str_replace_all(colnames(numreads_nozero), pattern = "[X]", "")
colnames(numreads_nozero) <- str_split_i(colnames(numreads_nozero), "[_]", 1)
```

## Phyloseq object 
```{r}
idx <- match(colnames(numreads_nozero), rownames(metadata))
metadata_3 <- metadata[idx,]

NUMREADS = otu_table(numreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_3)

ps.count <- phyloseq(NUMREADS, META)
ps.count
```

## Annotated genes only
I would like to filter this phyloseq object to only include genes that are annotated. My rationale is that if I am to pull out any discussion or prediction about functional potential of the differentially abundant genes, then it would be useful to have that be what I do the analysis with. 
```{r}
numreads_nozero_for_fxn_filt <- numreads_nozero %>%
  mutate(query = rownames(numreads_nozero))

# Add count (number of reads) to the function data frame
fxn.numreads_nozero <- fxn %>%
  left_join(numreads_nozero_for_fxn_filt, by = "query")

fxn.numreads_nozero

# Select out only the reads again
numreads_nozero_annotated_only <- as.data.frame(fxn.numreads_nozero[,22:62])
rownames(numreads_nozero_annotated_only) <- fxn.numreads_nozero$query

# It has occurrred to me that some of these annotations are actually missing from the abundance table so are showing up as NAs. I need to remove those samples. 
table(is.na(rowSums(numreads_nozero_annotated_only)))
table(is.finite(rowSums(numreads_nozero_annotated_only)))

numreads_nozero_annotated_only_noNA <-
  numreads_nozero_annotated_only[is.finite(rowSums(numreads_nozero_annotated_only)), ]

numreads_nozero_annotated_only_noNA %>% dim

# Make the actual phyloseq object
idx <- match(colnames(numreads_nozero_annotated_only_noNA), metadata$CoralID)
metadata_annotated <- metadata[idx,]

NUMREADS = otu_table(numreads_nozero_annotated_only_noNA, taxa_are_rows = TRUE)
META = sample_data(metadata_annotated)

ps.count.annotated <- phyloseq(NUMREADS, META)
ps.count.annotated
```
This drastically reduced the number of genes I am working with, which makes sense. Ultimately, I am only doing annotated genes because I want to be able to draw biological conclusions from the data. 

## Save PS annotated object
```{r}
saveRDS(ps.count.annotated, "../data/FLK_coassembly_phyloseq_count.annotated.rds")
```

## Quick start - load ps object
```{r}
# Now what is it like to read this back in?
ps.count.annotated <- readRDS("../data/FLK_coassembly_phyloseq_count.annotated.rds")
```

# Differential abundance 
## deseq2 using the following code:
Set the levels of the factor of interest
```{r}
sample_data(ps.count.annotated)$Disease_on_colony_when_coring_binary <-
  factor(sample_data(ps.count.annotated)$Disease_on_colony_when_coring_binary, 
         levels = c("Unaffected", "Diseased"))

head(sample_data(ps.count.annotated)$Disease_on_colony_when_coring_binary, 10) # levels look good
```

These levels look good, and we are good to move on to DESeq2 analysis on the OFAV florida keys samples between completely unaffected to persistantly diseased coral colonies. 
```{r}
dds = phyloseq_to_deseq2(ps.count.annotated, ~ Disease_on_colony_when_coring_binary)
dds

#Keep all rows and apply geometric mean to replace zeros
cts <- DESeq2::counts(dds)
geoMeans <- apply(cts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
dds <- estimateSizeFactors(dds, geoMeans = geoMeans)

# run the actual deseq function - check results at 0.05 p adjusted value
dds <- DESeq(dds,test="Wald", fitType="parametric")
res = results(dds, cooksCutoff = FALSE)
alpha = 0.05
sigtab = res[which(res$padj < alpha), ] # 5,894 DA genes

# Add the functional annotations to the subset differentially abundant genes
sigtab.fxn <- as(sigtab, "data.frame") %>% 
  mutate(query = rownames(sigtab)) %>% 
  left_join(fxn, by = "query")

# Write these taxa out to a file to be saved for later
write.table(sigtab.fxn,"../data/DESeq2_coassembly_results_disease_0.05.txt", sep="\t", 
            col.names = TRUE, row.names = FALSE)

# Write the 0.01 cutoff genes
alpha01 = 0.01
sigtab01 = res[which(res$padj < alpha01), ] # 2,449 DA genes

# Add the functional annotations to the subset differentially abundant genes
sigtab01.fxn <- as(sigtab01, "data.frame") %>% 
  mutate(query = rownames(sigtab01)) %>% 
  left_join(fxn, by = "query")

# Write these taxa out to a file to be saved for later
write.table(sigtab01.fxn,"../data/DESeq2_coassembly_results_disease_0.01.txt", sep="\t", 
            col.names = TRUE, row.names = FALSE)

sigtab.fxn
sigtab01.fxn
```

## Corncob using the following code:
```{r, eval=FALSE}
set.seed(10) # set seed for reproducibility
corncobDA.dis <- differentialTest(formula = ~ Disease_on_colony_when_coring_binary, 
                                   phi.formula = ~ Disease_on_colony_when_coring_binary,
                                   formula_null = ~ 1,
                                   phi.formula_null = ~ Disease_on_colony_when_coring_binary,
                                   test = "Wald", boot = FALSE,
                                   data = ps.count.annotated,
                                   fdr_cutoff = 0.05)

# Save the corncob phyloseq object so you can manipulate it in R
saveRDS(corncobDA.dis, "corncob_output_disease_at_sampling.rds")

## Try controlling for the differences in the symbiont community
set.seed(10) # set seed for reproducibility
corncobDA.dis.symb <- differentialTest(formula = ~ Disease_on_colony_when_coring_binary + Zoox_background, 
                                  phi.formula = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  formula_null = ~ Zoox_background,
                                  phi.formula_null = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  test = "Wald", boot = FALSE,
                                  data = ps.count.annotated,
                                  fdr_cutoff = 0.05)

# Save the corncob phyloseq object so you can manipulate it in R
saveRDS(corncobDA.dis.symb, "corncob_output_disease_at_sampling_zoox.rds")
```
I would like to see how the number of differentially abudnant genes differs between these two runs, but I feel like I should do the second one, because it accounts for background zoox, which significantly explained variation in the microbiome, so I feel it is best to account for both things. 

### disease at sampling only
Then, I manipulated and saved the corncob output with this code:
```{r}
# read in RDS file
corncobDA.dis <- readRDS("../data/corncob_output_disease_at_sampling.rds")

plot(corncobDA.dis)

# How many significant taxa are there?
corncobDA.dis$significant_taxa %>% length
```


```{r}

```

### diseae at sampling while controlling for differential abundance due to background zooxanthellae
```{r}
# read in RDS file
corncobDA.dis.symb <- readRDS("../data/corncob_output_disease_at_sampling_zoox.rds")

plot(corncobDA.dis.symb)

# How many significant taxa are there?
corncobDA.dis.symb$significant_taxa %>% length
```
There are so many taxa that are differentially abundant. To get the ones that change most, it may be worthwhile to identify the taxa with the greatest change between unaffected and diseased taxa, or those with the greatest coefficient (positive or negative).
```{r}
# I want to extract the important information from this into a table...
# check out results structure from one of the models
corncobDA.dis.symb$significant_models[[1]]$coefficients

# write a function to extract the corncob results
extractmods2 <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                        "Std.Error" = model$coefficients[2, 2], 
                        "p" = model$coefficients[2, 4])
  return(result)
}

# save the output as a list using the above function
corncob.dis.symb.models <- lapply(corncobDA.dis.symb$significant_models, extractmods2)
names(corncob.dis.symb.models) <- corncobDA.dis.symb$significant_taxa

# save a table of annotations for just the significant genes
sig.fxnl.genes.dis.symb <- filter(fxn, query %in% corncobDA.dis.symb$significant_taxa)

# turn the output from a list to a dataframe for ease of saving and manipulating
corncob.dis.symb.models.df <- plyr::ldply(corncob.dis.symb.models, data.frame) %>% 
  left_join(sig.fxnl.genes.dis.symb, by = c(".id" = "query"))

### Save these data so you don't have to re-run the model ###
write.table(corncob.dis.symb.models.df, "../data/Corncob_coassembly_results_dis.symb_0.05.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

## Corncob disease and symbiont
Read in the corncob data. I have a couple of questions. I want to see what the greatest 
```{r}
corn.dis.symb <- read.table("../data/Corncob_coassembly_results_dis.symb_0.05.txt", header = TRUE, sep = "\t")

corn.dis.symb$Estimate %>% summary

table(corn.dis.symb$Estimate > 0) 
table(corn.dis.symb$Estimate < -1) 
table(corn.dis.symb$Estimate < -3) 
table(corn.dis.symb$Estimate > 3) 
table(corn.dis.symb$Estimate < -3.5) 
table(corn.dis.symb$Estimate > 5) 
```
## Figure for MS

I imagine producing a table of all the differentially abundant genes, but for a general figure, I am interested in:
- What functions are these differentially abundant genes performing?
- What taxonomic groups are these differentially abundant genes part of?

Genes enriched in diseased - what taxonomic groups are there?
Genes enriched in unaffected - what taxonomic groups are there?

We conducted differential abundance tests in apparently healthy tissue between unaffected and diseased coral colonies while controlling for the effect of background symbiodiniaceae on differential abundance, and identified 1,605 differentially abundant genes (supplementary table). This included 303 enriched in diseased tissue, and 1302 enriched in unaffected coral tissue. These genes were from varied taxonomic groups, including x, y, and z. 

### Taxonomic info
```{r}
# What taxa ID is assigned to genes that were enriched in disease
corn.dis.symb_tax <- corn.dis.symb %>%
  left_join(taxa, by = c(".id" = "contig")) %>%
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE)

# How many of the total differentially abundant genes were actually annotated as bacteria?
# disease-enriched - 168 of 303 
dis.prok <- filter(corn.dis.symb_tax, Estimate > 0) %>% 
  filter(Domain %in% c("d_Bacteria", "d_Archaea")) %>%
  mutate(Phylum = case_when(
    is.na(Phylum) ~ paste0(Domain),
    TRUE ~ Phylum
    )) %>%
  mutate(Class = case_when(
    is.na(Class) ~ paste0(Phylum),
    TRUE ~ Class
    )) %>%
  mutate(Order = case_when(
    is.na(Order) ~ paste0(Class),
    TRUE ~ Order
    )) %>%
  mutate(Family = case_when(
    is.na(Family) ~ paste0(Order),
    TRUE ~ Family
    )) %>%
  mutate(Genus = case_when(
    is.na(Genus) ~ paste0(Family),
    TRUE ~ Genus
    )) %>%
  mutate(Species = case_when(
    is.na(Species) ~ paste0(Genus),
    TRUE ~ Species
    ))


# unaffected-enriched - 1,129 of 1,302
unaff.prok <- filter(corn.dis.symb_tax, Estimate < 0) %>% 
  filter(Domain %in% c("d_Bacteria", "d_Archaea")) %>%
  mutate(Phylum = case_when(
    is.na(Phylum) ~ paste0(Domain),
    TRUE ~ Phylum
    )) %>%
  mutate(Class = case_when(
    is.na(Class) ~ paste0(Phylum),
    TRUE ~ Class
    )) %>%
  mutate(Order = case_when(
    is.na(Order) ~ paste0(Class),
    TRUE ~ Order
    )) %>%
  mutate(Family = case_when(
    is.na(Family) ~ paste0(Order),
    TRUE ~ Family
    )) %>%
  mutate(Genus = case_when(
    is.na(Genus) ~ paste0(Family),
    TRUE ~ Genus
    )) %>%
  mutate(Species = case_when(
    is.na(Species) ~ paste0(Genus),
    TRUE ~ Species
    ))

1129/1303 # 86.6%

# Can I just make a stacked bar chart with two columns (one disease and one unaffected)?
dis.unaff.prok <- rbind(dis.prok, unaff.prok) %>%
  mutate(corncob_result = ifelse(Estimate > 0, "Disease_assoc", "Unaffected_assoc"))

x1 <- distinctColorPalette(50)

# Make a plot
ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Phylum)) +
  geom_bar() +
  scale_fill_manual(values = x1)

ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Phylum)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = x1)

ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Class)) +
  geom_bar() +
  scale_fill_manual(values = x1)

ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Class)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = x1)
```

# KEGG Diff. Abundance

I want to analyze the differentially abunant genes of only annotated (by KEGG) the microbial genes that were annotated. 

I recently re-ran Salmon to get gene abundances on all the prokaryote genes. The prokaryote genes were identified using EukRep. EukRep split the coassembly into prok and euk fractions. I then got predicted genes from prodigal (output both AA and nucleotide seqs). I took the prok fraction of predicted genes (nucleotide) and re-mapped the noneuk reads on the prok genes to get gene abundances. That is what I am now re-running all the PCA analyses on. 

I also took the AA file of the predicted genes and ran that in the GhostKOALA database to get KEGG annotations. Kegg is useful because it has all that hierarchical data. 

## Data 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)

# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)
```

```{r}
# load KO and Gene ids
# Read all lines from the file
lines <- readLines("../data/user_ko.txt")

# Split each line by tab ("\t")
split_lines <- strsplit(lines, "\t")

# Convert the split lines into a data frame
gene_ko <- do.call(rbind, lapply(split_lines, function(x) {
  if (length(x) == 1) {
    c(x[1], NA)  # If only one column, fill second column with NA
  } else {
    x  # Keep both columns
  }
}))

# Assign column names
colnames(gene_ko) <- c("geneID", "KO")

dim(gene_ko)
unique(gene_ko) %>% dim
gene_ko$geneID %>% unique %>% length

# Convert to a data frame
gene_ko <- as.data.frame(gene_ko, stringsAsFactors = FALSE)

# View the resulting data
print(head(gene_ko))

# Read in the table of the annotations
ko_info <- read_excel("../data/kegg_data.xlsx", col_names = TRUE, na = "")

# Get rid of the "KO" in the first column
ko_info$ENTRY <- str_split_i(ko_info$ENTRY, " ", 1)

head(ko_info)
dim(ko_info)

# How many unique KOs? 4563
ko_info$KO_ID %>% unique %>% length
ko_info$ENTRY %>% unique %>% length

# There's too many duplicate rows. 
ko_info_nodups <- unique(ko_info)

# add the annotation info to the gene_ko documant
gene_ko_join <- gene_ko %>%
  left_join(ko_info_nodups, by = c("KO" = "ENTRY"))

# How many genes are annotated? 32,980
sum(!is.na(gene_ko_join$KO))

# Filter to only include the annotated columns (should end with 32,980 rows)
gene_ko_annotated <- gene_ko_join[!is.na(gene_ko_join$KO), ]

# It looks like there are some duplicated gene IDs
gene_ko_annotated$geneID %>% unique %>% length

# Get rid of duplicate geneID's
unique_gene_ko_annotated<- gene_ko_annotated[!duplicated(gene_ko_annotated$geneID), ]
dim(unique_gene_ko_annotated)

# Rename the rows 
rownames(unique_gene_ko_annotated) <- unique_gene_ko_annotated[,"geneID"]
```

## Phyloseq object
Start with a phyloseq object jsut with the count data (no annotation information). The annotations aren't needed for the PCAs I want to make. 
```{r}
## Make phyloseq object of count-related data 
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced samples

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
FXN = tax_table(as.matrix(unique_gene_ko_annotated))

ps.prok.annotated <- phyloseq(NUMREADS, META, FXN)
ps.prok.annotated
```

## Corncob

I will use corncob to test the hypothesis that functional genes change in abundance between corals visibly affected by SCTLD or not visibly affected by the disease. Knowing that the composition of symbiodiniaceae also impacts the microbiome composition, I am going to control for the effect of symbiont composition. 

I am going to use raw counts of the different genes from Salmon since Corncob models the relative abundances. 

```{r}
## Set levels
sample_data(ps.prok.annotated)$Disease_on_colony_when_coring_binary <-
  factor(sample_data(ps.prok.annotated)$Disease_on_colony_when_coring_binary, 
         levels = c("Unaffected", "Diseased"))

head(sample_data(ps.prok.annotated)$Disease_on_colony_when_coring_binary, 10) # levels look good

sample_data(ps.prok.annotated)$Zoox_background <- 
  factor(sample_data(ps.prok.annotated)$Zoox_background, 
         levels = c("none", "C", "CD", "D"))

head(sample_data(ps.prok.annotated)$Zoox_background, 10) # levels look good

## Test disease on colony when coring and control for symbiont composition
set.seed(10) # set seed for reproducibility
prok.da.dis <- differentialTest(formula = ~ Disease_on_colony_when_coring_binary + Zoox_background, 
                                  phi.formula = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  formula_null = ~ Zoox_background,
                                  phi.formula_null = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  test = "Wald", boot = FALSE,
                                  data = ps.prok.annotated,
                                  fdr_cutoff = 0.05)

plot(prok.da.dis, level = "SYMBOL")
```

WOW! This resulted in 685 differentially abundant taxa!!! That is kinda a LOT. I would like to be thoughtful in how I represent that data. Maybe I will do the ones that are disease-enriched in one graph, and one that is healthy-enriched in the other graph? I could also plot them all as a heatmap and I DEFINTIELY want to graph them by their BRITE hierarchies. 

Out of curiosity, I am also going to see how many taxa there are with a more stringent p-value cutoff. Save all the data first
```{r}
# write a function to extract the corncob results
extractmods2 <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                        "Std.Error" = model$coefficients[2, 2], 
                        "p" = model$coefficients[2, 4])
  return(result)
}

# save the output as a list using the above function
sig.models <- lapply(prok.da.dis$significant_models, extractmods2)
names(sig.models) <- prok.da.dis$significant_taxa

# save a table of annotations for just the significant genes
sig.fxnl.genes <- filter(unique_gene_ko_annotated, geneID %in% prok.da.dis$significant_taxa)

# turn the output from a list to a dataframe for ease of saving and manipulating
sig.models.df <- plyr::ldply(sig.models, data.frame) %>% 
  left_join(sig.fxnl.genes, by = c(".id" = "geneID"))

### Save these data so you don't have to re-run the model ###
write.table(sig.models.df, "../data/Corncob_prok_annotated_results_dis.symb_0.05.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

How many different genes are there significant at a adjusted p value < 0.001?
```{r}
filter(sig.models.df, p < 0.001) #299 genes
filter(sig.models.df, Estimate > 0)
filter(sig.models.df, Estimate < 0)

sig.models.df$KO %>% length
sig.models.df$KO %>% unique %>% length

sig.taxa <- sig.models.df$KO %>% unique

write(sig.taxa, file = "sig.taxa.KO.txt")
```

I need to try to manipulate the data table output so that I can get a table that shows the KEGG Orthology information clearly.

Try looping through a bunch of KO ids. I will loop through all the KO IDs and get them combined. 

```{r}
# List of KO identifiers
ko_ids <- readLines("sig.taxa.KO.txt")

ko_ids_testgroup <- c("K09125", "K00950", "K15738", "K07457", "K24163", "K00059")

# Create an empty dataframe to store parsed data for all KO IDs
parsed_data <- data.frame(KO_ENTRY = character(), 
                          BRITEhierarchy = character(),
                          Level2 = character(),
                          Level3 = character(),
                          Level4 = character(),
                          Level5 = character(),
                          Level6 = character(),
                          KO = character(),
                          stringsAsFactors = FALSE)

# Loop through each KO ID
for (ko_id in ko_ids) {
  
  # Retrieve information for the KO
  ko_info <- keggGet(ko_id)
  
  # Note which KO ID you are on
  print(paste("Processing KO ID: ", ko_id))
  
  # Check if the KO ID has BRITE hierarchy data
  if (!is.null(ko_info[[1]]$BRITE)) {
    brite <- ko_info[[1]]$BRITE  # Extract the BRITE section
    
    # Initialize variables for hierarchy levels
    brite_hierarchy <- level2 <- level3 <- level4 <- level5 <- level6 <- NA
    
    # Process each line in the BRITE section
    for (line in brite) {
      
      # Determine indentation level (number of leading spaces)
      indentation <- attr(regexpr("^\\s*", line), "match.length")  # Count spaces
      content <- trimws(line)  # Remove leading/trailing spaces
      
      if (indentation == 0) {
          # Top-level category (Level1)
          brite_hierarchy <- content
          level2 <- level3 <- level4 <- level5 <- level6 <- NA  # Reset lower levels
      } else if (indentation == 1) {
          # Second-level category (Level2)
          level2 <- content
          level3 <- level4 <- level5 <- level6 <- NA  # Reset lower levels
      } else if (indentation == 2) {
          # Third-level category (Level3)
          level3 <- content
          level4 <- level5 <- level6 <- NA  # Reset lower level
      } else if (indentation == 3) {
          # KO entry or Fourth-level category (Level4)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko <- content
            parsed_data <- rbind(parsed_data, data.frame(KO_ENTRY = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO = ko,
                                                         stringsAsFactors = FALSE))
          } else { 
            level4 <- content
            level5 <- level6 <- NA
          }
      } else if (indentation == 4) {
          # KO entry or Fifth-level category (Level5)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko <- content
            parsed_data <- rbind(parsed_data, data.frame(KO_ENTRY = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO = ko,
                                                         stringsAsFactors = FALSE))
          } else {
            level5 <- content
            level6 <- NA
          }
      } else if (indentation == 5) {
          # KO entry or or Sixth-level category (Level6)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko <- content
            parsed_data <- rbind(parsed_data, data.frame(KO_ENTRY = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO = ko,
                                                         stringsAsFactors = FALSE))
          } else {
            level6 <- content
          }
      } else if (indentation == 6) {
        # KO entry or or Seventh-level category (Level7)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko <- content
            parsed_data <- rbind(parsed_data, data.frame(KO_ENTRY = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO = ko,
                                                         stringsAsFactors = FALSE))
          } else {
            ko <- paste("WARNING, doesn't match KO ID", content)
            parsed_data <- rbind(parsed_data, data.frame(KO_ENTRY = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO = ko,
                                                         stringsAsFactors = FALSE))
          } 
      } else {
          # For unexpected cases, add a catch-all (optional)
          warning(paste("Unexpected indentation level encountered: ", indentation, "for the following KO: ", ko_id))
      }
    }
  } else {
    warning(paste("No BRITE data found for KO ID:", ko_id))
  }
}

# View the parsed hierarchy
print(parsed_data)

```

Now, the `parsed_data` data frame is the BRITE hierarchy data for the significant taxa. Ultimately, I just want the KEGG orthology data for each of these KO IDs. 

Then, I will make a data table of the estimates and a heatmap of the corncob results. I will subset only of the KEGG Orthology annotations. Then I will focus on specifically metabolism-based KEGG Orthology annotations since those have an ecological function more than things like "DNA repair" or similar. 
```{r}
KEGG_Orthology_sig_ko <- parsed_data %>%
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism")

KEGG_Orthology_sig_ko$KO_ENTRY %>% unique %>% length

parsed_data$KO_ENTRY %>% unique %>% length
```


