---
title: "DA_code_to_reproduce_Figure_3"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(dplyr)
library(tidyr)
library(readxl)
library(stringr)
library(KEGGREST)

# For visualization
library(corncob)
library(DESeq2)
library(phyloseq)
library(randomcoloR)
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
```


# MS ANALYSIS - KEGG Diff. Abundance - Disease vs. Healthy

I want to analyze the differentially abunant genes of only annotated (by KEGG) the microbial genes that were annotated. 

I recently re-ran Salmon to get gene abundances on all the prokaryote genes. The prokaryote genes were identified using EukRep. EukRep split the coassembly into prok and euk fractions. I then got predicted genes from prodigal (output both AA and nucleotide seqs). I took the prok fraction of predicted genes (nucleotide) and re-mapped the noneuk reads on the prok genes to get gene abundances. That is what I am now re-running all the PCA analyses on. 

I also took the AA file of the predicted genes and ran that in the GhostKOALA database to get KEGG annotations. Kegg is useful because it has all that hierarchical data. 

I am also re-running this on 1/4 because I re-made the symbiodiniaceae composition (Zoox_composition) column in the metadata since I found errors in the Background Zoox column. 

I am re-running this to include a colony fate category. Additionally, the colony fate category was recently re-organized because we have updated colony information. Additionally, I changed the disease at sampling category to "Condition At Sampling".

## Data 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata
metadata <- as.data.frame(read_excel("../data/RRC_sample_CoreSiteData_v10_RRCCoreData_SP1_Keys_OFAV.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)

# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)

# Read in the taxonomy
prok_taxa <- read.table("../data/FLK_prok_coassembly_tax_mmseqs2.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")
```
### KO data
```{r}
# load KO and Gene ids
# Read all lines from the file
lines <- readLines("../data/user_ko.txt")

# Split each line by tab ("\t")
split_lines <- strsplit(lines, "\t")

# Convert the split lines into a data frame
gene_ko <- do.call(rbind, lapply(split_lines, function(x) {
  if (length(x) == 1) {
    c(x[1], NA)  # If only one column, fill second column with NA
  } else {
    x  # Keep both columns
  }
}))

# Assign column names
colnames(gene_ko) <- c("geneID", "KO")

dim(gene_ko)
unique(gene_ko) %>% dim

# Convert to a data frame
gene_ko <- as.data.frame(gene_ko, stringsAsFactors = FALSE)

# how many unique KO ids?
gene_ko$KO %>% unique %>% length

# View the resulting data
print(head(gene_ko))

# Read in the table of the annotations
ko_info <- read_excel("../data/kegg_data.xlsx", col_names = TRUE, na = "")

# Get rid of the "KO" in the first column
ko_info$ENTRY <- str_split_i(ko_info$ENTRY, " ", 1)

head(ko_info)
dim(ko_info)

# How many unique KOs? 4563
ko_info$KO_ID %>% unique %>% length
ko_info$ENTRY %>% unique %>% length

# There's too many duplicate rows. Should have only 4,563
ko_info_nodups <- unique(ko_info)

# add the annotation info to the gene_ko documant
gene_ko_join <- gene_ko %>%
  left_join(ko_info_nodups, by = c("KO" = "ENTRY"))

# How many genes are annotated? 32,980
sum(!is.na(gene_ko_join$KO))

# Filter to only include the annotated columns (should end with 32,980 rows)
gene_ko_annotated <- gene_ko_join[!is.na(gene_ko_join$KO), ]

# It looks like there are some duplicated gene IDs
gene_ko_annotated$geneID %>% unique %>% length

# Get rid of duplicate geneID's
unique_gene_ko_annotated<- gene_ko_annotated[!duplicated(gene_ko_annotated$geneID), ]
dim(unique_gene_ko_annotated)

# Rename the rows 
rownames(unique_gene_ko_annotated) <- unique_gene_ko_annotated[,"geneID"]
```

## Phyloseq object
Start with a phyloseq object jsut with the count data (no annotation information). The annotations aren't needed for the PCAs I want to make. 
```{r}
## Make phyloseq object of count-related data 
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced samples

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
FXN = tax_table(as.matrix(unique_gene_ko_annotated))

ps.prok.annotated <- phyloseq(NUMREADS, META, FXN)
ps.prok.annotated
```

## Corncob

I will use corncob to test the hypothesis that functional genes change in abundance between corals visibly affected by SCTLD or not visibly affected by the disease. Knowing that the composition of symbiodiniaceae also impacts the microbiome composition, I am going to control for the effect of symbiont composition. 

I am going to use raw counts of the different genes from Salmon since Corncob models the relative abundances. 

For the symbiont composition, I will use the new column called `Zoox_composition`, which doesn't have errors. DO NOT USE the Zoox_background column anymore. There are errors in it. 

3/2/25 - I UPDATED THIS with the new zoox composition groupings (`Zoox_composition_New`). This way the anlaysis reflects the actual symbiont groupings.  
```{r}
## Set levels
sample_data(ps.prok.annotated)$Condition_At_Sampling <-
  factor(sample_data(ps.prok.annotated)$Condition_At_Sampling, 
         levels = c("Apparently Healthy", "Actively Diseased"))

head(sample_data(ps.prok.annotated)$Condition_At_Sampling, 10) # levels look good

sample_data(ps.prok.annotated)$Zoox_composition_New <- 
  factor(sample_data(ps.prok.annotated)$Zoox_composition_New, 
         levels = c("B", "B_dom_plus_C", "B_dom_plus_CD", "B_dom_plus_D", "BD", "CD"))

head(sample_data(ps.prok.annotated)$Zoox_composition_New, 10) # levels look good

## Test disease on colony when coring and control for symbiont composition
set.seed(10) # set seed for reproducibility
prok.da.dis <- differentialTest(formula = ~ Condition_At_Sampling + Zoox_composition_New, 
                                  phi.formula = ~ Condition_At_Sampling + Zoox_composition_New,
                                  formula_null = ~ Zoox_composition_New,
                                  phi.formula_null = ~ Condition_At_Sampling + Zoox_composition_New,
                                  test = "Wald", boot = FALSE,
                                  data = ps.prok.annotated,
                                  fdr_cutoff = 0.05)

plot(prok.da.dis, level = "SYMBOL")
prok.da.dis$significant_taxa
```

WOW! This resulted in 327 differentially abundant taxa!!! That is kinda a LOT, but not as many as before, when it was 685 from the previous analysis that accounted for differences due to the old zoox composition groupings.  I would like to be thoughtful in how I represent that data. Maybe I will do the ones that are disease-enriched in one graph, and one that is healthy-enriched in the other graph? I could also plot them all as a heatmap and I DEFINTIELY want to graph them by their BRITE hierarchies. 

Out of curiosity, I am also going to see how many taxa there are with a more stringent p-value cutoff. Save all the data first
```{r}
# write a function to extract the corncob results
extractmods2 <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                        "Std.Error" = model$coefficients[2, 2], 
                        "p" = model$coefficients[2, 4])
  return(result)
}

# save the output as a list using the above function
sig.models <- lapply(prok.da.dis$significant_models, extractmods2)
names(sig.models) <- prok.da.dis$significant_taxa

# save a table of annotations for just the significant genes
sig.fxnl.genes <- filter(unique_gene_ko_annotated, geneID %in% prok.da.dis$significant_taxa)

# turn the output from a list to a dataframe for ease of saving and manipulating
sig.models.df <- plyr::ldply(sig.models, data.frame) %>% 
  left_join(sig.fxnl.genes, by = c(".id" = "geneID")) %>%
  left_join(prok_taxa, by = c(".id" = "contig"))

### Save these data so you don't have to re-run the model ###
write.table(sig.models.df, "../data/Corncob_prok_annotated_results_03.02.25.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

## Process output

How many different genes are there significant at a adjusted p value < 0.001?
```{r}
filter(sig.models.df, p < 0.001) #181 genes
filter(sig.models.df, Estimate > 0) #15 gnes
filter(sig.models.df, Estimate < 0) #312 genes

sig.models.df$KO %>% length #327
sig.models.df$KO %>% unique %>% length #299

sig.taxa <- sig.models.df$KO %>% unique

write(sig.taxa, file = "sig.taxa.KO.txt")
```

I need to try to manipulate the data table output so that I can get a table that shows the KEGG Orthology information clearly.

Try looping through a bunch of KO ids. I will loop through all the KO IDs and get them combined. 
```{r}
# List of KO identifiers
ko_ids <- readLines("sig.taxa.KO.txt")

ko_ids_testgroup <- c("K09125", "K00950", "K15738", "K07457", "K24163", "K00059")

# Create an empty dataframe to store parsed data for all KO IDs
parsed_data <- data.frame(KO = character(), 
                          BRITEhierarchy = character(),
                          Level2 = character(),
                          Level3 = character(),
                          Level4 = character(),
                          Level5 = character(),
                          Level6 = character(),
                          KO_DETAILS = character(),
                          stringsAsFactors = FALSE)

# Loop through each KO ID
for (ko_id in ko_ids) {
  
  # Retrieve information for the KO
  ko_info <- keggGet(ko_id)
  
  # Note which KO ID you are on
  print(paste("Processing KO ID: ", ko_id))
  
  # Check if the KO ID has BRITE hierarchy data
  if (!is.null(ko_info[[1]]$BRITE)) {
    brite <- ko_info[[1]]$BRITE  # Extract the BRITE section
    
    # Initialize variables for hierarchy levels
    brite_hierarchy <- level2 <- level3 <- level4 <- level5 <- level6 <- NA
    
    # Process each line in the BRITE section
    for (line in brite) {
      
      # Determine indentation level (number of leading spaces)
      indentation <- attr(regexpr("^\\s*", line), "match.length")  # Count spaces
      content <- trimws(line)  # Remove leading/trailing spaces
      
      if (indentation == 0) {
          # Top-level category (Level1)
          brite_hierarchy <- content
          level2 <- level3 <- level4 <- level5 <- level6 <- NA  # Reset lower levels
      } else if (indentation == 1) {
          # Second-level category (Level2)
          level2 <- content
          level3 <- level4 <- level5 <- level6 <- NA  # Reset lower levels
      } else if (indentation == 2) {
          # Third-level category (Level3)
          level3 <- content
          level4 <- level5 <- level6 <- NA  # Reset lower level
      } else if (indentation == 3) {
          # KO entry or Fourth-level category (Level4)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data <- rbind(parsed_data, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else { 
            level4 <- content
            level5 <- level6 <- NA
          }
      } else if (indentation == 4) {
          # KO entry or Fifth-level category (Level5)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data <- rbind(parsed_data, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else {
            level5 <- content
            level6 <- NA
          }
      } else if (indentation == 5) {
          # KO entry or or Sixth-level category (Level6)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data <- rbind(parsed_data, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else {
            level6 <- content
          }
      } else if (indentation == 6) {
        # KO entry or or Seventh-level category (Level7)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data <- rbind(parsed_data, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else {
            ko_details <- paste("WARNING, doesn't match KO ID", content)
            parsed_data <- rbind(parsed_data, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } 
      } else {
          # For unexpected cases, add a catch-all (optional)
          warning(paste("Unexpected indentation level encountered: ", indentation, "for the following KO: ", ko_id))
      }
    }
  } else {
    warning(paste("No BRITE data found for KO ID:", ko_id))
  }
}

# View the parsed hierarchy
print(parsed_data)
```

Now, the `parsed_data` data frame is the BRITE hierarchy data for the significant taxa. Ultimately, I just want the KEGG orthology data for each of these KO IDs. 

Then, I will make a data table of the estimates and a heatmap of the corncob results. I will subset only of the KEGG Orthology annotations. Then I will focus on specifically metabolism-based KEGG Orthology annotations since those have an ecological function more than things like "DNA repair" or similar. 
```{r}
KEGG_Orthology_sig_ko <- parsed_data %>%
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism")

# Potential functional hierarchies to look into and graph
parsed_data %>%
  filter(BRITEhierarchy != "KEGG Orthology (KO) [BR:ko00001]") %>%
  dplyr::select(BRITEhierarchy) %>%
  unique
 
KEGG_Orthology_sig_ko$KO %>% unique %>% length

parsed_data$KO %>% unique %>% length

# For each of the KO identifiers, match the gene ID and the associated abundance to that KO. There are multiple duplicate annotations for each KO, and duplicate genes that match a KO, so it is going to be a bit of an "all against all" match up. I also want to add the taxonomy results. 
sig.ko.tax.data <- sig.models.df %>%
  dplyr::select(.id:NAME, DBLINKS, annotation, complete_classification) %>%
  full_join(parsed_data, by = "KO", relationship = "many-to-many")

# Check that there are still just 327 unique genes
sig.ko.tax.data$.id %>% unique %>% length

### Save these data so you have the fully parsed out annotations for each gene (there will be more rows than columns) ###
write.table(sig.ko.tax.data, "../data/Corncob_prok_annotated_results_03.02.25_ANNOTATIONS.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

# Visualize condition at sampling corncob results

It is difficult to know how to parse all these data because the KO identifiers have a lot of different results. Even within one metabolism-based annotation, one gene may play a role in multiple metabolic pathways, so it can have multiple annotations. So when I visualize the data, that is going to be reflected. 

## Metabolism
I want to show the metabolism-based results. I will plot the coefficient since that reflects how much something increased or decreased in the different conditions. 

Outlying questions
- Is it ok if there are duplicate genes on an x or y axis? Maybe just note the # of genes?
- Is it ok to just show the coefficient?
- Should I include a heatmap? 
- How many genes have a metabolism function? 
- What is the function of the genes that are not involved in metabolism?

```{r metabolism graph}
# How many genes have a Metabolism function? 148
sig.ko.tax.data %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism") %>%
  dplyr::select(.id) %>%
  unique %>% 
  nrow

# How many KO ids have a Metabolism function? 131
sig.ko.tax.data %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism") %>%
  dplyr::select(KO) %>%
  unique %>% 
  nrow

# Subset genes with metabolism annotations
sig.metab.genes <- sig.ko.tax.data %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism")

# Make a graph of metabolism genes
# disease associated
ggplot(filter(sig.metab.genes, Estimate > 0), aes(x = Level4, y = Estimate)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y")

# healthy associated
ggplot(filter(sig.metab.genes, Estimate < 0), aes(x = Level4, y = Estimate)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y")

# all together
ggplot(sig.metab.genes, aes(x = Level4, y = Estimate)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y")

# I like the graph where they are together. I will "pretty" this graph up
ggplot(sig.metab.genes, aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  coord_flip() +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group")

ggplot(sig.metab.genes, aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(. ~ Level3, space = "free", scales = "free_x") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# How many disease-assoc genes with metabolism annotations? 11
sig.metab.genes %>% 
  filter(Estimate > 0) %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

# How many healthy-assoc genes with metabolism annotations? 137
sig.metab.genes %>% 
  filter(Estimate < 0) %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

# How many genes in the different categories?
summary_table <- sig.metab.genes %>%
  mutate(DisHealth = ifelse(Estimate > 0, "DiseaseAssoc", "HealthyAssoc")) %>%  # Classify results as diseased or healthy associated
  dplyr::select(.id, DisHealth, Level3) %>% # select columns of interest
  distinct(.id, DisHealth, Level3) %>% # Keep unique combinations of Gene, StatSign, and Category
  group_by(Level3, DisHealth) %>%
  summarize(UniqueGenes = n_distinct(.id), .groups = "drop") %>% # count unique genes
  spread(DisHealth, UniqueGenes)

print(summary_table)
```
I want to make a graph of the genes involved in natural product relevant metabolism. This will be a subset of the above graph that is hopefully easier to digest because there won't be as many categories and genes as the above graph. 
```{r}
sig.metab.genes.subsetforgraph <- sig.metab.genes %>%
  filter(Level3 %in% c("09109 Metabolism of terpenoids and polyketides", 
                       "09110 Biosynthesis of other secondary metabolites", 
                       "09111 Xenobiotics biodegradation and metabolism")) %>%
  mutate(DiseaseHealth = ifelse(Estimate > 0, "Disease", "Healthy"))

ggplot(sig.metab.genes.subsetforgraph, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Graph at the gene level, regardless of grouping
ggplot(sig.metab.genes.subsetforgraph, aes(x = NAME, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(size = 3) +
  coord_flip() +
  facet_grid(DiseaseHealth ~ ., space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "Gene") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Now add groupings, which get complicated since some genes fall under multiple groupings. 
ggplot(sig.metab.genes.subsetforgraph, aes(x = NAME, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(size = 3) +
  facet_grid(Level4 ~ ., space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "Gene") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_flip()

# How many genes and KOs are reflected in this figure?
sig.metab.genes.subsetforgraph %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

sig.metab.genes.subsetforgraph %>% 
  dplyr::select(KO) %>%
  unique %>%
  nrow

# FINAL GRAPH THAT COULD BE ADDED TO A PAPER
ggplot(sig.metab.genes.subsetforgraph, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology Metabolism Group") +  
  geom_vline(xintercept = 0, linetype = "dashed")

# What about cofactors and vitamins? NOT MY FAVE - I DON"T THINK THIS ADDS MUCH
sig.metab.genes.vitamins <- sig.metab.genes %>%
  filter(Level3 %in% c("09108 Metabolism of cofactors and vitamins")) %>%
  mutate(DiseaseHealth = ifelse(Estimate > 0, "Disease", "Healthy"))

ggplot(sig.metab.genes.vitamins, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology Metabolism Group") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
I like the final graph (not the vitamin one)
Things to mention in the paper
- number of different genes that were healthy or disease associated
- specific names of genes that were part of metabolizing natural or bioactive products

### Add taxonomy
```{r}
## SPLIT APART THE TAXONOMY INFO - CAN I COLOR BY TAXA?
sig.metab.genes.subsetforgraph.tax <- sig.metab.genes.subsetforgraph %>%
    separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE)

ggplot(sig.metab.genes.subsetforgraph.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, aes(color = Family)) +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology Metabolism Group") +  
  geom_vline(xintercept = 0, linetype = "dashed")
```


## Everything else?

It looks like 310 out of the 685 genes were involved in metabolism functions. What about the other half? I am going to go through and assess and graph the other genes. First, I need to subset the data to only look at the remaining genes. 

```{r}
# Get a list of all the genes from the first graph
metab.genes.id <- sig.metab.genes$.id %>% unique

# Subset the data to exclude the metab.genes.id values
sig.nonmetab.genes <- sig.ko.tax.data[!sig.ko.tax.data$.id %in% metab.genes.id, ]

# Double check how many unique genes in the new data frame (179)
sig.nonmetab.genes$.id %>% unique %>% length

# Subset the KO data since it includes all 179 genes
sig.nonmetab.genes.KO <- sig.nonmetab.genes %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]")

# How many level 2 groups are there?
sig.nonmetab.genes.KO$Level2 %>% factor %>% levels
sig.nonmetab.genes.KO$Level3 %>% factor %>% levels

# Try out a couple ways to graph the data
# Make a graph of not metabolism genes

# disease associated
ggplot(filter(sig.nonmetab.genes.KO, Estimate > 0), aes(x = Level4, y = Estimate)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y")

# healthy associated
ggplot(filter(sig.nonmetab.genes.KO, Estimate < 0), aes(x = Level4, y = Estimate)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y")

# all genes
ggplot(sig.nonmetab.genes.KO, aes(x = Level4, y = Estimate)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y")

# I like the graph where they are together. I will "pretty" this graph up
ggplot(sig.nonmetab.genes.KO, aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  coord_flip() +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group")

ggplot(sig.nonmetab.genes.KO, aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(. ~ Level2 + Level3, space = "free", scales = "free_x") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# How many disease-assoc genes with metabolism annotations? 4
sig.nonmetab.genes.KO %>% 
  filter(Estimate > 0) %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

# How many healthy-assoc genes with metabolism annotations? 175
sig.nonmetab.genes.KO %>% 
  filter(Estimate < 0) %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

# How many genes in the different categories?
summary_table_nonmetab <- sig.nonmetab.genes.KO %>%
  mutate(DisHealth = ifelse(Estimate > 0, "DiseaseAssoc", "HealthyAssoc")) %>%  # Classify results as diseased or healthy associated
  dplyr::select(.id, DisHealth, Level3) %>% # select columns of interest
  distinct(.id, DisHealth, Level3) %>% # Keep unique combinations of Gene, StatSign, and Category
  group_by(Level3, DisHealth) %>%
  summarize(UniqueGenes = n_distinct(.id), .groups = "drop") %>% # count unique genes
  spread(DisHealth, UniqueGenes)

print(summary_table_nonmetab)
```

Rather, i'd like to make a graph, without the protein families: genetic informatino processing, or unclassified: metabolism groups, and poorly characterized grous since they dont have genes that have more r less DA genes in disease or healthy corals. 

I also don't really want to focus on unclassified proteins. 

```{r}
sig.nonmetab.genes.KO.trimmed <- sig.nonmetab.genes.KO %>%
  filter(Level3 != "09182 Protein families: genetic information processing") %>%
  filter(Level3 != "09191 Unclassified: metabolism") %>%
  filter(Level3 != "09192 Unclassified: genetic information processing") %>%
  filter(Level3 != "09193 Unclassified: signaling and cellular processes") %>%
  filter(Level3 != "09194 Poorly characterized")

# How many genes did I lose out of the remaining 179? 116/179 (lost 63 genes)
sig.nonmetab.genes.KO.trimmed$.id %>% unique %>% length

ggplot(sig.nonmetab.genes.KO.trimmed, aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(. ~ Level2 + Level3, space = "free", scales = "free_x") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Many of the differentially abundant genes that I would hypothesize are playing a role in microbial-based disease susceptibility or resilience are in the Environmental Information Processing, Cellular processes, Drug resistance, and protein families, signalling and cellular processes groups. I will make a graph with just those categories. I will ignore the human disease ones because I anticipate those are relevant to human pathogens, which I don't think would be abundant on the reefs. 

```{r}
sig.nonmetab.genes.KO.subsetforgraph <- sig.nonmetab.genes.KO %>%
  filter(Level3 %in% c("09131 Membrane transport",
                       "09132 Signal transduction", 
                       "09143 Cell growth and death", 
                       "09145 Cellular community - prokaryotes",
                       "09175 Drug resistance: antimicrobial",
                       "09176 Drug resistance: antineoplastic",
                       "09183 Protein families: signaling and cellular processes"))

ggplot(sig.nonmetab.genes.KO.subsetforgraph, aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(. ~ Level2 + Level3, space = "free", scales = "free_x") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# There are a ton of transporters so I am going to get rid of those. 
sig.nonmetab.genes.KO.subsetforgraph_2 <- sig.nonmetab.genes.KO.subsetforgraph %>%
  filter(Level4 != "02000 Transporters") %>%
  filter(Level4 != "02010 ABC transporters") %>%
  mutate(DiseaseHealth = ifelse(Estimate > 0, "Disease", "Healthy"))

# Graph at the gene level, regardless of grouping
ggplot(sig.nonmetab.genes.KO.subsetforgraph_2, aes(x = NAME, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(size = 3) +
  coord_flip() +
  facet_grid(DiseaseHealth ~ ., space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "Gene") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Now add groupings, which get complicated since some genes fall under multiple groupings. 
ggplot(sig.nonmetab.genes.KO.subsetforgraph_2, aes(x = NAME, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(size = 3) +
  facet_grid(. ~ DiseaseHealth + Level4, space = "free", scales = "free_x") +
  labs(y = "Differential Abundance Coefficient", x = "Gene") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# How many genes and KOs are reflected in this figure? 28
sig.nonmetab.genes.KO.subsetforgraph_2 %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

sig.nonmetab.genes.KO.subsetforgraph_2 %>%  #24
  dplyr::select(KO) %>%
  unique %>%
  nrow

## POTENTIAL MANUSCRIPT FIGURE - Maybe keep the transporters for now?
ggplot(sig.nonmetab.genes.KO.subsetforgraph, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed")
```

### Add taxonomy?
```{r}
## SPLIT APART THE TAXONOMY INFO - CAN I COLOR BY TAXA?
sig.nonmetab.genes.KO.subsetforgraph.tax <- sig.nonmetab.genes.KO.subsetforgraph %>%
    separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE)

ggplot(sig.nonmetab.genes.KO.subsetforgraph.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, aes(color = Family)) +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed")
```

# MS ANALYSIS - FATE

## Corncob

I would like to analyze the Colony fate category. This time, I will use that, and again control for zoox. This will be interesting to add to the analysis and is something the co-authors and likely reviewers will ask for. 
```{r}
## Set levels
sample_data(ps.prok.annotated)$Fate_after_SP1 <-
  factor(sample_data(ps.prok.annotated)$Fate_after_SP1, 
         levels = c("Unaffected", "Recovered", "Will be diseased", "Diseased"))

head(sample_data(ps.prok.annotated)$Fate_after_SP1, 10) # levels look good

sample_data(ps.prok.annotated)$Zoox_composition_New <- 
  factor(sample_data(ps.prok.annotated)$Zoox_composition_New, 
         levels = c("B", "B_dom_plus_C", "B_dom_plus_CD", "B_dom_plus_D", "BD", "CD"))

head(sample_data(ps.prok.annotated)$Zoox_composition_New, 10) # levels look good

## Test disease on colony when coring and control for symbiont composition
set.seed(10) # set seed for reproducibility
prok.da.fate <- differentialTest(formula = ~ Fate_after_SP1 + Zoox_composition_New, 
                                  phi.formula = ~ Fate_after_SP1 + Zoox_composition_New,
                                  formula_null = ~ Zoox_composition_New,
                                  phi.formula_null = ~ Fate_after_SP1 + Zoox_composition_New,
                                  test = "Wald", boot = FALSE,
                                  data = ps.prok.annotated,
                                  fdr_cutoff = 0.05)

plot(prok.da.fate, level = "SYMBOL")
prok.da.fate$significant_taxa
```
Looks like there are 392 differentially abundant genes by different categories! I'll dig into these to see how best to represent these different groupings. 

Save all the data. NEED TO MAKE A NEW FUNCTION TO GET THESE DATA OUT OF CORNCOB
```{r}
# write a function to extract the corncob results
extractmods3 <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2:4, 1], 
                        "Std.Error" = model$coefficients[2:4, 2], 
                        "p" = model$coefficients[2:4, 4], 
                       "Comparison" = c("Unaffected_vs_Recovered",
                                        "Unaffected_vs_WillbeDiseased",
                                        "Unaffected_vs_Diseased"))
  return(result)
}

# save the output as a list using the above function
sig.models.fate <- lapply(prok.da.fate$significant_models, extractmods3)
names(sig.models.fate) <- prok.da.fate$significant_taxa

# save a table of annotations for just the significant genes
sig.fxnl.genes.fate <- filter(unique_gene_ko_annotated, geneID %in% prok.da.fate$significant_taxa)

# turn the output from a list to a dataframe for ease of saving and manipulating
sig.models.df.fate <- plyr::ldply(sig.models.fate, data.frame) %>% 
  mutate(significant = ifelse(p < 0.05, TRUE, FALSE)) %>%
  left_join(sig.fxnl.genes.fate, by = c(".id" = "geneID")) %>%
  left_join(prok_taxa, by = c(".id" = "contig"))

### Save these data so you don't have to re-run the model ###
write.table(sig.models.df.fate, "../data/Corncob_prok_annotated_results_FATE_07.28.25.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

## Process output

How many different genes are there significant at a adjusted p value < 0.001?
```{r}
filter(sig.models.df.fate, p < 0.001) #162 comparisons
filter(sig.models.df.fate, Estimate > 0) #536 comparisons 
filter(sig.models.df.fate, Estimate < 0) #640 comparisons

sig.models.df.fate$KO %>% length #1176
sig.models.df.fate$KO %>% unique %>% length #356

sig.taxa.fate <- sig.models.df.fate$KO %>% unique

write(sig.taxa.fate, file = "sig.taxa.KO.fate.txt")
```

I need to try to manipulate the data table output so that I can get a table that shows the KEGG Orthology information clearly.

Try looping through a bunch of KO ids. I will loop through all the KO IDs and get them combined. 
```{r}
# List of KO identifiers
ko_ids.fate <- readLines("sig.taxa.KO.fate.txt")

ko_ids_testgroup <- c("K09125", "K00950", "K15738", "K07457", "K24163", "K00059")

# Create an empty dataframe to store parsed data for all KO IDs
parsed_data.fate <- data.frame(KO = character(), 
                          BRITEhierarchy = character(),
                          Level2 = character(),
                          Level3 = character(),
                          Level4 = character(),
                          Level5 = character(),
                          Level6 = character(),
                          KO_DETAILS = character(),
                          stringsAsFactors = FALSE)

# Loop through each KO ID
for (ko_id in ko_ids.fate) {
  
  # Retrieve information for the KO
  ko_info <- keggGet(ko_id)
  
  # Note which KO ID you are on
  print(paste("Processing KO ID: ", ko_id))
  
  # Check if the KO ID has BRITE hierarchy data
  if (!is.null(ko_info[[1]]$BRITE)) {
    brite <- ko_info[[1]]$BRITE  # Extract the BRITE section
    
    # Initialize variables for hierarchy levels
    brite_hierarchy <- level2 <- level3 <- level4 <- level5 <- level6 <- NA
    
    # Process each line in the BRITE section
    for (line in brite) {
      
      # Determine indentation level (number of leading spaces)
      indentation <- attr(regexpr("^\\s*", line), "match.length")  # Count spaces
      content <- trimws(line)  # Remove leading/trailing spaces
      
      if (indentation == 0) {
          # Top-level category (Level1)
          brite_hierarchy <- content
          level2 <- level3 <- level4 <- level5 <- level6 <- NA  # Reset lower levels
      } else if (indentation == 1) {
          # Second-level category (Level2)
          level2 <- content
          level3 <- level4 <- level5 <- level6 <- NA  # Reset lower levels
      } else if (indentation == 2) {
          # Third-level category (Level3)
          level3 <- content
          level4 <- level5 <- level6 <- NA  # Reset lower level
      } else if (indentation == 3) {
          # KO entry or Fourth-level category (Level4)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data.fate <- rbind(parsed_data.fate, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else { 
            level4 <- content
            level5 <- level6 <- NA
          }
      } else if (indentation == 4) {
          # KO entry or Fifth-level category (Level5)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data.fate <- rbind(parsed_data.fate, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else {
            level5 <- content
            level6 <- NA
          }
      } else if (indentation == 5) {
          # KO entry or or Sixth-level category (Level6)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data.fate <- rbind(parsed_data.fate, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else {
            level6 <- content
          }
      } else if (indentation == 6) {
        # KO entry or or Seventh-level category (Level7)
          if (grepl("^K[0-9]+", content)) {  # Check if it's a KO entry
            ko_details <- content
            parsed_data.fate <- rbind(parsed_data.fate, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } else {
            ko_details <- paste("WARNING, doesn't match KO ID", content)
            parsed_data.fate <- rbind(parsed_data.fate, data.frame(KO = ko_id, 
                                                         BRITEhierarchy = brite_hierarchy,
                                                         Level2 = level2,
                                                         Level3 = level3,
                                                         Level4 = level4,
                                                         Level5 = level5,
                                                         Level6 = level6,
                                                         KO_DETAILS = ko_details,
                                                         stringsAsFactors = FALSE))
          } 
      } else {
          # For unexpected cases, add a catch-all (optional)
          warning(paste("Unexpected indentation level encountered: ", indentation, "for the following KO: ", ko_id))
      }
    }
  } else {
    warning(paste("No BRITE data found for KO ID:", ko_id))
  }
}

# View the parsed hierarchy
print(parsed_data.fate)
```


Now, the `parsed_data.fate` data frame is the BRITE hierarchy data for the significant taxa. Ultimately, I just want the KEGG orthology data for each of these KO IDs. 

Then, I will make a data table of the estimates and a heatmap of the corncob results. I will subset only of the KEGG Orthology annotations. Then I will focus on specifically metabolism-based KEGG Orthology annotations since those have an ecological function more than things like "DNA repair" or similar. 
```{r}
KEGG_Orthology_sig_ko.fate <- parsed_data.fate %>%
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism")

# Potential functional hierarchies to look into and graph
parsed_data.fate %>%
  filter(BRITEhierarchy != "KEGG Orthology (KO) [BR:ko00001]") %>%
  dplyr::select(BRITEhierarchy) %>%
  unique
 
#did i capture all the KOs? NO - only 165 out of all of them
KEGG_Orthology_sig_ko.fate$KO %>% unique %>% length

parsed_data.fate$KO %>% unique %>% length

# For each of the KO identifiers, match the gene ID and the associated abundance to that KO. There are multiple duplicate annotations for each KO, and duplicate genes that match a KO, so it is going to be a bit of an "all against all" match up. I also want to add the taxonomy results. 
sig.ko.tax.data.fate <- sig.models.df.fate %>%
  dplyr::select(.id:NAME, DBLINKS, annotation, complete_classification) %>%
  full_join(parsed_data.fate, by = "KO", relationship = "many-to-many")

# Check that there are still just 392 unique genes
sig.ko.tax.data.fate$.id %>% unique %>% length

### Save these data so you have the fully parsed out annotations for each gene (there will be more rows than columns) ###
write.table(sig.ko.tax.data.fate, "../data/Corncob_prok_annotated_results_FATE_07.28.25_ANNOTATIONS.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

## Visualize/explore fate

It is difficult to know how to parse all these data because the KO identifiers have a lot of different results. Even within one metabolism-based annotation, one gene may play a role in multiple metabolic pathways, so it can have multiple annotations. So when I visualize the data, that is going to be reflected. 

Additionally, since I did a corncob test with multiple levels, I am not sure I want to show the output of those that are NOT significant, since it might be misleading. So I'll have to think about how to display that. 

### Metabolism
I want to show the metabolism-based results. I will plot the coefficient since that reflects how much something increased or decreased in the different conditions. 

Outlying questions
- Is it ok if there are duplicate genes on an x or y axis? Maybe just note the # of genes?
- Is it ok to just show the coefficient?
- Should I include a heatmap? 
- How many genes have a metabolism function? 
- What is the function of the genes that are not involved in metabolism?

```{r metabolism graph}
# How many genes have a Metabolism function? 180
sig.ko.tax.data.fate %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism") %>%
  dplyr::select(.id) %>%
  unique %>% 
  nrow

# How many KO ids have a Metabolism function? 165
sig.ko.tax.data.fate %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism") %>%
  dplyr::select(KO) %>%
  unique %>% 
  nrow

# Subset genes with metabolism annotations - need to facet with the different categories...
sig.metab.genes.fate <- sig.ko.tax.data.fate %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(Level2 == "09100 Metabolism") %>%
  mutate(Comparison = factor(Comparison, levels = c("Unaffected_vs_Recovered", "Unaffected_vs_WillbeDiseased", "Unaffected_vs_Diseased")))

# Make a graph of metabolism genes
# need to facet by the groupings
ggplot(sig.metab.genes.fate, aes(x = Level4, y = Estimate)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y")

# change the color based on whether the relationship is significant - 
ggplot(sig.metab.genes.fate, aes(x = Level4, y = Estimate, color = significant)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y")

# ultimately, I think showing the results of the non-signficant relationships will just be more confusing 
# only significant comparisons. 
ggplot(filter(sig.metab.genes.fate, significant == TRUE), aes(x = Level4, y = Estimate, color = significant)) +
  geom_point() +
  coord_flip() +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y")

# I like showing only the significant relationships. I will pretty this up. 
ggplot(filter(sig.metab.genes.fate, significant == TRUE), aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  coord_flip() +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group")

# How many genes in the different categories?
summary_table <- sig.metab.genes.fate %>%
  filter(significant == TRUE) %>% 
  mutate(posneg = ifelse(Estimate > 0, "positive", "negative")) %>%
  dplyr::select(.id, Comparison, posneg, Level3) %>% #slect columns of interest
  distinct(.id, Comparison, posneg, Level3) %>% 
  group_by(Level3, posneg, Comparison) %>%
  summarize(UniqueGenes = n_distinct(.id), .groups = "drop") %>% 
  spread(posneg, UniqueGenes)

print(summary_table)
```
I want to make a graph of the genes involved in natural product relevant metabolism. This will be a subset of the above graph that is hopefully easier to digest because there won't be as many categories and genes as the above graph. 
```{r}
sig.metab.genes.fate.subsetforgraph <- sig.metab.genes.fate %>%
  filter(significant == TRUE) %>%
  filter(Level3 %in% c("09109 Metabolism of terpenoids and polyketides", 
                       "09110 Biosynthesis of other secondary metabolites", 
                       "09111 Xenobiotics biodegradation and metabolism"))

ggplot(sig.metab.genes.fate.subsetforgraph, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Metabolism Group") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# How many genes and KOs are reflected in this figure? 16
sig.metab.genes.fate.subsetforgraph %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

# GRAPH THAT COULD BE ADDED TO A PAPER
ggplot(sig.metab.genes.fate.subsetforgraph, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology Metabolism Group") +  
  geom_vline(xintercept = 0, linetype = "dashed")

# What about cofactors and vitamins? Let's add this to the previous info...
sig.metab.genes.fate.subsetforgraph.vits <- sig.metab.genes.fate %>%
  filter(significant == TRUE) %>%
  filter(Level3 %in% c("09109 Metabolism of terpenoids and polyketides", 
                       "09110 Biosynthesis of other secondary metabolites", 
                       "09111 Xenobiotics biodegradation and metabolism",
                       "09108 Metabolism of cofactors and vitamins"))

ggplot(sig.metab.genes.fate.subsetforgraph.vits, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology Metabolism Group") +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_vline(xintercept = 0, linetype = "dashed")
```
I like this graph A LOT! Indeed, I am thinking this analysis is better suited to the paper more and more and more. 
Some trends I am noticing thaat would be good to include in the paper:
- Recovered corals are enriched for a LOT of vitamin metabolism pathways and some secondary metabolite pathways --> does this further HELP the coral maybe?
- The diseased corals have significant declines in vitamin metabolism and secondary metabolites, especially antibiotics --> loss of potentially protective functional capacity of the coral microbiome

#### Add taxonomy
```{r}
## SPLIT APART THE TAXONOMY INFO - CAN I COLOR BY TAXA?
sig.metab.genes.fate.subsetforgraph.vits.tax <- sig.metab.genes.fate.subsetforgraph.vits %>%
    separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE)

ggplot(sig.metab.genes.fate.subsetforgraph.vits.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, aes(color = Family)) +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology Metabolism Group") +  
  geom_vline(xintercept = 0, linetype = "dashed")
```
I do like the addition of the taxonomic information. I think this is useful and will be good to include down the road! 

### Everything else?

It looks like 180 out of the 392 genes were involved in metabolism functions. What about the other half? I am going to go through and assess and graph the other genes. First, I need to subset the data to only look at the remaining genes. 

```{r}
# Get a list of all the genes from the first graph
metab.genes.fate.id <- sig.metab.genes.fate$.id %>% unique

# Subset the data to exclude the metab.genes.id values
sig.nonmetab.genes.fate <- sig.ko.tax.data.fate[!sig.ko.tax.data.fate$.id %in% metab.genes.fate.id, ]

# Double check how many unique genes in the new data frame (212 - good! adds up to 392 with 180)
sig.nonmetab.genes.fate$.id %>% unique %>% length

# Subset the KO data since it includes all 179 genes
sig.nonmetab.genes.fate.KO <- sig.nonmetab.genes.fate %>% 
  filter(BRITEhierarchy == "KEGG Orthology (KO) [BR:ko00001]") %>%
  filter(significant == TRUE) %>%
  mutate(Comparison = factor(Comparison, levels = c("Unaffected_vs_Recovered", "Unaffected_vs_WillbeDiseased", "Unaffected_vs_Diseased")))

# How many level 2 groups are there?
sig.nonmetab.genes.fate.KO$Level2 %>% factor %>% levels
sig.nonmetab.genes.fate.KO$Level3 %>% factor %>% levels

# Try out a couple ways to graph the data
# Make a graph of not metabolism genes

# all genes - only significant comparisons represented
ggplot(sig.nonmetab.genes.fate.KO, aes(x = Level4, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  coord_flip() +
  facet_grid(Level2 + Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(y = "Differential Abundance Coefficient", x = "KEGG Orthology Group") +
  geom_hline(yintercept = 0, linetype = "dashed")

# How many genes in the different categories?
summary_table_nonmetab <- sig.nonmetab.genes.fate.KO %>%
  mutate(posneg = ifelse(Estimate > 0, "positive", "negative")) %>%
  dplyr::select(.id, Comparison, posneg, Level3) %>% #slect columns of interest
  distinct(.id, Comparison, posneg, Level3) %>% 
  group_by(Level3, posneg, Comparison) %>%
  summarize(UniqueGenes = n_distinct(.id), .groups = "drop") %>% # count unique genes
  spread(posneg, UniqueGenes)

print(summary_table_nonmetab)
```
Okay, there's a lot of interesting stuff here. Overall, I don't need some of these groups in the resultant graph. For example, all the human disease stuff is pretty useless, except beta-Lactam resistance. I don't need any unclassified groupings. I don't need "metabolism" or "genetic information processing" either. 

I also don't really want to focus on unclassified proteins.

Instead, I'd like to focus on groupings that would likely lead to potential disease or health of the coral, not stuff (transcription) required for normal cell growth. 

Many of the differentially abundant genes that I would hypothesize are playing a role in microbial-based disease susceptibility or resilience are in the Environmental Information Processing, Cellular processes, Drug resistance, and protein families, signalling and cellular processes groups. I will make a graph with just those categories. I will ignore the human disease ones because I anticipate those are relevant to human pathogens, which I don't think would be abundant on the reefs. 

```{r}
sig.nonmetab.genes.fate.KO.subsetforgraph <- sig.nonmetab.genes.fate.KO %>%
  filter(Level3 %in% c("09131 Membrane transport",
                       "09132 Signal transduction", 
                       "09142 Cell motility",
                       "09143 Cell growth and death", 
                       "09145 Cellular community - prokaryotes",
                       "09175 Drug resistance: antimicrobial",
                       "09183 Protein families: signaling and cellular processes"))

# How many genes and KOs are reflected in this figure? 73
sig.nonmetab.genes.fate.KO.subsetforgraph %>% 
  dplyr::select(.id) %>%
  unique %>%
  nrow

sig.nonmetab.genes.fate.KO.subsetforgraph %>%  #64
  dplyr::select(KO) %>%
  unique %>%
  nrow

## POTENTIAL MANUSCRIPT FIGURE - Maybe keep the transporters for now
ggplot(sig.nonmetab.genes.fate.KO.subsetforgraph, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level2 + Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed")
```
Thius is super interesting again. Like with the stuff previously, I think these fate categories are putting together a much much better story. A couple of thinks I'm noticiing:
- The recovered grouping has these signalling things ENRICHED relative to unaffected corals. This tells me these are things that may be helping the coral be resilient to SCTLD over a long time frame. 
- In contrast, many gene groups that are helping corals be resilient are DEPLETED in diseased corals relative to the unaffected corals. For example, disease corals have depleted secretion systems and quorum sensing. Interestingly, the Will be diseased grouping is particularly telling here. 

### Add taxonomy?
```{r}
## SPLIT APART THE TAXONOMY INFO - CAN I COLOR BY TAXA?
sig.nonmetab.genes.fate.KO.subsetforgraph.tax <- sig.nonmetab.genes.fate.KO.subsetforgraph %>%
    separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE)

ggplot(sig.nonmetab.genes.fate.KO.subsetforgraph.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, aes(color = Family)) +
  facet_grid(Level2 + Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed")
```

### Will be diseased deep dive

Honestly, the "Will be diseased" category is super interesting. I'd be curious to see the abundance of these differentially abundant genes across the disease history versus not disease history spectrum or in realtion to how soon after the sampling they get disease. but it is a bit confusing. I also think this may not be the most useful comparison since the dispersion wasn't significant and there are only 3 in that category that have a history of disease. 
```{r}

```



# MS FIGURES

This will use data and tables from the prevous code chunks, from the section `MS ANALYSIS - KEGG Diff. Abundance`

## Supplement - big graphs
```{r}
# Only genes with "09100 Metabolism" annotation in the KEGG Orthology (KO) [BR:ko00001] (148 genes)
a <- ggplot(sig.metab.genes, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +
  geom_vline(xintercept = 0, linetype = "dashed")

# Any remaining genes with any KO annotation except "09100 Metabolism" (169 genes)
b <- ggplot(sig.nonmetab.genes.KO, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7) +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +
  geom_vline(xintercept = 0, linetype = "dashed")

a
#ggsave("../figures/DA_metabolism_genes_03.02.25.pdf", width = 6, height = 11)

b
#ggsave("../figures/DA_non-metabolism_genes_03.02.25.pdf", width = 6, height = 11)
```

## Summary tables
```{r}
summary_table <- sig.metab.genes %>%
  mutate(DisHealth = ifelse(Estimate > 0, "DiseaseAssoc", "HealthyAssoc")) %>%  # Classify results as diseased or healthy associated
  dplyr::select(.id, DisHealth, Level3) %>% # select columns of interest
  distinct(.id, DisHealth, Level3) %>% # Keep unique combinations of Gene, StatSign, and Category
  group_by(Level3, DisHealth) %>%
  summarize(UniqueGenes = n_distinct(.id), .groups = "drop") %>% # count unique genes
  spread(DisHealth, UniqueGenes)

print(summary_table)

# How many genes in the different categories?
summary_table_nonmetab <- sig.nonmetab.genes.KO %>%
  mutate(DisHealth = ifelse(Estimate > 0, "DiseaseAssoc", "HealthyAssoc")) %>%  # Classify results as diseased or healthy associated
  dplyr::select(.id, DisHealth, Level3) %>% # select columns of interest
  distinct(.id, DisHealth, Level3) %>% # Keep unique combinations of Gene, StatSign, and Category
  group_by(Level3, DisHealth) %>%
  summarize(UniqueGenes = n_distinct(.id), .groups = "drop") %>% # count unique genes
  spread(DisHealth, UniqueGenes)

print(summary_table_nonmetab)
```

## Main Figure
I want to show the genes that are differentially abundant and may be involved in signalling, natural product generation and processing. 
Use a color-blind friendly palette from hex codes here: https://davidmathlogic.com/colorblind/#%23332288-%23117733-%2344AA99-%2388CCEE-%23DDCC77-%23CC6677-%23AA4499-%23882255
```{r}
# Metabolism genes subset
c <- ggplot(sig.metab.genes.subsetforgraph.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, size = 3, aes(color = Family)) +
  facet_grid(Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed") +
  xlim(-5, 2) +
  scale_color_manual(values = c("#117733", "#CC6677", "#AA4499", "#332288"))
c



# Non-metabolism genes subset
d <- ggplot(sig.nonmetab.genes.KO.subsetforgraph.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, size = 3, aes(color = Family)) +
  facet_grid(Level2 + Level3 ~ ., space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed") +
  xlim(-5, 2) +
  scale_color_manual(values = c("#CC6677","#AA4499", "#332288", "#88CCEE" ))
d


wrap_plots(c,d, nrow = 2, heights = c(1,1.3), axis_titles = "collect")
#ggsave("../figures/DA_MainFig_genes_03.02.25.pdf", width = 8, height = 7)
```
## Main FATE Figure
```{r}
# Metabolism genes subset
c <- ggplot(sig.metab.genes.fate.subsetforgraph.vits.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, size = 3, aes(color = Family)) +
  facet_grid(Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed") +
#  xlim(-5, 2) +
  scale_color_manual(values = c("#B2DF8A", "#33A02C", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99","#B15928"))
c

# Non-metabolism genes subset
d <- ggplot(sig.nonmetab.genes.fate.KO.subsetforgraph.tax, aes(x = Estimate, y = Level4)) +
  geom_errorbar(aes(xmin = Estimate-Std.Error, xmax = Estimate+Std.Error), width = 0.2, color = "gray") +
  geom_point(alpha = 0.7, size = 3, aes(color = Family)) +
  facet_grid(Level2 + Level3 ~ Comparison, space = "free", scales = "free_y") +
  labs(x = "Differential Abundance Coefficient", y = "KEGG Orthology") +  
  geom_vline(xintercept = 0, linetype = "dashed") +
 # xlim(-5, 2) +
  scale_color_manual(values = c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99","#B15928"))
d

wrap_plots(c,d, nrow = 2, heights = c(1.3,1), axis_titles = "collect")
#ggsave("../figures/DA_MainFig_genes_FATE_07.28.25.pdf", width = 10, height = 9)
```


## Supplement - gene names of just the subset genes
```{r}
sig.metab.genes.subsetforgraph.tax
sig.nonmetab.genes.KO.subsetforgraph.tax

write.table(sig.metab.genes.subsetforgraph.tax, "../data/DA_MainFig_Gene_Information_a_03.02.25.txt", sep="\t", row.names = FALSE, col.names = TRUE)
write.table(sig.nonmetab.genes.KO.subsetforgraph.tax, "../data/DA_MainFig_Gene_Information_b_03.02.25.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

## Fate Supplement - gene names of just the subset fate genes for MS
```{r}
sig.metab.genes.fate.subsetforgraph.vits.tax
sig.nonmetab.genes.fate.KO.subsetforgraph.tax

write.table(sig.metab.genes.fate.subsetforgraph.vits.tax, "../data/DA_MainFig_Gene_Information_FATE_a_07.29.25.txt", sep="\t", row.names = FALSE, col.names = TRUE)
write.table(sig.nonmetab.genes.fate.KO.subsetforgraph.tax, "../data/DA_MainFig_Gene_Information_FATE_b_07.29.25.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

# ZOOX ANALYSIS - RRC Report

In addition to focusing on the diseased/unaffected colonies, one of the main deliverables of the RRC report was investigating microbial genes associated with the symbionts. Since I have the new symbiont composition categories, i can do a corncob analysis to look for differentially abundant genes. 

I will use the similar code to that I have above. 

## Data 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# Load metadata
metadata <- as.data.frame(read_excel("../data/RRC_sample_CoreSiteData_v10_RRCCoreData_SP1.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)

# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)

# Read in the taxonomy
prok_taxa <- read.table("../data/FLK_prok_coassembly_tax_mmseqs2.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")
```

### KO data
```{r}
# load KO and Gene ids
# Read all lines from the file
lines <- readLines("../data/user_ko.txt")

# Split each line by tab ("\t")
split_lines <- strsplit(lines, "\t")

# Convert the split lines into a data frame
gene_ko <- do.call(rbind, lapply(split_lines, function(x) {
  if (length(x) == 1) {
    c(x[1], NA)  # If only one column, fill second column with NA
  } else {
    x  # Keep both columns
  }
}))

# Assign column names
colnames(gene_ko) <- c("geneID", "KO")

dim(gene_ko)
unique(gene_ko) %>% dim

# Convert to a data frame
gene_ko <- as.data.frame(gene_ko, stringsAsFactors = FALSE)

# how many unique KO ids?
gene_ko$KO %>% unique %>% length

# View the resulting data
print(head(gene_ko))

# Read in the table of the annotations
ko_info <- read_excel("../data/kegg_data.xlsx", col_names = TRUE, na = "")

# Get rid of the "KO" in the first column
ko_info$ENTRY <- str_split_i(ko_info$ENTRY, " ", 1)

head(ko_info)
dim(ko_info)

# How many unique KOs? 4563
ko_info$KO_ID %>% unique %>% length
ko_info$ENTRY %>% unique %>% length

# There's too many duplicate rows. Should have only 4,563
ko_info_nodups <- unique(ko_info)

# add the annotation info to the gene_ko documant
gene_ko_join <- gene_ko %>%
  left_join(ko_info_nodups, by = c("KO" = "ENTRY"))

# How many genes are annotated? 32,980
sum(!is.na(gene_ko_join$KO))

# Filter to only include the annotated columns (should end with 32,980 rows)
gene_ko_annotated <- gene_ko_join[!is.na(gene_ko_join$KO), ]

# It looks like there are some duplicated gene IDs
gene_ko_annotated$geneID %>% unique %>% length

# Get rid of duplicate geneID's
unique_gene_ko_annotated<- gene_ko_annotated[!duplicated(gene_ko_annotated$geneID), ]
dim(unique_gene_ko_annotated)

# Rename the rows 
rownames(unique_gene_ko_annotated) <- unique_gene_ko_annotated[,"geneID"]
```

## Phyloseq object
Start with a phyloseq object jsut with the count data (no annotation information). The annotations aren't needed for the PCAs I want to make. 
```{r}
## Make phyloseq object of count-related data 
idx <- match(colnames(prokreads_nozero), rownames(metadata))
metadata_prok <- metadata[idx,] # subset the metadata to only the sequenced samples

NUMREADS = otu_table(prokreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_prok)
FXN = tax_table(as.matrix(unique_gene_ko_annotated))

ps.prok.annotated <- phyloseq(NUMREADS, META, FXN)
ps.prok.annotated
```

## Corncob

I will use corncob to test the hypothesis that functional genes change in abundance between corals containing different compositions of symbiodiniaceae in their tissues (`Zoox_composition_New`). 

```{r}
## Set levels
sample_data(ps.prok.annotated)$Zoox_composition_New <- 
  factor(sample_data(ps.prok.annotated)$Zoox_composition_New, 
         levels = c("B", "B_dom_plus_C", "B_dom_plus_CD", "B_dom_plus_D", "BD", "CD"))

head(sample_data(ps.prok.annotated)$Zoox_composition_New, 10) # levels look good

## Test disease on colony when coring and control for symbiont composition
set.seed(10) # set seed for reproducibility
prok.da.zoox <- differentialTest(formula = ~ Zoox_composition_New, 
                                  phi.formula = ~ Zoox_composition_New,
                                  formula_null = ~ 1,
                                  phi.formula_null = ~ Zoox_composition_New,
                                  test = "Wald", boot = FALSE,
                                  data = ps.prok.annotated,
                                  fdr_cutoff = 0.05)

plot(prok.da.zoox, level = "SYMBOL")
prok.da.zoox$significant_taxa %>% length # Wow 2000 differentially abundant taxa
```

WOW! This resulted in 2000 differentially abundant taxa!!! That is a LOT, which may partially be beacuase some of the groups only had 1 sample in them, so I'll be curious how that worked. My goal next is to save all the data so I can report it in a supplementary Table.Definitiealy the most differentially abundant taxa in the BD group compared to B.

## Process output

Out of curiosity, I am also going to see how many taxa there are with a more stringent p-value cutoff. Save all the data first
```{r}
# write a function to extract the corncob results when there are 5 test levels (relavtive to the 6th baseline category)
extractmods5 <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2:6, 1], 
                        "Std.Error" = model$coefficients[2:6, 2], 
                       "t.value" = model$coefficients[2:6, 3], 
                        "p" = model$coefficients[2:6, 4],
                       "Comparison" = c("B_vs_B_dom_plus_C",
                                        "B_vs_B_dom_plus_CD",
                                        "B_vs_B_dom_plus_D",
                                        "B_vs_BD",
                                        "B_vs_CD"))
  return(result)
}

# save the output as a list using the above function
sig.models.zoox <- lapply(prok.da.zoox$significant_models, extractmods5)
names(sig.models.zoox) <- prok.da.zoox$significant_taxa

# save a table of annotations for just the significant genes
sig.fxnl.genes.zoox <- filter(unique_gene_ko_annotated, geneID %in% prok.da.zoox$significant_taxa)

# turn the output from a list to a dataframe for ease of saving and manipulating
# Additionally add further annotation information
sig.models.zoox.df <- plyr::ldply(sig.models.zoox, data.frame) %>% 
  mutate(Significant = ifelse(p < 0.05, TRUE, FALSE)) %>%
  left_join(sig.fxnl.genes.zoox, by = c(".id" = "geneID")) %>%
  left_join(prok_taxa, by = c(".id" = "contig"))

### Save these data so you don't have to re-run the model ###
write.table(sig.models.zoox.df, "../data/Corncob_prok_annotated_ZooxComposition.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

So this output is slightly confusing because for a given overall differentiallky abundant gene, it may be signficantly differentially abundant between B and CD, but not differentially abundant for other genes. I will count up different summary stats which may be more informative to report. 

How many overall differentially abundant genes?
```{r}
sig.models.zoox.df$.id %>% unique %>% length
```
Looks like 2000 genes. 

How many of these genes were differentially abundant in each comparison?
```{r}
sig.models.zoox.df %>% 
  group_by(Comparison) %>% 
  summarise(Significant_Gene_Count = sum(Significant == TRUE))

sample_data(ps.prok.annotated)$Zoox_composition_New %>% table
```

Looks like the most significantly differentially abundant genes were those in the B_vs_BD comparison (1,762 genes!) as well as the B_vs_CD comparison (291 genes). This makes sense because the CD and BD groupings have large groups of either C and D symbionts. I want to exercise caution however, because these goups were also significantly underreppresented in the dataset. There was only one of each CD and BD symbiont composition type. The most abundant grouping was the B only, which was 21 samples (about half the corals). The corals that had small proportions of either C, CD, or D included 5-7 corals, which is better representation. 

How many signfiicantly differentially abundant genes if we ignore any genes differentially abundant ONLY in the CD or BD groups?
```{r}
sig.NO_CD_or_BD <- sig.models.zoox.df %>% 
  filter(Comparison %in% c("B_vs_B_dom_plus_C", "B_vs_B_dom_plus_CD", "B_vs_B_dom_plus_D")) %>%
  filter(Significant == TRUE)

# how many genes?
sig.NO_CD_or_BD$.id %>% unique %>% length
```

WAAAAYYYY Fewer genes. Good to know. I will include all of them for reference, but this is a good reminder that sometimes considering groupings are important.

I'm also now curious, for each grouping, how many were enriched in the test group and how many were enriched in the baseline group. 
```{r}
sig.models.zoox.df %>% 
  filter(Significant == TRUE) %>%
  group_by(Comparison) %>% 
  summarise(Enriched_in_B = sum(Estimate < 0))


sig.models.zoox.df %>% 
  filter(Significant == TRUE) %>%
  group_by(Comparison) %>% 
  summarise(Enriched_in_test_group = sum(Estimate > 0))
```

