---
title: "DA_code_to_reproduce_Figure_3"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(dplyr)
library(tidyr)
library(readxl)

# For visualization
library(corncob)
library(DESeq2)
library(phyloseq)
library(randomcoloR)
```

# Read in data

## Load raw data 

```{r}
# Load table with count abundances from salmon ("numreads")
numreads <- read.table("../data/FLK_coassembly_salmon_quant_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
numreads <- numreads[, -1] # get rid of the first column cause it is just numbered lines

# Load the table with taoxnomic information from mmseqs2
taxa <- read.table("../data/FLK_coassembly_tax_mmseqs2_lca.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

# Load the table with functional information from eggnog mapper
fxn <- read_excel("../data/FLK_coassembly_fxn_eggnog.xlsx", skip = 2, na = "-")

# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID

# how many contigs are zero across all samples?
table(rowSums(numreads) == 0)

# remove them
numreads_nozero <- numreads[rowSums(numreads) != 0, ]
dim(numreads_nozero)

# any samples with no reads?
table(colSums(numreads_nozero) == 0)

# Fix the sample names
colnames(numreads_nozero) <- str_replace_all(colnames(numreads_nozero), pattern = "[X]", "")
colnames(numreads_nozero) <- str_split_i(colnames(numreads_nozero), "[_]", 1)
```

## Phyloseq object 
```{r}
idx <- match(colnames(numreads_nozero), rownames(metadata))
metadata_3 <- metadata[idx,]

NUMREADS = otu_table(numreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_3)

ps.count <- phyloseq(NUMREADS, META)
ps.count
```

## Annotated genes only
I would like to filter this phyloseq object to only include genes that are annotated. My rationale is that if I am to pull out any discussion or prediction about functional potential of the differentially abundant genes, then it would be useful to have that be what I do the analysis with. 
```{r}
numreads_nozero_for_fxn_filt <- numreads_nozero %>%
  mutate(query = rownames(numreads_nozero))

# Add count (number of reads) to the function data frame
fxn.numreads_nozero <- fxn %>%
  left_join(numreads_nozero_for_fxn_filt, by = "query")

fxn.numreads_nozero

# Select out only the reads again
numreads_nozero_annotated_only <- as.data.frame(fxn.numreads_nozero[,22:62])
rownames(numreads_nozero_annotated_only) <- fxn.numreads_nozero$query

# It has occurrred to me that some of these annotations are actually missing from the abundance table so are showing up as NAs. I need to remove those samples. 
table(is.na(rowSums(numreads_nozero_annotated_only)))
table(is.finite(rowSums(numreads_nozero_annotated_only)))

numreads_nozero_annotated_only_noNA <-
  numreads_nozero_annotated_only[is.finite(rowSums(numreads_nozero_annotated_only)), ]

numreads_nozero_annotated_only_noNA %>% dim

# Make the actual phyloseq object
idx <- match(colnames(numreads_nozero_annotated_only_noNA), metadata$CoralID)
metadata_annotated <- metadata[idx,]

NUMREADS = otu_table(numreads_nozero_annotated_only_noNA, taxa_are_rows = TRUE)
META = sample_data(metadata_annotated)

ps.count.annotated <- phyloseq(NUMREADS, META)
ps.count.annotated
```
This drastically reduced the number of genes I am working with, which makes sense. Ultimately, I am only doing annotated genes because I want to be able to draw biological conclusions from the data. 

## Save PS annotated object
```{r}
saveRDS(ps.count.annotated, "../data/FLK_coassembly_phyloseq_count.annotated.rds")
```

## Quick start - load ps object
```{r}
# Now what is it like to read this back in?
ps.count.annotated <- readRDS("../data/FLK_coassembly_phyloseq_count.annotated.rds")
```

# Differential abundance 
## deseq2 using the following code:
Set the levels of the factor of interest
```{r}
sample_data(ps.count.annotated)$Disease_on_colony_when_coring_binary <-
  factor(sample_data(ps.count.annotated)$Disease_on_colony_when_coring_binary, 
         levels = c("Unaffected", "Diseased"))

head(sample_data(ps.count.annotated)$Disease_on_colony_when_coring_binary, 10) # levels look good
```

These levels look good, and we are good to move on to DESeq2 analysis on the OFAV florida keys samples between completely unaffected to persistantly diseased coral colonies. 
```{r}
dds = phyloseq_to_deseq2(ps.count.annotated, ~ Disease_on_colony_when_coring_binary)
dds

#Keep all rows and apply geometric mean to replace zeros
cts <- DESeq2::counts(dds)
geoMeans <- apply(cts, 1, function(row) if (all(row == 0)) 0 else exp(mean(log(row[row != 0]))))
dds <- estimateSizeFactors(dds, geoMeans = geoMeans)

# run the actual deseq function - check results at 0.05 p adjusted value
dds <- DESeq(dds,test="Wald", fitType="parametric")
res = results(dds, cooksCutoff = FALSE)
alpha = 0.05
sigtab = res[which(res$padj < alpha), ] # 5,894 DA genes

# Add the functional annotations to the subset differentially abundant genes
sigtab.fxn <- as(sigtab, "data.frame") %>% 
  mutate(query = rownames(sigtab)) %>% 
  left_join(fxn, by = "query")

# Write these taxa out to a file to be saved for later
write.table(sigtab.fxn,"../data/DESeq2_coassembly_results_disease_0.05.txt", sep="\t", 
            col.names = TRUE, row.names = FALSE)

# Write the 0.01 cutoff genes
alpha01 = 0.01
sigtab01 = res[which(res$padj < alpha01), ] # 2,449 DA genes

# Add the functional annotations to the subset differentially abundant genes
sigtab01.fxn <- as(sigtab01, "data.frame") %>% 
  mutate(query = rownames(sigtab01)) %>% 
  left_join(fxn, by = "query")

# Write these taxa out to a file to be saved for later
write.table(sigtab01.fxn,"../data/DESeq2_coassembly_results_disease_0.01.txt", sep="\t", 
            col.names = TRUE, row.names = FALSE)

sigtab.fxn
sigtab01.fxn
```

## Corncob using the following code:
```{r, eval=FALSE}
set.seed(10) # set seed for reproducibility
corncobDA.dis <- differentialTest(formula = ~ Disease_on_colony_when_coring_binary, 
                                   phi.formula = ~ Disease_on_colony_when_coring_binary,
                                   formula_null = ~ 1,
                                   phi.formula_null = ~ Disease_on_colony_when_coring_binary,
                                   test = "Wald", boot = FALSE,
                                   data = ps.count.annotated,
                                   fdr_cutoff = 0.05)

# Save the corncob phyloseq object so you can manipulate it in R
saveRDS(corncobDA.dis, "corncob_output_disease_at_sampling.rds")

## Try controlling for the differences in the symbiont community
set.seed(10) # set seed for reproducibility
corncobDA.dis.symb <- differentialTest(formula = ~ Disease_on_colony_when_coring_binary + Zoox_background, 
                                  phi.formula = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  formula_null = ~ Zoox_background,
                                  phi.formula_null = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  test = "Wald", boot = FALSE,
                                  data = ps.count.annotated,
                                  fdr_cutoff = 0.05)

# Save the corncob phyloseq object so you can manipulate it in R
saveRDS(corncobDA.dis.symb, "corncob_output_disease_at_sampling_zoox.rds")
```
I would like to see how the number of differentially abudnant genes differs between these two runs, but I feel like I should do the second one, because it accounts for background zoox, which significantly explained variation in the microbiome, so I feel it is best to account for both things. 

### disease at sampling only
Then, I manipulated and saved the corncob output with this code:
```{r}
# read in RDS file
corncobDA.dis <- readRDS("../data/corncob_output_disease_at_sampling.rds")

plot(corncobDA.dis)

# How many significant taxa are there?
corncobDA.dis$significant_taxa %>% length
```


```{r}

```

### diseae at sampling while controlling for differential abundance due to background zooxanthellae
```{r}
# read in RDS file
corncobDA.dis.symb <- readRDS("../data/corncob_output_disease_at_sampling_zoox.rds")

plot(corncobDA.dis.symb)

# How many significant taxa are there?
corncobDA.dis.symb$significant_taxa %>% length
```
There are so many taxa that are differentially abundant. To get the ones that change most, it may be worthwhile to identify the taxa with the greatest change between unaffected and diseased taxa, or those with the greatest coefficient (positive or negative).
```{r}
# I want to extract the important information from this into a table...
# check out results structure from one of the models
corncobDA.dis.symb$significant_models[[1]]$coefficients

# write a function to extract the corncob results
extractmods2 <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                        "Std.Error" = model$coefficients[2, 2], 
                        "p" = model$coefficients[2, 4])
  return(result)
}

# save the output as a list using the above function
corncob.dis.symb.models <- lapply(corncobDA.dis.symb$significant_models, extractmods2)
names(corncob.dis.symb.models) <- corncobDA.dis.symb$significant_taxa

# save a table of annotations for just the significant genes
sig.fxnl.genes.dis.symb <- filter(fxn, query %in% corncobDA.dis.symb$significant_taxa)

# turn the output from a list to a dataframe for ease of saving and manipulating
corncob.dis.symb.models.df <- plyr::ldply(corncob.dis.symb.models, data.frame) %>% 
  left_join(sig.fxnl.genes.dis.symb, by = c(".id" = "query"))

### Save these data so you don't have to re-run the model ###
write.table(corncob.dis.symb.models.df, "../data/Corncob_coassembly_results_dis.symb_0.05.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

## Corncob disease and symbiont
Read in the corncob data. I have a couple of questions. I want to see what the greatest 
```{r}
corn.dis.symb <- read.table("../data/Corncob_coassembly_results_dis.symb_0.05.txt", header = TRUE, sep = "\t")

corn.dis.symb$Estimate %>% summary

table(corn.dis.symb$Estimate > 0) 
table(corn.dis.symb$Estimate < -1) 
table(corn.dis.symb$Estimate < -3) 
table(corn.dis.symb$Estimate > 3) 
table(corn.dis.symb$Estimate < -3.5) 
table(corn.dis.symb$Estimate > 5) 
```
## Figure for MS

I imagine producing a table of all the differentially abundant genes, but for a general figure, I am interested in:
- What functions are these differentially abundant genes performing?
- What taxonomic groups are these differentially abundant genes part of?

Genes enriched in diseased - what taxonomic groups are there?
Genes enriched in unaffected - what taxonomic groups are there?

We conducted differential abundance tests in apparently healthy tissue between unaffected and diseased coral colonies while controlling for the effect of background symbiodiniaceae on differential abundance, and identified 1,605 differentially abundant genes (supplementary table). This included 303 enriched in diseased tissue, and 1302 enriched in unaffected coral tissue. These genes were from varied taxonomic groups, including x, y, and z. 

### Taxonomic info
```{r}
# What taxa ID is assigned to genes that were enriched in disease
corn.dis.symb_tax <- corn.dis.symb %>%
  left_join(taxa, by = c(".id" = "contig")) %>%
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE)

# How many of the total differentially abundant genes were actually annotated as bacteria?
# disease-enriched - 168 of 303 
dis.prok <- filter(corn.dis.symb_tax, Estimate > 0) %>% 
  filter(Domain %in% c("d_Bacteria", "d_Archaea")) %>%
  mutate(Phylum = case_when(
    is.na(Phylum) ~ paste0(Domain),
    TRUE ~ Phylum
    )) %>%
  mutate(Class = case_when(
    is.na(Class) ~ paste0(Phylum),
    TRUE ~ Class
    )) %>%
  mutate(Order = case_when(
    is.na(Order) ~ paste0(Class),
    TRUE ~ Order
    )) %>%
  mutate(Family = case_when(
    is.na(Family) ~ paste0(Order),
    TRUE ~ Family
    )) %>%
  mutate(Genus = case_when(
    is.na(Genus) ~ paste0(Family),
    TRUE ~ Genus
    )) %>%
  mutate(Species = case_when(
    is.na(Species) ~ paste0(Genus),
    TRUE ~ Species
    ))


# unaffected-enriched - 1,129 of 1,302
unaff.prok <- filter(corn.dis.symb_tax, Estimate < 0) %>% 
  filter(Domain %in% c("d_Bacteria", "d_Archaea")) %>%
  mutate(Phylum = case_when(
    is.na(Phylum) ~ paste0(Domain),
    TRUE ~ Phylum
    )) %>%
  mutate(Class = case_when(
    is.na(Class) ~ paste0(Phylum),
    TRUE ~ Class
    )) %>%
  mutate(Order = case_when(
    is.na(Order) ~ paste0(Class),
    TRUE ~ Order
    )) %>%
  mutate(Family = case_when(
    is.na(Family) ~ paste0(Order),
    TRUE ~ Family
    )) %>%
  mutate(Genus = case_when(
    is.na(Genus) ~ paste0(Family),
    TRUE ~ Genus
    )) %>%
  mutate(Species = case_when(
    is.na(Species) ~ paste0(Genus),
    TRUE ~ Species
    ))

1129/1303 # 86.6%

# Can I just make a stacked bar chart with two columns (one disease and one unaffected)?
dis.unaff.prok <- rbind(dis.prok, unaff.prok) %>%
  mutate(corncob_result = ifelse(Estimate > 0, "Disease_assoc", "Unaffected_assoc"))

x1 <- distinctColorPalette(50)

# Make a plot
ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Phylum)) +
  geom_bar() +
  scale_fill_manual(values = x1)

ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Phylum)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = x1)

ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Class)) +
  geom_bar() +
  scale_fill_manual(values = x1)

ggplot(dis.unaff.prok, aes(x = corncob_result, fill = Class)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = x1)
```

