---
title: "Benefitial_gene_code_to_reproduce_Figures_4-6"
author: "Cynthia Becker"
date: "2024-08-01"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.path = "../figures/fig-")
setwd("~/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/")
```

# Setup
## Install necessary packages

```{r packages, message=FALSE, warning=FALSE, results='hide'}
# For data wrangling
library(dplyr)
library(readxl)
library(tidyr)

# For visualization
library(ggplot2)
```

# Load data

## metabolic data
```{r}
# Load METABOLIC functional information
HMM_hits <- read_excel("../data/FLK_coassembly_METABOLIC_result.xlsx", sheet = "HMMHitNum")
fxns_hits <- read_excel("../data/FLK_coassembly_METABOLIC_result.xlsx", sheet = "FunctionHit")
KEGG_hits <- read_excel("../data/FLK_coassembly_METABOLIC_result.xlsx", sheet = "KEGGModuleHit")
KEGG_step_hits <- read_excel("../data/FLK_coassembly_METABOLIC_result.xlsx", sheet = "KEGGModuleStepHit")

# Load table with count abundances from salmon ("numreads")
numreads <- read.table("../data/FLK_coassembly_salmon_quant_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
numreads <- numreads[, -1] # get rid of the first column cause it is just numbered lines

# Load the table with taoxnomic information from mmseqs2
taxa <- read.table("../data/FLK_coassembly_tax_mmseqs2_lca.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

# Load the table with functional information from eggnog mapper
fxn <- read_excel("../data/FLK_coassembly_fxn_eggnog.xlsx", skip = 2, na = "-")

# Load metadata
metadata <- as.data.frame(read_excel("../data/metadata_SP1_v9.4.xlsx", na = "NA"))
rownames(metadata) <- metadata$CoralID
```

## Count phyloseq object that is centered log-ratio transformation.

Create a phyloseq object with the count data. Remove any genes with zero abundances. Also make a phyloseq object with only the annotated genes. 
```{r}
# remove contigs with zero abundacne
numreads_nozero <- numreads[rowSums(numreads) != 0, ]
dim(numreads_nozero)

# any samples with no reads?
table(colSums(numreads_nozero) == 0)

# Fix the sample names
colnames(numreads_nozero) <- str_replace_all(colnames(numreads_nozero), pattern = "[X]", "")
colnames(numreads_nozero) <- str_split_i(colnames(numreads_nozero), "[_]", 1)
```

## phyloseq object - count data
```{r}
idx <- match(colnames(numreads_nozero), rownames(metadata))
metadata_3 <- metadata[idx,]

NUMREADS = otu_table(numreads_nozero, taxa_are_rows = TRUE)
META = sample_data(metadata_3)

ps.count <- phyloseq(NUMREADS, META)
ps.count
```

## clr-transf phyloseq object

I need to do a centered log-ratio transformation, which addresses the compositional nature of the data before I compare between groups.
```{r}
ps.count_clr <- microbiome::transform(ps.count, 'clr')
ps.count_clr
```

## Nitrogen

I am interested in analyzing which nitrogen cycling genes change with Fate.

### Organize data
Get the data organized into a form for graphing and statistical analyses. 
```{r}
n_hits <- fxns_hits %>%
  filter(Category == "Nitrogen cycling") %>%
  filter(FLK_OFAV_MG_pred.Function.presence == "Present")

n_hits

n_contig_hits <- HMM_hits %>%
  filter(Category == "Nitrogen cycling") %>%
  filter(FLK_OFAV_MG_pred.Hmm.presence == "Present") %>%
  mutate(FLK_OFAV_MG_pred.Hits = str_split(FLK_OFAV_MG_pred.Hits, ",")) %>%
  unnest_wider(FLK_OFAV_MG_pred.Hits, names_sep = "_") %>%
  gather(key = "arbitrary_hit_number", value = "contig", FLK_OFAV_MG_pred.Hits_1:FLK_OFAV_MG_pred.Hits_14) %>%
  filter(!is.na(contig))

# Which contigs contain nitrogen cycling genes?
n_contig_hits$contig

n_cycling_contigs <- n_contig_hits$contig

# make a ps object with only those n cycling contigs
ps.count.clr_n_cycle <- prune_taxa(n_cycling_contigs, ps.count_clr)

n_cycle_forgraph_clr <- as.data.frame(otu_table(ps.count.clr_n_cycle)) %>%
  mutate(contig = rownames(.)) %>%
  left_join(n_contig_hits, by = "contig", relationship = "many-to-many") %>% # need to set many to many because the k127_62437_2 contig has two annotations
  gather(key = "CoralID", value = "clr", 1:41) %>% 
  left_join(metadata, by = "CoralID") %>%   # add the metadata
  left_join(taxa, by = "contig") %>% # add the taxonomy
  mutate(Fate_after_SP1 = factor(Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", 
                                            "Will be diseased", "Diseased"))) %>%
  mutate(Disease_on_colony_when_coring_binary = factor(Disease_on_colony_when_coring_binary, 
                                 levels = c("Unaffected", "Diseased")))

n_cycle_forgraph_clr
```

### Wilcox test
Now do the statistical test on the group, by disease at sampling (Disease_on_colony_when_coring_binary). I will run a wilcoxon test, then graph with all the genes. 
```{r}
# Get data for one contig each, getting rid of the duplicate contigs
n_cycle_clr_forstat <- as.data.frame(otu_table(ps.count.clr_n_cycle)) %>%
  mutate(contig = rownames(.)) %>%
  gather(key = "CoralID", value = "clr", 1:41) %>% 
  left_join(metadata, by = "CoralID") %>%   # add the metadata
  left_join(taxa, by = "contig") %>% # add the taxonomy
  mutate(Disease_on_colony_when_coring_binary = factor(Disease_on_colony_when_coring_binary, 
                                 levels = c("Unaffected", "Diseased")))

summary(plyr::count(n_cycle_clr_forstat$contig)$freq) # verify you have the same number of samples for each contig - Works!

set.seed(100)

# perform wilcoxon test for each contig between the subset fate categories
wilcoxon_results_Fate <- n_cycle_clr_forstat %>% 
  group_by(contig) %>%
  dplyr::summarize(
    p_value = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$p.value,
    statistic = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$statistic,
    method = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$method
  )

# which genes are significantly different?
results0.05_N <- dplyr::filter(wilcoxon_results_Fate, p_value < 0.05) %>%
  left_join(n_contig_hits, by = "contig")
results0.05_N 

results0.01_N <- dplyr::filter(wilcoxon_results_Fate, p_value < 0.01) %>%
  left_join(n_contig_hits, by = "contig")
results0.01_N

# Try a bonferroni correction -  tests 54
ntests_N <- n_cycle_clr_forstat$contig %>% unique %>% length

N_resultsBC <- dplyr::filter(wilcoxon_results_Fate, p_value < 0.05/ntests_N) %>%
  left_join(n_contig_hits, by = "contig")
N_resultsBC
```

### Main figure
Make a graph of the results at the 0.01 level. 
```{r}
safe_colorblind_palette <- c("#000000", "#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

palette_for_mags <- c("#B3E2CD", "#FDCDAC", "#CBD5E8", "#F4CAE4", "#E6F5C9", "#FFF2AE", "#F1E2CC", "#CCCCCC")

n_cycle_forgraph_clr_sig01 <- n_cycle_forgraph_clr %>%
  filter(contig %in% results0.01_N$contig)

ggplot(n_cycle_forgraph_clr_sig01, aes(x = Disease_on_colony_when_coring_binary, y = clr)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position = position_jitter(width = .1, height = 0), 
              aes(color = complete_classification, shape = Disease_on_colony_when_coring_binary), size = 4) +
  ylab("clr-transformed abundance") +
  facet_wrap(. ~ Function + contig, scales = "free_y", ncol = 7) +
  theme(axis.text.x = element_text(angle = 45, hjust= 1, vjust = 1)) +
  scale_color_manual(values = c("#000000", "#882255","#AA4499", "#999933", "#DDCC77", "#117733", "#332288", "#6699CC", "#44AA99", "#88CCEE",  "#CC6677", "#661100","#888888")) +
  labs(shape = "Disease at sampling", color = "Taxonomic annotation of contig", x = "Fate after sampling")

#ggsave("../figures/box_whisker_N_cycling_fate.pdf", width = 22, height = 3)

ggplot(n_cycle_forgraph_clr_sig01, aes(x = Disease_on_colony_when_coring_binary, y = clr)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position = position_jitter(width = .1, height = 0), 
              aes(color = complete_classification, shape = Disease_on_colony_when_coring_binary), size = 4) +
  ylab("clr-transformed abundance") +
  facet_wrap(. ~ Function + contig, scales = "free_y", ncol = 7) +
  theme(axis.text.x = element_text(angle = 45, hjust= 1, vjust = 1)) +
  scale_color_manual(values = c("#000000", "#882255","#AA4499", "#999933", "#DDCC77", "#117733", "#332288", "#6699CC", "#44AA99", "#88CCEE",  "#CC6677", "#661100","#888888")) +
  labs(shape = "Disease at sampling", color = "Taxonomic annotation of contig", x = "Fate after sampling") +
  theme(legend.position = "none")
```

## Sulfur

### Organize data
```{r}
s_hits <- fxns_hits %>%
  filter(Category == "Sulfur cycling") %>%
  filter(FLK_OFAV_MG_pred.Function.presence == "Present")

s_hits

s_contig_hits_summary <- HMM_hits %>%
  filter(Category %in% c("Sulfur cycling", "Sulfur cycling enzymes (detailed)")) %>%
  filter(FLK_OFAV_MG_pred.Hmm.presence == "Present")

s_contig_hits_summary

s_contig_hits_summary %>% 
  dplyr::select(Gene.abbreviation, Gene.name) %>%
  mutate(combo = paste0(Gene.abbreviation, " = ", Gene.name)) %>%
  dplyr::select(combo) %>%
  unique %>%
  as.data.frame

# I also found some DMSP degrading activity in the eggnog results, cause there werent any in the metabolic database
s_eggnog_hits_DMSP <- fxn %>% 
  filter(Description %in% c("Mediates cleavage of dimethlysulfonioproprionate (DMSP) into dimethyl sulfide (DMS) and acrylate. DMS is the principal form by which sulfur is transported from oceans to the atmosphere and is a key component of the ocean sulfur cycle", "Dimethlysulfonioproprionate lyase")) %>%
  left_join(taxa, by = c("query" = "contig"))

s_eggnog_hits_DMSP

s_contig_hits <- HMM_hits %>%
  filter(Category %in% c("Sulfur cycling", "Sulfur cycling enzymes (detailed)")) %>%
  filter(FLK_OFAV_MG_pred.Hmm.presence == "Present") %>%
  mutate(FLK_OFAV_MG_pred.Hits = str_split(FLK_OFAV_MG_pred.Hits, ",")) %>%
  unnest_wider(FLK_OFAV_MG_pred.Hits, names_sep = "_") %>%
  gather(key = "arbitrary_hit_number", value = "contig", FLK_OFAV_MG_pred.Hits_1:FLK_OFAV_MG_pred.Hits_50) %>%
  filter(!is.na(contig))

s_contig_hits

# Which contigs contain sulfur cycling genes? Combine the eggnog sulfur contigs and the metabolic sulfur contigs
s_contig_hits$contig
s_eggnog_hits_DMSP$query

s_cycling_contigs <- c(s_contig_hits$contig, s_eggnog_hits_DMSP$query)

# make a ps object with only those n cycling contigs
ps.count.clr_s_cycle <- prune_taxa(s_cycling_contigs, ps.count_clr)

s_cycle_forgraph_clr <- as.data.frame(otu_table(ps.count.clr_s_cycle)) %>%
  mutate(contig = rownames(.)) %>%
  left_join(s_contig_hits, by = "contig", relationship = "many-to-many") %>% # need to set many to many because multiple contigs have multiple annotations
  left_join(s_eggnog_hits_DMSP, by = c("contig" = "query")) %>%
  gather(key = "CoralID", value = "clr", 1:41) %>% 
  left_join(metadata, by = "CoralID") %>%   # add the metadata
  left_join(taxa, by = "contig") %>% # add the taxonomy
  mutate(Fate_after_SP1 = factor(Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", 
                                            "Will be diseased", "Diseased"))) %>%
  mutate(Disease_on_colony_when_coring_binary = factor(Disease_on_colony_when_coring_binary, 
                                 levels = c("Unaffected", "Diseased")))

s_cycle_forgraph_clr
```

### Wilcox test

Wow, there are a lot of organosulfur protein genes as well as inorganic sulfur. Now do the statistical test on the group, by fate...So I need to subset the data by just the unaffected and diseased fates, run the wilcoxon test, then graph with all the genes. 
```{r}
s_cycle_clr_forstat <- as.data.frame(otu_table(ps.count.clr_s_cycle)) %>%
  mutate(contig = rownames(.)) %>%
  gather(key = "CoralID", value = "clr", 1:41) %>% 
  left_join(metadata, by = "CoralID") %>%   # add the metadata
  left_join(taxa, by = "contig") %>% # add the taxonomy
  mutate(Disease_on_colony_when_coring_binary = factor(Disease_on_colony_when_coring_binary, 
                                 levels = c("Unaffected", "Diseased")))

summary(plyr::count(s_cycle_clr_forstat$contig)$freq) # verify you have the same number of samples for each contig

set.seed(100)

# perform wilcoxon test for each contig between the subset fate categories
S_wilcoxon_results_Fate <- s_cycle_clr_forstat %>% 
  group_by(contig) %>%
  dplyr::summarize(
    p_value = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$p.value,
    statistic = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$statistic,
    method = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$method
  )

# which genes are significantly different?
S_results0.01 <- dplyr::filter(S_wilcoxon_results_Fate, p_value < 0.01) %>%
  left_join(s_contig_hits, by = "contig") %>%
  left_join(s_eggnog_hits_DMSP, by = c("contig" = "query"))
S_results0.01

S_results0.01$contig %>% unique %>% length

# which genes are significantly different?
S_results0.001 <- dplyr::filter(S_wilcoxon_results_Fate, p_value < 0.001) %>%
  left_join(s_contig_hits, by = "contig") %>%
  left_join(s_eggnog_hits_DMSP, by = c("contig" = "query"))
S_results0.001

S_results0.001$contig %>% unique %>% length

# Try a bonferroni correction - 261 tests
ntests_S <- s_cycle_clr_forstat$contig %>% unique %>% length

S_resultsBC <- dplyr::filter(S_wilcoxon_results_Fate, p_value < 0.05/ntests_S) %>%
  left_join(s_contig_hits, by = "contig") %>%
  left_join(s_eggnog_hits_DMSP, by = c("contig" = "query"))
S_resultsBC
```

### main figure
```{r}
s_cycle_forgraph_clr_sig01 <- s_cycle_forgraph_clr %>%
  filter(contig %in% S_results0.01$contig) %>%
  mutate(gene_shortname = paste0(Gene.abbreviation, PFAMs)) %>%
  mutate(gene_shortname = str_replace_all(gene_shortname, "NA", ""))

safe_colorblind_palette <- c("#000000", "#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

palette_for_mags <- c("#B3E2CD", "#FDCDAC", "#CBD5E8", "#F4CAE4", "#E6F5C9", "#FFF2AE", "#F1E2CC", "#CCCCCC")

s_cycle_forgraph_clr_sig01 <- s_cycle_forgraph_clr_sig01 %>%
  mutate(contig = factor(contig,
                         levels = c("k127_1087367_3", "k127_1157184_1", "k127_881950_25", "k127_129157_4",
                                    "k127_199314_1", "k127_191079_2", "k127_3369475_10", "k127_1050565_91",
                                    "k127_3384539_4", "k127_959642_3", "k127_2757884_1"))) %>%
  mutate(gene_shortname = factor(gene_shortname,
                         levels = c("sdo", "ahcY", "metH", "TPP_enzyme_C",
                                    "metK", "dcyD", "luxS", "megL")))

ggplot(s_cycle_forgraph_clr_sig01, aes(x = Disease_on_colony_when_coring_binary, y = clr)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position = position_jitter(width = .1, height = 0), 
              aes(color = complete_classification.y, shape = Fate_after_SP1), size = 4) +
  ylab("clr-transformed abundance") +
  facet_wrap(. ~ gene_shortname + contig, scales = "free_y", ncol = 4) +
  theme(axis.text.x = element_text(angle = 45, hjust= 1, vjust = 1)) +
  #scale_color_manual(values = c("#888888", "#000000", "#88CCEE","#CC6677", "#332288", "#AA4499", 
     #                        "#44AA99", "#999933", "#882255", "#661100", "#888888")) +
  labs(shape = "Disease at sampling", color = "Taxonomic annotation of contig", x = "Fate after sampling")

#ggsave("../figures/box_whisker_S_cycling_fate.pdf", width = 15, height = 7)

ggplot(s_cycle_forgraph_clr_sig01, aes(x = Disease_on_colony_when_coring_binary, y = clr)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position = position_jitter(width = .1, height = 0), 
              aes(color = complete_classification.y, shape = Fate_after_SP1), size = 4) +
  ylab("clr-transformed abundance") +
  facet_wrap(. ~ gene_shortname + contig, scales = "free_y", ncol = 4) +
  theme(axis.text.x = element_text(angle = 45, hjust= 1, vjust = 1)) +
  #scale_color_manual(values = c("#888888", "#000000", "#88CCEE","#CC6677", "#332288", "#AA4499", 
     #                        "#44AA99", "#999933", "#882255", "#661100", "#888888")) +
  labs(shape = "Disease at sampling", color = "Taxonomic annotation of contig", x = "Fate after sampling") +
  theme(legend.position = "none")
```

## Vitamin and cofactor biosynthesis genes
I want to see how vitamin biosynthesis genes are changing with coral fate. I would hypothesize that these are involved in coral health. This is more complicated to find, because the METABOLIC output that identifies the present/absent modules is not combined with a contig ID result. But the KO ID results are in the eggnog output. So I will combine the METABOLIC and the eggnog output. 

### Organize data
```{r}
vitamin_hits_broad <- KEGG_hits %>%
  filter(Module.Category == "Cofactor and vitamin metabolism") %>%
  filter(FLK_OFAV_MG_pred.Module.presence == "Present")

vitamin_hits_broad

# Need a list of the KO ID's so that I can cross-reference it with the eggnog output
vitamin_hits <- KEGG_step_hits %>%
  filter(Module.Category == "Cofactor and vitamin metabolism") %>%
  filter(FLK_OFAV_MG_pred.Module.step.presence == "Present") %>%
  mutate(KO.id.test = str_split(KO.id, ",")) %>% # make mutliple KOs different columns 
  unnest_wider(KO.id.test, names_sep = "_") %>%
  gather(key = "KO_arbitrary_id", value = "KO", KO.id.test_1:KO.id.test_6) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_split(KO, "[+]")) %>% # make mutliple KOs different columns that have plusses
  unnest_wider(KO, names_sep = "_") %>%
  gather(key = "KO_arbitrary_id2", value = "KO", KO_1:KO_3) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_split(KO, "[-]")) %>% # make mutliple KOs different columns that have minuses
  unnest_wider(KO, names_sep = "_") %>%
  gather(key = "KO_arbitrary_id3", value = "KO", KO_1:KO_2) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_replace_all(KO, "[)]", "")) %>% # remove the random parentheses
  mutate(KO = str_replace_all(KO, "[(]", ""))

vitamin_hits

# Now I need a list of the contigs that go with the different KO ids from the eggnog functional results
vitamin_eggnog_ko_contigs <- fxn %>%
  dplyr::select(query, KEGG_ko, KEGG_Module) %>%
  mutate(KEGG_ko = str_split(KEGG_ko, ",")) %>% # make mutliple KOs different columns 
  unnest_wider(KEGG_ko, names_sep = "_") %>%
  gather(key = "KO_random_number", value = "KO", KEGG_ko_1:KEGG_ko_15) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_replace_all(KO, "ko:", "")) %>% # remove the random "ko:"
  dplyr::select(!KO_random_number)

vitamin_eggnog_ko_contigs

# Add the contig ID information to the KEGG module information
vitamin_ko_contig_hits <- vitamin_hits %>%
  left_join(vitamin_eggnog_ko_contigs, by = "KO", relationship = "many-to-many") # need many to many because there are multiple x and y matches

vitamin_ko_contig_hits

# Which contigs contain vitamin biosynthesis genes?
vitamin_ko_contig_hits$query

vitamin_cycling_contigs <- vitamin_ko_contig_hits$query %>% unique

# make a ps object with only those vitamin biosynthesis contigs
ps.count.clr_vitamin <- prune_taxa(vitamin_cycling_contigs, ps.count_clr)

vitamin_forgraph_clr <- as.data.frame(otu_table(ps.count.clr_vitamin)) %>%
  mutate(contig = rownames(.)) %>%
  left_join(vitamin_ko_contig_hits, by = c("contig" = "query"), relationship = "many-to-many") %>% # need to set many to many because multiple contigs have multiple annotations
  gather(key = "CoralID", value = "clr", 1:41) %>% 
  left_join(metadata, by = "CoralID") %>%   # add the metadata
  left_join(taxa, by = "contig") %>% # add the taxonomy
  mutate(Fate_after_SP1 = factor(Fate_after_SP1, 
                                 levels = c("Recovered", "Unaffected", "Will be diseased", "Diseased"))) %>%
  mutate(Disease_on_colony_when_coring_binary = factor(Disease_on_colony_when_coring_binary, 
                                 levels = c("Unaffected", "Diseased")))

vitamin_forgraph_clr
```

### Wilcox test

So now I want to compare the results between the different genes, but I don't want to count contigs twice. Instead, if a contig has multiple annotations, I need to make sure my statistical tests are only incorporating one set of the contigs. 
```{r}
vitamin_clr_forstat <- as.data.frame(otu_table(ps.count.clr_vitamin)) %>%
  mutate(contig = rownames(.)) %>%
  gather(key = "CoralID", value = "clr", 1:41) %>% 
  left_join(metadata, by = "CoralID") %>%   # add the metadata
  left_join(taxa, by = "contig") %>% # add the taxonomy
  mutate(Disease_on_colony_when_coring_binary = factor(Disease_on_colony_when_coring_binary, 
                                 levels = c("Unaffected", "Diseased")))

summary(plyr::count(vitamin_clr_forstat$contig)$freq) # verify you have the same number of samples for each contig

set.seed(100)

# perform wilcoxon test for each contig between the subset fate categories
vitamin_wilcoxon_results_Fate <- vitamin_clr_forstat %>% 
  group_by(contig) %>%
  dplyr::summarize(
    p_value = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$p.value,
    statistic = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$statistic,
    method = wilcox.test(clr ~ Disease_on_colony_when_coring_binary)$method
  )

# which genes are significantly different?
dplyr::filter(vitamin_wilcoxon_results_Fate, p_value < 0.01) %>% nrow #22 genes

vit_results0.01 <- dplyr::filter(vitamin_wilcoxon_results_Fate, p_value < 0.01) %>%
  left_join(vitamin_ko_contig_hits, by = c("contig" = "query"), relationship = "many-to-many")

vit_results0.01

vit_results0.01$contig %>% unique %>% length

# Try a bonferroni correction - 261 tests
ntests_V <- vitamin_clr_forstat$contig %>% unique %>% length

V_resultsBC <- dplyr::filter(vitamin_wilcoxon_results_Fate, p_value < 0.05/ntests_V) %>%
  left_join(vitamin_ko_contig_hits, by = c("contig" = "query"), relationship = "many-to-many")

V_resultsBC
```


You know what would be better? z-score standardize each gene across the samples so I can see which one is higher/lower in each sample.

```{r}
# Step 1: select the data 
data <- prune_taxa(unique(vit_results0.01$contig), ps.count_clr)

data_wide <- as.data.frame(otu_table(data)) %>%
  t() %>%
  as.data.frame()

# Step 3: Standardize the data using z-score standardization for each gene across samples
sample_ids <- rownames(data_wide)
gene_data_scaled <- as.data.frame(scale(data_wide))
gene_data_scaled <- cbind(CoralID = sample_ids, gene_data_scaled)

# Step 4: Reshape back to long format
gene_data_scaled_long <- gene_data_scaled %>%
  gather(key = "contig", value = "z_score_clr", k127_20672_2:k127_3398162_2) %>%
  left_join(vitamin_ko_contig_hits, by = c("contig" = "query"), relationship = "many-to-many") %>%
  left_join(metadata, by = "CoralID") %>%   # add the metadata
  left_join(taxa, by = "contig") %>% # add the taxonomy
  mutate(Disease_on_colony_when_coring_binary = factor(Disease_on_colony_when_coring_binary, 
                                 levels = c("Unaffected", "Diseased"))) %>%
  mutate(module_broad = str_split_i(Module, ",", 1))
```

### main figure
Heatmap figure
```{r}
safe_colorblind_palette <- c("#000000", "#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

palette_for_mags <- c("#B3E2CD", "#FDCDAC", "#CBD5E8", "#F4CAE4", "#E6F5C9", "#FFF2AE", "#F1E2CC", "#CCCCCC")

# Step 5: Make a tiled graph that should be scaled for each gene
ggplot(gene_data_scaled_long, aes(x = CoralID, y = contig)) +
  geom_tile(aes(fill = z_score_clr)) +
  facet_grid(module_broad ~ Disease_on_colony_when_coring_binary, scales = "free", space = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust= 1, vjust = 1)) +
  scale_fill_gradientn(colors = hcl.colors(256, "Spectral", rev = TRUE)) +
  labs(fill = "Z-score of clr-based abundance", x = "Coral ID", y = "Contig in co-assembly")

# ggsave("../figures/heatmap_vitamin_fate.pdf", width = 11, height = 8)
```

### Supplementary figure
Supplementary box and whisker plot
```{r}
vitamin_forgraph_clr_sig <- vitamin_forgraph_clr %>%
  filter(contig %in% vit_results0.01$contig) 

# Make a pretty box and whisker plot graph 
ggplot(vitamin_forgraph_clr_sig, aes(x = Disease_on_colony_when_coring_binary, y = clr)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position = position_jitter(width = .1, height = 0), 
              aes(color = complete_classification, shape = Disease_on_colony_when_coring_binary), size = 4) +
  ylab("clr-transformed abundance") +
  facet_wrap(. ~ contig, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust= 1, vjust = 1)) +
  labs(shape = "Fate after sampling", 
       color = "Taxonomic annotation of contig", 
       x = "Disease at sampling") 

#ggsave("../figures/box_whisker_vitamin_fate.png", width = 16, height = 8)

ggplot(vitamin_forgraph_clr_sig, aes(x = Disease_on_colony_when_coring_binary, y = clr)) +
  geom_boxplot(lwd = 1, outlier.shape = NA) +
  stat_boxplot(geom = "errorbar", width = 0.2, lwd = 1) + 
  geom_jitter(position = position_jitter(width = .1, height = 0), 
              aes(color = complete_classification, shape = Disease_on_colony_when_coring_binary), size = 4) +
  ylab("clr-transformed abundance") +
  facet_wrap(. ~ contig, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust= 1, vjust = 1)) +
  labs(shape = "Fate after sampling", 
       color = "Taxonomic annotation of contig", 
       x = "Disease at sampling") +
  theme(legend.position = "none")
```

# Corncob on METABOLIC genes

I want to use corncob because corncob can control for different variables, like zoox_background. I am interested in nitrogen-cycling genes, sulfur-cycling genes, and vitamin biosynthesis genes as these are genes predicted to have a beneficial role in the coral holobiont. I hypothesize that these genes will be enriched in apparently healthy tissue from unaffected colonies relative to apparently healthy tissue from diseased colonies. Given the known role of background zooxanthellae in influencing microbial functional community variability, I want to control for the effect of background zooxanthellae when conducting my statistical test. 

## Nitrogen contig list
```{r}
n_hits <- fxns_hits %>%
  filter(Category == "Nitrogen cycling") %>%
  filter(FLK_OFAV_MG_pred.Function.presence == "Present")

n_hits

n_contig_hits <- HMM_hits %>%
  filter(Category == "Nitrogen cycling") %>%
  filter(FLK_OFAV_MG_pred.Hmm.presence == "Present") %>%
  mutate(FLK_OFAV_MG_pred.Hits = str_split(FLK_OFAV_MG_pred.Hits, ",")) %>%
  unnest_wider(FLK_OFAV_MG_pred.Hits, names_sep = "_") %>%
  gather(key = "arbitrary_hit_number", value = "contig", FLK_OFAV_MG_pred.Hits_1:FLK_OFAV_MG_pred.Hits_14) %>%
  filter(!is.na(contig))

# Which contigs contain nitrogen cycling genes?
n_contig_hits$contig

n_cycling_contigs <- n_contig_hits$contig

plyr::count(n_cycling_contigs) %>% nrow # number of unique genes
```

## Sulfur contig list
```{r}
# inorganic sulfur cycling functions
s_hits <- fxns_hits %>%
  filter(Category == "Sulfur cycling") %>%
  filter(FLK_OFAV_MG_pred.Function.presence == "Present")

s_hits

# inorganic and organic sulfur cycling
s_contig_hits_summary <- HMM_hits %>%
  filter(Category %in% c("Sulfur cycling", "Sulfur cycling enzymes (detailed)")) %>%
  filter(FLK_OFAV_MG_pred.Hmm.presence == "Present")

s_contig_hits_summary

# Check out the gene names
s_contig_hits_summary %>% 
  dplyr::select(Gene.abbreviation, Gene.name) %>%
  mutate(combo = paste0(Gene.abbreviation, " = ", Gene.name)) %>%
  dplyr::select(combo) %>%
  unique %>%
  as.data.frame

# I also found some DMSP degrading activity in the eggnog results, cause there werent any in the metabolic database
s_eggnog_hits_DMSP <- fxn %>% 
  filter(Description %in% c("Mediates cleavage of dimethlysulfonioproprionate (DMSP) into dimethyl sulfide (DMS) and acrylate. DMS is the principal form by which sulfur is transported from oceans to the atmosphere and is a key component of the ocean sulfur cycle", "Dimethlysulfonioproprionate lyase")) %>%
  left_join(taxa, by = c("query" = "contig"))

s_eggnog_hits_DMSP

s_contig_hits <- HMM_hits %>%
  filter(Category %in% c("Sulfur cycling", "Sulfur cycling enzymes (detailed)")) %>%
  filter(FLK_OFAV_MG_pred.Hmm.presence == "Present") %>%
  mutate(FLK_OFAV_MG_pred.Hits = str_split(FLK_OFAV_MG_pred.Hits, ",")) %>%
  unnest_wider(FLK_OFAV_MG_pred.Hits, names_sep = "_") %>%
  gather(key = "arbitrary_hit_number", value = "contig", FLK_OFAV_MG_pred.Hits_1:FLK_OFAV_MG_pred.Hits_50) %>%
  filter(!is.na(contig))

s_contig_hits

# Which contigs contain sulfur cycling genes? Combine the eggnog sulfur contigs and the metabolic sulfur contigs
s_contig_hits$contig
s_eggnog_hits_DMSP$query

s_cycling_contigs <- c(s_contig_hits$contig, s_eggnog_hits_DMSP$query)

plyr::count(s_cycling_contigs) %>% nrow # number of unique genes
```

## Vitamin biosynthesis contig list
```{r}
vitamin_hits_broad <- KEGG_hits %>%
  filter(Module.Category == "Cofactor and vitamin metabolism") %>%
  filter(FLK_OFAV_MG_pred.Module.presence == "Present")

vitamin_hits_broad

# Need a list of the KO ID's so that I can cross-reference it with the eggnog output
vitamin_hits <- KEGG_step_hits %>%
  filter(Module.Category == "Cofactor and vitamin metabolism") %>%
  filter(FLK_OFAV_MG_pred.Module.step.presence == "Present") %>%
  mutate(KO.id.test = str_split(KO.id, ",")) %>% # make mutliple KOs different columns 
  unnest_wider(KO.id.test, names_sep = "_") %>%
  gather(key = "KO_arbitrary_id", value = "KO", KO.id.test_1:KO.id.test_6) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_split(KO, "[+]")) %>% # make mutliple KOs different columns that have plusses
  unnest_wider(KO, names_sep = "_") %>%
  gather(key = "KO_arbitrary_id2", value = "KO", KO_1:KO_3) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_split(KO, "[-]")) %>% # make mutliple KOs different columns that have minuses
  unnest_wider(KO, names_sep = "_") %>%
  gather(key = "KO_arbitrary_id3", value = "KO", KO_1:KO_2) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_replace_all(KO, "[)]", "")) %>% # remove the random parentheses
  mutate(KO = str_replace_all(KO, "[(]", ""))

vitamin_hits

# Now I need a list of the contigs that go with the different KO ids from the eggnog functional results
vitamin_eggnog_ko_contigs <- fxn %>%
  dplyr::select(query, KEGG_ko, KEGG_Module) %>%
  mutate(KEGG_ko = str_split(KEGG_ko, ",")) %>% # make mutliple KOs different columns 
  unnest_wider(KEGG_ko, names_sep = "_") %>%
  gather(key = "KO_random_number", value = "KO", KEGG_ko_1:KEGG_ko_15) %>%
  filter(!is.na(KO)) %>%
  mutate(KO = str_replace_all(KO, "ko:", "")) %>% # remove the random "ko:"
  dplyr::select(!KO_random_number)

vitamin_eggnog_ko_contigs

# Add the contig/gene ID information to the KEGG module information
vitamin_ko_contig_hits <- vitamin_hits %>%
  left_join(vitamin_eggnog_ko_contigs, by = "KO", relationship = "many-to-many") # need many to many because there are multiple x and y matches

vitamin_ko_contig_hits

# Which contigs contain vitamin biosynthesis genes?
vitamin_ko_contig_hits$query

vitamin_cycling_contigs <- vitamin_ko_contig_hits$query %>% unique

plyr::count(vitamin_cycling_contigs) %>% nrow # number of unique genes
```

## Subset count phyloseq object
I need to combine the contig lists (actually these are gene lists, which are subsets of the contigs). With that, then I need to subset the phyloseq count data set that I previously made. 
```{r}
# list of all functional genes
beneficial_fxn_genes <- c(n_cycling_contigs, s_cycling_contigs, vitamin_cycling_contigs)

# subset data from ps object (count data)
ps.beneficial_fxn_genes <- prune_taxa(beneficial_fxn_genes, ps.count)
```

## Run corncob analysis
I want to test for differentially abundant genes within healthy tissue between unaffected and diseased coral colonies, while contorlling for the effect of background zooxanthellae on the abundance of the taxa. I will not test for differential dispersion.
```{r}
# Set the levels so that corncob will choose unaffected as baseline
sample_data(ps.beneficial_fxn_genes)$Disease_on_colony_when_coring_binary <- factor(sample_data(ps.beneficial_fxn_genes)$Disease_on_colony_when_coring_binary,
                                                                     levels = c("Unaffected", "Diseased"))

## Control for the differences in the symbiont community
set.seed(10) # set seed for reproducibility
DA.METABOLIC.dis.symb <- differentialTest(formula = ~ Disease_on_colony_when_coring_binary + Zoox_background, 
                                  phi.formula = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  formula_null = ~ Zoox_background,
                                  phi.formula_null = ~ Disease_on_colony_when_coring_binary + Zoox_background,
                                  test = "Wald", boot = FALSE,
                                  data = ps.beneficial_fxn_genes,
                                  fdr_cutoff = 0.05)

plot(DA.METABOLIC.dis.symb)

# Save the corncob phyloseq object so you can manipulate it in R
saveRDS(DA.METABOLIC.dis.symb, "../data/corncob_METABOLIC_output_disease_at_sampling_zoox.rds")
```
OKay! So there are some differentially abundant genes! This is great and I feel better about how I am running the statistical tests on these data. I'm not exactly sure how to parse the data or visualize it yet, but this is at least cool that there are some clear results! Like before, I want to save a table of the corncob data output. 

## Save corncob output to table
```{r}
# I want to extract the important information from this into a table...
# check out results structure from one of the models
DA.METABOLIC.dis.symb$significant_models[[1]]$coefficients

# write a function to extract the corncob results
extractmods2 <- function(model) {
  result <- data.frame("Estimate" = model$coefficients[2, 1], 
                        "Std.Error" = model$coefficients[2, 2], 
                        "p" = model$coefficients[2, 4])
  return(result)
}

# save the output as a list using the above function
METABOLIC_corncob.dis.symb.models <- lapply(DA.METABOLIC.dis.symb$significant_models, extractmods2)
names(METABOLIC_corncob.dis.symb.models) <- DA.METABOLIC.dis.symb$significant_taxa

# turn the output from a list to a dataframe for ease of saving and manipulating
METABOLIC_corncob.dis.symb.models.df <- plyr::ldply(METABOLIC_corncob.dis.symb.models, data.frame)

### Save these data so you don't have to re-run the model ###
write.table(METABOLIC_corncob.dis.symb.models.df, "../data/Corncob_coassembly_results_METABOLIC_dis.symb_0.05.txt", sep="\t", row.names = FALSE, col.names = TRUE)
```

## Make tables of significance for N, S, Vitamin
### Nitrogen 
```{r}
corn.dis.symb <- read.table("../data/Corncob_coassembly_results_METABOLIC_dis.symb_0.05.txt", header = TRUE, sep = "\t")

N.da <- corn.dis.symb %>%
  filter(.id %in% n_cycling_contigs) %>%
  dplyr::left_join(n_contig_hits, by = c(".id" = "contig"))

ggplot(N.da, aes(x = .id, y = Estimate, fill = Function)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), 
                color = "black", width = .3, position=position_dodge(.9)) +
  geom_point(size = 4, pch = 21) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray") +
  labs(x = "Taxa", y = "Coefficient", title = "Nitrogen cycling DA genes")
```

### Sulfur
```{r}
corn.dis.symb <- read.table("../data/Corncob_coassembly_results_METABOLIC_dis.symb_0.05.txt", header = TRUE, sep = "\t")

S.da <- corn.dis.symb %>%
  filter(.id %in% s_cycling_contigs) %>%
  dplyr::left_join(s_contig_hits, by = c(".id" = "contig")) %>%
  dplyr::left_join(s_eggnog_hits_DMSP, by = c(".id" = "query")) %>%
  mutate(Function = case_when(
    .id == "k127_1971158_9" ~ "Mediates cleavage of DMSP to DMS and acrylate", TRUE ~ Function))

ggplot(S.da, aes(x = .id, y = Estimate, fill = Function)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), 
                color = "black", width = .3, position=position_dodge(.9)) +
  geom_point(size = 4, pch = 21) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray") +
  labs(x = "Taxa", y = "Coefficient", title = "Sulfur cycling DA genes") +
  scale_fill_brewer(palette = "Paired") +
  facet_grid(Function ~ ., space = "free", scales = "free") +
  coord_flip()
```

### Vitamin
```{r}
corn.dis.symb <- read.table("../data/Corncob_coassembly_results_METABOLIC_dis.symb_0.05.txt", header = TRUE, sep = "\t")

Vit.da <- corn.dis.symb %>%
  filter(.id %in% vitamin_cycling_contigs) %>%
  dplyr::left_join(vitamin_ko_contig_hits, by = c(".id" = "query")) %>% 
  mutate(module_broad = str_split_i(Module, ",", 1))
  
 
x1 <- distinctColorPalette(20)

ggplot(Vit.da, aes(x = .id, y = Estimate, fill = module_broad)) +
  geom_errorbar(aes(ymin = Estimate-Std.Error, ymax = Estimate+Std.Error), 
                color = "black", width = .3, position=position_dodge(.9)) +
  geom_point(size = 4, pch = 21) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray") +
  labs(x = "Taxa", y = "Coefficient", title = "Vitamin biosynthesis DA genes") +
  scale_fill_manual(values = x1) +
  facet_grid(module_broad ~ ., space = "free", scales = "free") +
  coord_flip()
```

