---
title: "Blank_analysis"
output: html_document
date: "2025-01-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Microbiome blank analysis

Since the metagenomics analysis is expensive and requires specific concentrations, we didn't send any DNA extraction blanks to the sequencing center for shotgun metagenomics. In order to evaluate any potential contamination, though, we did subject the blanks to 16S rRNA microbiome sequencing. While this includes an amplification step, it will give us a sense of what microbes are in the blank samples that may have also been sequenced metagenomically. If there are some taxa that overlap, I could consider removing them from the analysis.

# Criteria before I start

- Samples have been demultiplexed, i.e. split into individual per-sample fastq files.
YES! 
- Non-biological nucleotides have been removed, e.g. primers, adapters, linkers, etc.
YES! It looks like Julie used cutadapt to remove the different primers and adapters. 
- If paired-end sequencing data, the forward and reverse fastq files contain reads in matched order.
Not sure about this, but hopefully yes!

# Load packages and data

```{r}
library(dada2); packageVersion("dada2")
library(phyloseq)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw())
library(microViz)
```

## data
```{r}
path <- "/Users/cynthiabecker/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/blank_analysis/cutadapt"
list.files(path)
```

Read in the names of the fastq files and use string manipulation to get the different file names and match the fo4rward and reverse reads. 
```{r}
# Forward and reverse fastq filenames have format: SAMPLENAME_L001_R1_cut.fastq.gz and SAMPLENAME_L001_R2_cut.fastq.gz
fnFs <- sort(list.files(path, pattern="_L001_R1_cut.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_L001_R2_cut.fastq.gz", full.names = TRUE))

# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

# Plot and inspect quality profiles
```{r}
plotQualityProfile(fnFs[1:4])
```
IN a shock to noone, they are pretty mediocre quality profiles. This is because they are low biomass blank samples, so they look fairly shitty. That is ok though! Surprisingly, actually, there are a lot of reads in EC2, about 38k. That's quite a bit!

```{r}
plotQualityProfile(fnRs[1:4])
```
As expected, these have the same number of reads as the forward reads. Additionally, they are of lower quality, which is normal for reverse reads. Since these are 150 bp reads, they aren't getting cut off as much, which is great. I will just cut 5 bp off the end to get rid of any really big dros in quality. 

# Filter and trim

Assign file names for the filtered fastq files 

```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_cut_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_cut_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

Use stansdard filtering parameters along with the truncation lengths indicated from the quality profiles. 
```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(145,145),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE)
head(out)

# What percent of reads retained?
as.data.frame(out) %>% 
  mutate(reads.retained = reads.out/reads.in)
```

Looks like a lot of reads were dropped for EC3 adnd EC4, which had pretty few reads overall. ZThe first 2, which had a lot more reeads, retained 75% to 91% of reads. 

# Learn Error Rates

The DADA2 algorithm makes use of a parametric error model (`err`) and every amplicon dataset has a different set of error rates. The `learnErrors` method learns this error model from the data, by alternating estimation of the error rates and inference of sample composition until they converge on a jointly consistent solution. As in many machine-learning problems, the algorithm must begin with an initial guess, for which the maximum possible error rates in this data are used (the error rates if only the most abundant sequence is correct and all the rest are errors).

```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)

# Check out the error rates
plotErrors(errF, nominalQ=TRUE)
```

The error rates for each possible transition (A→C, A→G, …) are shown. Points are the observed error rates for each consensus quality score. The black line shows the estimated error rates after convergence of the machine-learning algorithm. The red line shows the error rates expected under the nominal definition of the Q-score. Here the estimated error rates (black line) are a good fit to the observed rates (points), and the error rates drop with increased quality as expected. Everything looks reasonable and we proceed with confidence.

# Sample inference

Time to apply the core sample inference model to the filtered and trimmed data

```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)

# Check out the returned dada-class object
dadaFs[[1]]
```
For the first sample, it found 110 sequence variants in the data. 

# Merge paired reads
```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```
# Construct the sequence table
This is the amplicon sequence variant (ASV) table that is a higher-resolution version of the OTU table
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```

Looks like overall, there are 231 amplicon sequence variants. most of them are the expected 253 basepairs. There is some variability in this, which is expected with a hypervariable region of the 16S rRNA gene

# Remove Chimeras
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)

sum(seqtab.nochim)/sum(seqtab)
```

Looks like there were 23 chimeras. That isn't too bad. Since 98% of the ASVs were not chimeric, that makes me feel good that the data are pretty good. Yay!

# Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))

# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# Save this table! It is useful to have
write.table(track, "blank_reads_through_pipeline.csv", sep = ",", row.names = TRUE, col.names = TRUE)
```


# Assign taxonomy
It is common at this point, especially in 16S/18S/ITS amplicon sequencing, to assign taxonomy to the sequence variants. The DADA2 package provides a native implementation of the naive Bayesian classifier method for this purpose. The `assignTaxonomy` function takes as input a set of sequences to be classified and a training set of reference sequences with known taxonomy, and outputs taxonomic assignments with at least minBoot bootstrap confidence.

We maintain formatted training fastas for the RDP training set, GreenGenes clustered at 97% identity, and the Silva reference database, and additional trainings fastas suitable for protists and certain specific environments have been contributed. For fungal taxonomy, the General Fasta release files from the UNITE ITS database can be used as is. To follow along, download the silva_nr_v132_train_set.fa.gz file, and place it in the directory with the fastq files.

I will get the silva database sequences and put it in the folder with my datasets. The current database is version 138.2.

```{r}
taxa <- assignTaxonomy(seqtab.nochim, "taxa/silva_nr99_v138.2_toGenus_trainset.fa.gz", multithread=TRUE)
taxa <- addSpecies(taxa, "taxa/silva_v138.2_assignSpecies.fa.gz") 
```

Inspect the taxonomic assignments for fragments that exist in the blanks. 
```{r}
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

#Make outputtable ASV tables and taxonomy files
Make sure the seqtab.nochim file with all the ASVs and the taxa are all in the same order. Save the sequences with an ASV# identifier.
```{r}
otus <- seqtab.nochim
taxonomy <- taxa

idx <- match(rownames(taxonomy), colnames(otus))
#looks like they were all aligned, but doesn't hurt just to make it very safe.
otus <- otus[,idx]

#save a dataframe with a new ASV identifier and the sequence from the rownames for taxa
#This is very important because you want to save the sequence for REPRODUCIBILITY and TRACTABILITY
ASVseqs <- data.frame("asv" = paste0("ASV", seq(from = 1, to = ncol(seqtab.nochim), by = 1)), 
                      "sequence" = rownames(taxa))

#rename otu and taxa dataframe so they are easier to interpret
colnames(otus) <- ASVseqs$asv
otus <- t(otus) #change OTU table so that the ASVs are rows and samples are columns
rownames(taxonomy) <- ASVseqs$asv
```


###Recap: Write these tables so you have them for supplementary data
```{r}
setwd("../blank_analysis/results/") #write out these as .txt files
write.table(otus, file = "ASV_Blank_RRC.txt", sep = "\t", row.names = TRUE, col.names = TRUE)
write.table(taxonomy, file = "taxonomy_Blank_RRC.txt", sep = "\t", row.names = TRUE, col.names = TRUE)
write.table(ASVseqs, file = "ASVsequences_Blank_RRC.txt", sep = "\t", row.names = TRUE, col.names = TRUE)
```

Interesting...I don't know acetobacter well. It seems common here in the results though.

# Phyloseq preliminary analysis
I would like to use phyloseq to investigate the composition of different taxonomic groups in the blanks. Questions I have are:
- What taxonomic groupings are common in the blanks?
- Are there groups that I am also seeing in the metagenomes that I should consider taking OUT given their representation in the blank samples?

```{r}
samples.out <- rownames(seqtab.nochim)

# Make phyloseq object
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               tax_table(taxa))
ps
```

It is more convenient to use short names for our ASVs (e.g. ASV21) rather than the full DNA sequence when working with some of the tables and visualizations from phyloseq, but we want to keep the full DNA sequences for other purposes like merging with other datasets or indexing into reference databases. For that reason we’ll store the DNA sequences of our ASVs in the refseq slot of the phyloseq object, and then rename our taxa to a short string. That way, the short new taxa names will appear in tables and plots, and we can still recover the DNA sequences corresponding to each ASV as needed with refseq(ps).

```{r}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
# check out the dna object
dna

ps <- merge_phyloseq(ps, dna)

taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```
now we have the `refseq()` part of the phyloseq object. This is great! I will use this if I ever want to relate the DNA sequence to any databases. 

Quick richness check
```{r}
plot_richness(ps, measures = "Observed")
```

# Top 20 taxa
Visualize a stacked bar plot of the different taxa in the control samples. 
```{r}
top20 <- names(sort(taxa_sums(ps), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)

plot_bar(ps.top20, fill="Family")
plot_bar(ps.top20, fill="Genus")

ggsave("results/top_20_blank_ASVs.pdf")
```
Next let's visualize all the taxa
```{r}
ps.ra <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))

plot_bar(ps.ra, fill="Family")
```

Okay, so it looks like a lot of different taxa showed up. Take a quick look using raw counts. I suspect the EC3 and EC4 samples will have MUCH more taxa. 
```{r}
plot_bar(ps, fill = "Family")

plot_bar(ps, fill = "Genus")
```
# Prevalent contaminants

Which Families are in all 4 of the samples?
```{r}
ps.family <- tax_glom(ps, taxrank = "Family")

# Define core contaminants as taxa that show up with a positive abundance in all four samples
ps.corecontam <- filter_taxa(ps.family, function(x) sum(x > 0) == 4, TRUE)

plot_bar(ps.corecontam, fill = "Family")
```
Which ASVs are in all 4 of the samples?
```{r}
# Define core contaminants as taxa that show up with a positive abundance in at least 3 samples
ps.corecontam.asv <- filter_taxa(ps, function(x) sum(x > 0) >= 3, TRUE)
ps.corecontam.asv

plot_bar(ps.corecontam.asv, fill = "Genus")
```
Which genera are in at least 3 controls?
```{r}
ps.genus <- tax_glom(ps, taxrank = "Genus")

# Define core contaminants as taxa that show up with a positive abundance in at least 3 samples
ps.corecontam.genus <- filter_taxa(ps.genus, function(x) sum(x > 0) >= 3, TRUE)
ps.corecontam.genus

plot_bar(ps.corecontam.genus, fill = "Genus")
```
# MAG taxa to blanks

Let's look at the different ASVs that fall into this. Keep in mind, the MAGs I recovered (indicating lots of coverage of these genera) in the metagenome coassembly for FL Keys corals were:
d__Bacteria;p__Pseudomonadota;c__Gammaproteobacteria;o__Burkholderiales;f__Burkholderiaceae_B;g__Pelomonas;s__
d__Bacteria;p__Actinomycetota;c__Actinomycetia;o__Mycobacteriales;f__Mycobacteriaceae;g__Lawsonella;s__Lawsonella clevelandensis_A
d__Bacteria;p__Actinomycetota;c__Actinomycetia;o__Propionibacteriales;f__Propionibacteriaceae;g__Cutibacterium;s__Cutibacterium namnetense
d__Bacteria;p__Actinomycetota;c__Actinomycetia;o__Propionibacteriales;f__Propionibacteriaceae;g__Cutibacterium;s__Cutibacterium acnes
d__Bacteria;p__Bacillota;c__Bacilli;o__Lactobacillales;f__Lactobacillaceae;g__Schleiferilactobacillus;s__Schleiferilactobacillus harbinensis
d__Bacteria;p__Bacillota_A;c__Clostridia;o__Peptostreptococcales;f__Caminicellaceae;g__JAKSBP01;s__JAKSBP01 sp022361035
d__Bacteria;p__Pseudomonadota;c__Alphaproteobacteria;o__Acetobacterales;f__Acetobacteraceae;g__Acetobacter;s__
d__Bacteria;p__Bacillota;c__Bacilli;o__Lactobacillales;f__Streptococcaceae;g__Streptococcus;s__
d__Bacteria;p__Pseudomonadota;c__Gammaproteobacteria;o__Xanthomonadales;f__Xanthomonadaceae;g__Xanthomonas_B;s__Xanthomonas_B massiliensis

Families to looks for:
* Burkholderiaceae
* Mycobacteriaceae
* Propionibacteriaceae
* Lactobacillaceae
* Caminicellaceae
* Acetobacteraceae
* Streptococcaceae
* Xanthomonadaceae

Genera to look for:
* Pelomonas
* Lawsonella
* Cutibacterium
* Schleiferilactobacillus
* JAKSBP01 (Clostridia)
* Acetobacter
* Streptococcus
* Xanthomonas

Next step...Look for and plot these specific families and genera
```{r}
ps.MAG.familes <- subset_taxa(ps, Family=="Burkholderiaceae" | Family=="Mycobacteriaceae" | Family=="Propionibacteriaceae" | Family=="Lactobacillaceae" | Family=="Caminicellaceae" | Family=="Acetobacteraceae" | Family=="Streptococcaceae" | Family=="Xanthomonadaceae")

plot_bar(ps.MAG.familes, fill = "Genus")
plot_bar(ps.MAG.familes, fill = "Family")

tax_table(ps.MAG.familes)
```

Next step...Look for and plot these specific Genera
```{r}
ps.MAG.genera <- subset_taxa(ps, Genus=="Pelomonas" | Genus=="Lawsonella" | Genus=="Cutibacterium" | Genus=="Schleiferilactobacillus" | Genus=="JAKSBP01" | Genus=="Acetobacter" | Genus=="Streptococcus" | Genus=="Xanthomonas")

plot_bar(ps.MAG.genera, fill = "Genus")
plot_bar(ps.MAG.genera, fill = "Family")

tax_table(ps.MAG.genera)
```

What if the 16S primers just don't do a good job amplifying the pelomonas, lawsonella, and cutibacterium? THe ones the MAGs generated?
I downloaded from silva the analysis of the 515FY and 806RB primers and how well they would match (or not) various taxa within the SILVA datbase. It is called `515FY_806RB_tax_matches_SILVA.csv` and it is at the following path: `/Users/cynthiabecker/Documents/GitHub/FLK_SCTLD_Resistance_OFAV/data`. 

Looks like it is mostly a mismatch for Cutibacterium. But it is a match for some propionibacteriaceae
Looks like it is a good match for Burkholderiaceae (Pelomonas isn't in the DB)
Looks like it is an ok match for Lawsonella (half match, half mismatch). ANd it is a good match for most Mycobacteriales. 
Looks like it is a great match for streptococcus. 

# Relate metagenome taxa to blanks
Rather than reading in all the taxa, I will just read in the phyloseq object of the count data. 
```{r}
# Load table with count abundances from salmon ("numreads")
prokreads <- read.table("../data/prok_FLK_coassembly_salmon_quant_all_NumReads.tsv", header = TRUE, sep = "\t", row.names = 2)
prokreads <- prokreads[, -1] # get rid of the first column cause it is just numbered lines

# how many genes are zero across all samples? Like 32k
table(rowSums(prokreads) == 0)

# remove them
prokreads_nozero <- prokreads[rowSums(prokreads) != 0, ]
dim(prokreads_nozero)

# any samples with no reads? No.
table(colSums(prokreads_nozero) == 0)

# Fix the sample names bc R doesn't like column names that start with a number
colnames(prokreads_nozero) <- str_replace_all(colnames(prokreads_nozero), pattern = "[X]", "")
colnames(prokreads_nozero) <- str_split_i(colnames(prokreads_nozero), "[_]", 1)

# Read in the taxonomy
prok_taxa <- read.table("../data/FLK_prok_coassembly_tax_mmseqs2.tsv", header = FALSE, col.names = c("contig", "ncbi_taxid", "rank", "annotation", "complete_classification"), na.strings = "", sep = "\t")

prok_taxa <- prok_taxa %>% 
  separate(complete_classification, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), 
           sep = "[;]", remove = FALSE) %>%
  dplyr::select(Domain:Species, contig) %>%
  as.data.frame()

rownames(prok_taxa) <- prok_taxa$contig
prok_taxa_matrix <- as.matrix(prok_taxa)
```

Make phyloseq object
```{r}
ps.mg <- phyloseq(otu_table(prokreads_nozero, taxa_are_rows=TRUE), 
               tax_table(prok_taxa_matrix))
ps.mg
```

# MG Taxa 
What are the most abundant taxa from the metagenomic analysis?
```{r}
ps.mg.family <- tax_glom(ps.mg, taxrank = "Family")
ps.mg.family.ra <- transform_sample_counts(ps.mg.family, function(OTU) OTU/sum(OTU))

# how many families total? 777
length(unique(tax_table(ps.mg.family)[,"Family"]))

ps.mg.family.ra %>% comp_barplot(tax_level = "Family", n_taxa = 15, other_name = "Other Families (n = 762)") +
  coord_flip()
```
What about the top most abundant genera?
```{r}
ps.mg.genus <- tax_glom(ps.mg, taxrank = "Genus")

# how many genera? 1747
ps.mg.genus

comp_barplot(ps.mg.genus, tax_level="Genus", n_taxa = 15, other_name = "Other Genera (n = 1,732)") +
  coord_flip()


```

Ugh....I'm REAAALLLY not sure how to proceed. It is clear that we have these 4 blanks, and they have varying degrees of contaminants.


# OPTIONS

## Cutibacterium

Remove cutibacterium. It isn't associated with the environemtn. It is a human skin contaminant. It doesn't show up in the blanks, because it likely isn't amplified well by the 515FY/806RB primers. Let's see. How many genes would it remove if I get rid of Cutibacterium-annotated contigs from my dataset?

```{r}
ps.mg

# NOTE - subset_taxa will also remove NA's, so make sure to keep the NA's 
ps.mg.Cuti_remove <- subset_taxa(ps.mg, Genus != "g_Cutibacterium" | is.na(Genus))

ps.mg.Cuti_remove

Cuti <- subset_taxa(ps.mg, Genus == "g_Cutibacterium")
Cuti
```
## Top 20 contaminants

Remove any genera that match the genera of the top 20 ASVs found in blank samples. Get a list of these genera first. 
```{r}
ps.mg

tax_table(ps.top20)[ ,"Genus"] %>% unique

subset_taxa(ps.mg, Genus == "g_Acetobacter") # 4408
subset_taxa(ps.mg, Genus == "g_Schleiferilactobacillus") # 2698
# NONE subset_taxa(ps.mg, Genus == "g_SN8")
# NONE subset_taxa(ps.mg, Genus == "g_Brucella")
subset_taxa(ps.mg, Genus == "g_Liquorilactobacillus") # 397
# NONE subset_taxa(ps.mg, Genus == "g_Peribacillus")
subset_taxa(ps.mg, Genus == "g_Lacticaseibacillus") # 304
subset_taxa(ps.mg, Genus == "g_Novosphingobium") # 1
subset_taxa(ps.mg, Genus == "g_Lentilactobacillus") # 324
subset_taxa(ps.mg, Genus == "g_Staphylococcus") # 5345
# NONE subset_taxa(ps.mg, Genus == "g_Niallia")
subset_taxa(ps.mg, Genus == "g_Xanthobacter") # 3 
subset_taxa(ps.mg, Genus == "g_Corynebacterium") # 824

ps.mg.top20_contam_remove <- subset_taxa(ps.mg, 
                                         Genus != "g_Acetobacter" & 
                                           Genus != "g_Schleiferilactobacillus" & 
                                           Genus != "g_Liquorilactobacillus" & 
                                           Genus != "g_Lacticaseibacillus" &
                                           Genus != "g_Novosphingobium" &
                                           Genus != "g_Lentilactobacillus" &
                                           Genus != "g_Staphylococcus" &
                                           Genus != "g_Xanthobacter" &
                                           Genus != "g_Corynebacterium" | is.na(Genus))
ps.mg.top20_contam_remove
```

## Highly prevalent contaminants

Remove Genera that were identified in 75% of the blanks (at least 3 of them). I called these prevalent contaminants or "core" contaminants. 
```{r}
ps.mg

tax_table(ps.corecontam.genus)[ ,"Genus"] %>% unique

subset_taxa(ps.mg, Genus == "g_Acetobacter") # 4408
subset_taxa(ps.mg, Genus == "g_Staphylococcus") # 5345
# NONE subset_taxa(ps.mg, Genus == "g_Niallia")
subset_taxa(ps.mg, Genus == "g_Xanthobacter") # 3 
subset_taxa(ps.mg, Genus == "g_Corynebacterium") # 824
subset_taxa(ps.mg, Genus == "g_Acinetobacter") # 23
# NONE subset_taxa(ps.mg, Genus == "g_Pleomorphomonas") # 
# NONE subset_taxa(ps.mg, Genus == "g_Cloacibacterium") # 
subset_taxa(ps.mg, Genus == "g_Veillonella") # 8
subset_taxa(ps.mg, Genus == "g_Sphingobium") # 2


ps.mg.prev_contam_remove <- subset_taxa(ps.mg, 
                                         Genus != "g_Acetobacter" & 
                                           Genus != "g_Staphylococcus" & 
                                           Genus != "g_Xanthobacter" & 
                                           Genus != "g_Corynebacterium" &
                                           Genus != "g_Acinetobacter" &
                                           Genus != "g_Veillonella" &
                                           Genus != "g_Sphingobium" | is.na(Genus))
ps.mg.prev_contam_remove
```

Save the Taxonomy table and ASV abundance table for the blanks. Also save the phyloseq object. 

```{r}
saveRDS(ps, "../data/blank_ps_object.rds")

write.table(as(otu_table(ps), "matrix"),
            "../data/blank_ASV_abundance.txt", sep="\t", col.names=NA)
write.table(as(tax_table(ps), "matrix"),
            "../data/blank_ASV_taxa.txt", sep="\t", col.names=NA)
```

